---
title: "Homework 5"
author: "Alex Ojemann"
date: "2024-11-01"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

## Problem 1

```{r}
# Load required libraries
library(dplyr)
library(tableone)
library(labelled)

# Read the data
tico_data <- read.csv("tico_data.csv")

tico_data$gender <- factor(tico_data$gender, levels = c(1, 2), labels = c("Male", "Female"))

# Convert "history of" and other binary categorical variables to factors with "No" and "Yes" labels
binary_vars <- c("asthmahx", "copdhx", "diabeteshx", "chfhx", "hypertenshx", 
                 "malignhx", "renalhx", "immuno_supp", "active")  # add other binary variables as needed
tico_data[binary_vars] <- lapply(tico_data[binary_vars], factor, levels = c(0, 1), labels = c("No", "Yes"))

# Assign descriptive variable labels using the labelled package
tico_data <- tico_data %>%
  set_variable_labels(
    trial = "Trial Arm",
    active = "Randomized Group (Active = 1, Placebo = 0)",
    gender = "Gender",
    age = "Age (years)",
    asthmahx = "History of Asthma",
    copdhx = "History of COPD",
    diabeteshx = "History of Diabetes",
    chfhx = "History of CHF",
    hypertenshx = "History of Hypertension",
    malignhx = "History of Malignancy",
    renalhx = "History of Renal Impairment",
    immuno_supp = "Immunosuppressed Status",
    bmi = "Body Mass Index (BMI)",
    newsbl = "NEWS Score at Baseline",
    creatbl = "Baseline Creatinine (mg/dL)",
    crpbl = "Baseline C-Reactive Protein (CRP)",
    lymphbl = "Baseline Lymphocyte Count",
    ordpulmbl = "Pulmonary Ordinal Scale",
    vl_rslt_log10 = "Viral Load Result (Log10)"
  )

# Set levels for categorical variables in the same way as gender for binary representation
tico_data$trial <- factor(tico_data$trial, 
                          levels = c("A", "B", "C", "D", "E"),
                          labels = c("Trial A", "Trial B", "Trial C", "Trial D", "Trial E"))


# Specify the variables to include in the table
baseline_vars <- c("trial", "active", "gender", "age", "asthmahx", "copdhx", "diabeteshx",
                   "chfhx", "hypertenshx", "malignhx", "renalhx", "immuno_supp", "bmi",
                   "newsbl", "creatbl", "crpbl", "lymphbl", "ordpulmbl", "vl_rslt_log10")

# Create the summary table stratified by treatment
table1 <- CreateTableOne(vars = baseline_vars, strata = "corticobl", data = tico_data, test = FALSE)

# Print the table with variable labels included
print(table1, varLabels = TRUE, quote = FALSE, noSpaces = FALSE)

```

## Problem 2

```{r}
# Part 2: Manual Calculation of Mean Difference, Confidence Interval, and p-value
outcome <- "hospital_free_28"
treatment <- "corticobl"

# Calculate means for each treatment group
mean_treated <- mean(tico_data$hospital_free_28[tico_data$corticobl == 1])
mean_control <- mean(tico_data$hospital_free_28[tico_data$corticobl == 0])

# Calculate the mean difference
mean_diff <- mean_treated - mean_control

# Calculate standard errors for each group
sd_treated <- sd(tico_data$hospital_free_28[tico_data$corticobl == 1])
sd_control <- sd(tico_data$hospital_free_28[tico_data$corticobl == 0])
n_treated <- sum(tico_data$corticobl == 1)
n_control <- sum(tico_data$corticobl == 0)

# Standard error of the mean difference
se_diff <- sqrt((sd_treated^2 / n_treated) + (sd_control^2 / n_control))

# 95% confidence interval for the mean difference
ci_lower <- mean_diff - 1.96 * se_diff
ci_upper <- mean_diff + 1.96 * se_diff

# Calculate the z-score
z_score <- mean_diff / se_diff

# Calculate the p-value for the two-tailed test
p_value <- 2 * (1 - pnorm(abs(z_score)))

# Output the results
cat("Mean Difference in Hospital-Free Days:\n")
cat("Treated Group Mean:", mean_treated, "\n")
cat("Control Group Mean:", mean_control, "\n")
cat("Mean Difference (Treated - Control):", mean_diff, "\n")
cat("95% Confidence Interval: (", ci_lower, ",", ci_upper, ")\n")
cat("z-score:", z_score, "\n")
cat("p-value:", p_value, "\n\n")

```
## Problem 3

This regression model is fit with the existing covariates. In this case, the ATT is the same as the ATE because there are no interaction terms so the predicted response has the same difference between treatment and control for all values.

```{r}
# Required Libraries
library(dplyr)
library(broom)

# Define covariates to include all relevant baseline characteristics
covariates <- c("age", "gender", "bmi", "asthmahx", "copdhx", "diabeteshx", 
                "chfhx", "hypertenshx", "malignhx", "renalhx", "immuno_supp", 
                "newsbl", "creatbl", "crpbl", "lymphbl", "ordpulmbl", "vl_rslt_log10")

# Outcome Variable
outcome_var <- "hospital_free_28"

# Step 1: Fit the Outcome Model on the Full Dataset for ATE

# Define the variables to keep (outcome variable, treatment, and covariates)
variables_to_keep <- c(outcome_var, "corticobl", covariates)

# Remove rows with any missing values in the specified columns using base R subsetting
data_clean <- tico_data[complete.cases(tico_data[, variables_to_keep]), variables_to_keep]

# We use a linear regression model to estimate ATE
outcome_model <- lm(as.formula(paste(outcome_var, "~ corticobl +", paste(covariates, collapse = " + "))), data = data_clean)


ate <- coef(summary(outcome_model))["corticobl", "Estimate"]
ate_se <- coef(summary(outcome_model))["corticobl", "Std. Error"]
ate_ci_lower <- ate - 1.96 * ate_se
ate_ci_upper <- ate + 1.96 * ate_se

ate_p_value <- coef(summary(outcome_model))["corticobl", "Pr(>|t|)"]

# Print the ATE results
cat("Average Treatment Effect (ATE):\n")
cat("- Estimate:", ate, "\n")
cat("- Standard Error:", ate_se, "\n")
cat("- 95% Confidence Interval: (", ate_ci_lower, ",", ate_ci_upper, ")\n")
cat("- p-value:", ate_p_value, "\n\n")

# Step 3: Calculate ATT using Counterfactual Predictions with Interaction Model for Treated Group
# Filter data to include only the treated group
treated_data <- data_clean %>% filter(corticobl == 1)

# Create two copies of the treated dataset for prediction under both treatment and control conditions
treated_data_treated <- treated_data %>% mutate(corticobl = 1)  # Predict as treated
treated_data_control <- treated_data %>% mutate(corticobl = 0)  # Predict as if untreated

# Predict outcomes for the treated group under both scenarios
predicted_as_treated <- predict(outcome_model, newdata = treated_data_treated)
predicted_as_control <- predict(outcome_model, newdata = treated_data_control)

# Calculate ATT as the mean difference between predicted treated and control outcomes for treated individuals
diffs <- na.omit(unname(predicted_as_treated - predicted_as_control))
att <- mean(diffs)

# Estimate standard error and confidence interval for the ATT
att_se <- ate_se * nrow(data_clean) / nrow(treated_data)
att_ci_lower <- att - 1.96 * att_se
att_ci_upper <- att + 1.96 * att_se

# Print ATT Results
cat("Average Treatment Effect among the Treated (ATT):\n")
cat("- Estimate:", att, "\n")
cat("- Standard Error:", att_se, "\n")
cat("- 95% Confidence Interval: (", att_ci_lower, ",", att_ci_upper, ")\n")
```

This regression model includes interaction terms. This was added in order to be able to show a meaningful ATT (not the same as the ATE). Forward stepwise variable selection was used to avoid too many insignificant covariates. Different choices of the penalty for additional covariates in stepwise variable selection were experimented with and these resulted in significantly different models and treatment effects. The model shown uses the standard AIC penalty. The estimate and standard error from the model can't be used as the ATE and standard error of the ATE due to the presence of interaction terms so they are calculated manually.

```{r}
# Required Libraries
library(dplyr)
library(broom)

# Define covariates to include all relevant baseline characteristics
covariates <- c("age", "gender", "bmi", "asthmahx", "copdhx", "diabeteshx", 
                "chfhx", "hypertenshx", "malignhx", "renalhx", "immuno_supp", 
                "newsbl", "creatbl", "crpbl", "lymphbl", "ordpulmbl", "vl_rslt_log10")


# Outcome Variable
outcome_var <- "hospital_free_28"


# Define the variables to keep (outcome variable, treatment, and covariates)
variables_to_keep <- c(outcome_var, "corticobl", covariates)

# Remove rows with any missing values in the specified columns using base R subsetting
data_clean <- tico_data[complete.cases(tico_data[, variables_to_keep]), variables_to_keep]



# Step 1: Fit the Outcome Model with Treatment-Covariate Interactions and Forward Stepwise Selection using Custom Penalty

# Define the base model (forces corticobl to be a part of it)
base_model <- lm(as.formula(paste(outcome_var, "~ corticobl")), data = data_clean)

# Define the full model with interactions for forward selection
full_formula <- as.formula(paste(outcome_var, "~ corticobl * (", paste(covariates, collapse = " + "), ")"))

# Set a custom `k` value 
custom_k <- 2

# Perform forward stepwise selection based on the custom penalty
# Perform forward stepwise selection based on the custom penalty without printing each step
outcome_model <- step(base_model, 
                      scope = list(lower = ~ corticobl, upper = full_formula), 
                      direction = "forward", 
                      k = custom_k, 
                      trace = 0)


# Step 2: Calculate the ATE from the selected model
# Create two copies of the full dataset for predictions under treatment and control conditions
data_treated <- data_clean %>% mutate(corticobl = 1)  # Set treatment to 1
data_control <- data_clean %>% mutate(corticobl = 0)  # Set treatment to 0

# Predict outcomes for each individual under both treatment conditions
predicted_treated <- predict(outcome_model, newdata = data_treated)
predicted_control <- predict(outcome_model, newdata = data_control)

# Calculate ATE as the mean difference between predicted treated and control outcomes
ate <- mean(predicted_treated - predicted_control)

# Estimate standard error and confidence interval for the ATT
ate_se <- sd(predicted_treated - predicted_control) / sqrt(nrow(data_clean))
ate_ci_lower <- ate - 1.96 * ate_se
ate_ci_upper <- ate + 1.96 * ate_se

ate_p_value <- coef(summary(outcome_model))["corticobl", "Pr(>|t|)"]

# Print the ATE results
cat("Average Treatment Effect (ATE):\n")
cat("- Estimate:", ate, "\n")
cat("- Standard Error:", ate_se, "\n")
cat("- 95% Confidence Interval: (", ate_ci_lower, ",", ate_ci_upper, ")\n")
cat("- p-value:", ate_p_value, "\n\n")

# Step 3: Calculate ATT using Counterfactual Predictions with Interaction Model for Treated Group
# Filter data to include only the treated group
treated_data <- data_clean %>% filter(corticobl == 1)

# Create two copies of the treated dataset for prediction under both treatment and control conditions
treated_data_treated <- treated_data %>% mutate(corticobl = 1)  # Predict as treated
treated_data_control <- treated_data %>% mutate(corticobl = 0)  # Predict as if untreated

# Predict outcomes for the treated group under both scenarios
predicted_as_treated <- predict(outcome_model, newdata = treated_data_treated)
predicted_as_control <- predict(outcome_model, newdata = treated_data_control)

# Calculate ATT as the mean difference between predicted treated and control outcomes for treated individuals
diffs <- na.omit(unname(predicted_as_treated - predicted_as_control))
att <- mean(diffs)

# Estimate standard error and confidence interval for the ATT
att_se <- sd(diffs) / sqrt(nrow(treated_data))
att_ci_lower <- att - 1.96 * att_se
att_ci_upper <- att + 1.96 * att_se

# Print ATT Results
cat("Average Treatment Effect among the Treated (ATT):\n")
cat("- Estimate:", att, "\n")
cat("- Standard Error:", att_se, "\n")
cat("- 95% Confidence Interval: (", att_ci_lower, ",", att_ci_upper, ")\n")
```
## Problem 4

### Part a

```{r}
# First fit propensity score model

# List of categorical covariates (binary or multi-level)
categorical_covariates <- c("gender", "asthmahx", "copdhx", "diabeteshx", 
                            "chfhx", "hypertenshx", "malignhx", "renalhx", 
                            "immuno_supp")

# List of numeric covariates
continuous_covariates <- c("age", "bmi", "newsbl", "creatbl", "crpbl", 
                        "lymphbl", "ordpulmbl", "vl_rslt_log10")


# List of covariates (both categorical and continuous)
covariates <- c(categorical_covariates, continuous_covariates)

# Create a new data frame with the necessary variables for the propensity model
propensity_data <- data_clean[, c(covariates, "corticobl")]

# Fit a logistic regression model to estimate propensity scores
propensity_model <- glm(corticobl ~ ., 
                        data = propensity_data, 
                        family = binomial())

# Add propensity scores to the data
data_clean$pscore <- predict(propensity_model, type = "response")

# Calculate inverse probability weights
data_clean <- data_clean %>%
  mutate(weights = ifelse(corticobl == 1, 1 / pscore, 1 / (1 - pscore)))


# Ensure categorical variables are numeric (0 and 1) for calculation
data_clean <- data_clean %>%
  mutate(across(all_of(categorical_covariates), as.numeric))

# Calculate SMDs using the same process for both continuous and binary categorical variables
calculate_smd_ipw <- function(var) {
  treated_mean <- weighted.mean(data_clean %>% filter(corticobl == 1) %>% pull(var), 
                                w = data_clean %>% filter(corticobl == 1) %>% pull(weights))
  control_mean <- weighted.mean(data_clean %>% filter(corticobl == 0) %>% pull(var), 
                                w = data_clean %>% filter(corticobl == 0) %>% pull(weights))
  pooled_sd <- sd(data_clean[[var]], na.rm = TRUE)
  (treated_mean - control_mean) / pooled_sd
}

# Apply SMD calculation across all covariates (continuous and categorical)
cat("Weighted Standardized Mean Differences of all Covariates:\n")
smd_ipw <- sapply(c(categorical_covariates, continuous_covariates), calculate_smd_ipw)
print(smd_ipw)

```
### Part b

```{r}
# Assuming your code includes propensity scores and propensity score stratification as follows

# Divide data into quintiles based on propensity scores (for example, 5 quintiles)
data_clean <- data_clean %>%
  mutate(stratum = ntile(pscore, 5))

# Calculate stratum-specific effects
stratum_effects <- data_clean %>%
  group_by(stratum) %>%
  summarize(
    n_stratum = n(),
    n_treated = sum(corticobl == 1),
    n_control = sum(corticobl == 0),
    treated_mean = mean(hospital_free_28[corticobl == 1], na.rm = TRUE),
    control_mean = mean(hospital_free_28[corticobl == 0], na.rm = TRUE),
    effect = treated_mean - control_mean,
    var_effect = (var(hospital_free_28[corticobl == 1], na.rm = TRUE) / n_treated) + 
                 (var(hospital_free_28[corticobl == 0], na.rm = TRUE) / n_control)
  )

# Total sample sizes
N <- nrow(data_clean)
N_treated <- sum(data_clean$corticobl == 1)

# 1. Calculate ATE
ate <- sum(stratum_effects$effect * (stratum_effects$n_stratum / N))

# 2. Calculate ATT
att <- sum(stratum_effects$effect * (stratum_effects$n_treated / N_treated))

# 3. Calculate standard errors for ATE and ATT
# Variance of ATE
ate_var <- sum((stratum_effects$n_stratum / N)^2 * stratum_effects$var_effect / stratum_effects$n_stratum)
ate_se <- sqrt(ate_var)

# Variance of ATT
att_var <- sum((stratum_effects$n_treated / N_treated)^2 * stratum_effects$var_effect / stratum_effects$n_treated)
att_se <- sqrt(att_var)

# 4. Confidence intervals for ATE and ATT
ate_ci <- c(ate - 1.96 * ate_se, ate + 1.96 * ate_se)
att_ci <- c(att - 1.96 * att_se, att + 1.96 * att_se)

# Print results
cat("ATE using Propensity Score Stratification:\n")
cat("- Estimate:", ate, "\n")
cat("- Standard Error:", ate_se, "\n")
cat("- 95% CI:", ate_ci, "\n\n")

cat("ATT using Propensity Score Stratification:\n")
cat("- Estimate:", att, "\n")
cat("- Standard Error:", att_se, "\n")
cat("- 95% CI:", att_ci, "\n")

```


### Part c

```{r}
# Required libraries
library(survey)
library(dplyr)

# Required libraries
library(survey)
library(dplyr)

# Step 1: Calculate ATE using the IPW weights and the survey design
# Use the existing weights in a survey design to estimate the ATE
design <- svydesign(ids = ~1, weights = ~weights, data = data_clean)
ate_model <- svyglm(hospital_free_28 ~ corticobl, design = design)
ate <- coef(summary(ate_model))["corticobl", "Estimate"]
ate_se <- coef(summary(ate_model))["corticobl", "Std. Error"]
ate_ci <- confint(ate_model)["corticobl", ]

# Step 2: Calculate ATT using IPW2 weights to reweight control outcomes
# Calculate the mean outcome for the treated group (observed treated)
treated_mean <- mean(data_clean$hospital_free_28[data_clean$corticobl == 1])

# Calculate the weighted mean outcome for the control group (reweighted to match treated distribution)
weighted_control_mean <- with(data_clean[data_clean$corticobl == 0, ], 
                              sum(weights * hospital_free_28) / sum(weights))

# ATT is the difference between the treated mean and the weighted control mean
att <- treated_mean - weighted_control_mean

# Calculate standard error for ATT (assuming normal approximation for simplicity)
att_se <- sd(data_clean$hospital_free_28[data_clean$corticobl == 1] - weighted_control_mean) / sqrt(nrow(data_clean[data_clean$corticobl == 1, ]))
att_ci_lower <- att - 1.96 * att_se
att_ci_upper <- att + 1.96 * att_se

# Print results
cat("ATE using IPW:\n")
cat("- Estimate:", ate, "\n")
cat("- 95% CI:", ate_ci, "\n\n")

cat("ATT using IPW:\n")
cat("- Estimate:", att, "\n")
cat("- Standard Error:", att_se, "\n")
cat("- 95% CI: (", att_ci_lower, ",", att_ci_upper, ")\n")

```

### Part d

```{r}
library(purrr)

# Add a unique identifier if it doesn’t already exist
data_clean$id <- seq_len(nrow(data_clean))

# Ensure treated and control datasets are set up correctly
treated_data <- data_clean %>% filter(corticobl == 1)
control_data <- data_clean %>% filter(corticobl == 0)

# Create the matched_pairs data frame by iterating over each treated row and finding the closest control match
matched_pairs <- treated_data %>%
  mutate(
    control_match_id = map_dbl(pscore, ~ {
      control_data %>%
        mutate(distance = abs(pscore - .x)) %>%
        arrange(distance) %>%
        slice(1) %>%
        pull(id)
    })
  )

# Retrieve outcomes for the matched pairs directly without joining
treated_outcomes <- matched_pairs$hospital_free_28
control_outcomes <- data_clean$hospital_free_28[match(matched_pairs$control_match_id, data_clean$id)]

# Calculate the ATT for the matched pairs
att <- mean(treated_outcomes - control_outcomes, na.rm = TRUE)

# Calculate the standard error and 95% confidence interval
att_se <- sd(treated_outcomes - control_outcomes, na.rm = TRUE) / sqrt(nrow(matched_pairs))
att_ci_lower <- att - 1.96 * att_se
att_ci_upper <- att + 1.96 * att_se

# Print the ATT results
cat("ATT using Propensity Score Matching:\n")
cat("- Estimate:", att, "\n")
cat("- Standard Error:", att_se, "\n")
cat("- 95% Confidence Interval: (", att_ci_lower, ",", att_ci_upper, ")\n")

```

## Problem 5

Looking at each of these estimates, we can see that the unadjusted mean difference reflects a significant negative average treatment effect, but the ATE and ATT estimates when using both outcome regression models, propensity score stratification, inverse probability weighting, and propensity score matching are all much closer to zero with zero being included in most of the 95% confidence intervals for these estimates, indicating that we can not draw much of a conclusion either way regarding the effectiveness of this treatment.

