---
title: "PubH7485 Homework5 Answerkey"
date: "2024-11-01"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

set.seed(7485)

library(dplyr)
library(labelled)
library(causaldata)
library(knitr)
library(MASS)
library(reshape2)
library(ggplot2)
library(survey)
library(tableone)
library(randomForest)
library(MatchIt)

library(formattable)
library(rms)
library(forestplot)
library(Matching)
```

```{r, include=F}
# load and impute dataset
tico_7485 <- read.csv(file.choose(), header=T) 

tico_7485$active <- factor(tico_7485$active, levels = c(0,1), labels = c('placebo', 'active'))
tico_7485$gender <- factor(tico_7485$gender, levels = c(1,2), labels = c('male', 'female'))
tico_7485$asthmahx <- factor(tico_7485$asthmahx, levels = c(0,1), labels = c('no', 'yes'))
tico_7485$copdhx <- factor(tico_7485$copdhx, levels = c(0,1), labels = c('no', 'yes'))
tico_7485$diabeteshx <- factor(tico_7485$diabeteshx, levels = c(0,1), labels = c('no', 'yes'))
tico_7485$chfhx <- factor(tico_7485$chfhx, levels = c(0,1), labels = c('no', 'yes'))
tico_7485$hypertenshx <- factor(tico_7485$hypertenshx, levels = c(0,1), labels = c('no', 'yes'))
tico_7485$malignhx <- factor(tico_7485$malignhx, levels = c(0,1), labels = c('no', 'yes'))
tico_7485$renalhx <- factor(tico_7485$renalhx, levels = c(0,1), labels = c('no', 'yes'))
tico_7485$immuno_supp <- factor(tico_7485$immuno_supp, levels = c(0,1), labels = c('no', 'yes'))
tico_7485$ordpulmbl <- factor(tico_7485$ordpulmbl, levels = c(2,3,4), labels = c('no O2', 'low O2', 'high O2'))
tico_7485$corticobl <- factor(tico_7485$corticobl, levels = c(0,1), labels = c('no', 'yes'))

tico_7485 <- set_variable_labels(tico_7485, asthmahx = "Asthma", 
                                 copdhx = 'COPD',
                                 diabeteshx = 'Diabetes requiring medication' ,
                                 chfhx = 'CHF',
                                 hypertenshx = 'Hypertension requiring medication', 
                                 malignhx = 'Malignancy (active or receiving treatment)',
                                 renalhx= 'Renal impairment',
                                 immuno_supp = 'Immunosuppressed',
                                 newsbl = 'New Score',
                                 creatbl = 'Creatinine (mg/dL)',
                                 crpbl='C-reactive protein',
                                 lymphbl = 'Lymphocytes (10^9/L)',
                                 ordpulmbl = 'Pulm ordinal', 
                                 vl_rslt_log10 = 'Log 10 of Viral load copies/mL')

# sapply(tico_7485, function(x) sum(is.na(x)))

mode_function <- function(x) {
  x <- as.character(x)
  tbl <- table(x, useNA = "ifany")
  as.character(names(tbl)[which.max(tbl)])
}

for (var in names(tico_7485)) {
  if (is.numeric(tico_7485[[var]])) {
    tico_7485[[var]][is.na(tico_7485[[var]])] <- median(tico_7485[[var]], na.rm = TRUE)
    print(paste("Imputed numeric variable:", var))
    
  } else if (is.factor(tico_7485[[var]])) {
    tico_7485[[var]][is.na(tico_7485[[var]])] <- mode_function(tico_7485[[var]])
    print(paste("Imputed categorical variable:", var))
  }
}

t1 <- CreateTableOne(var = names(tico_7485)[2:19], data = tico_7485, strata = 'corticobl')
t11 <- print(t1, smd = TRUE, showAllLevels = TRUE, varLabels = TRUE)
```

## Question 1 (10 points)

```{r, echo=F}
kable(t11, caption = "Summary Statistics Stratified by the use of corticosteroids")
```


## Question 2 (5 points)

The table below presented the unadjusted mean difference in hospitalized free days and 95% CI. The difference is significant since the 95% CI does not contain 0. 

```{r, echo=F}
num_t1 <- table(tico_7485$corticobl)[2]
num_c1 <- table(tico_7485$corticobl)[1]

## for hospital_free_28
hf28_t <- tico_7485$hospital_free_28[tico_7485$corticobl == 'yes']
hf28_c <- tico_7485$hospital_free_28[tico_7485$corticobl == 'no']
et_hf28 <- mean(hf28_t, na.rm = T)
ec_hf28 <- mean(hf28_c, na.rm = T)

# Unadjusted ATE
ate_hf28 <- et_hf28 - ec_hf28
# Standard Error
se_hf28 <- sqrt(var(hf28_t, na.rm = T)/num_t1 + var(hf28_c, na.rm = T)/num_c1)
# Confidence Interval
conf_int_hf28 <- ate_hf28 + c(-1, 1)*qnorm(0.975)*se_hf28

output1 <- data.frame(
  Type = 'ATE', 
  Estimate = round(ate_hf28, 3),
  SE = round(unname(se_hf28), 3),
  CI_Lower = round(conf_int_hf28[1], 3),
  CI_Upper = round(conf_int_hf28[2], 3)
)

kable(output1, caption = "Unadjusted ATE for Hospital-Free Days")
```


## Question3 (10 points)

The ATE and ATT estimates using regression model are presented below. The differences are not significant since both confidence intervals contain 0. 

```{r, include=F}
calculate_ATE_ATT <- function(model_formula, data, num_bootstraps = 100) {
 
  # Fit initial outcome model
  hf28_model <- lm(model_formula, data = data)
  
  # Data for ATE (all individuals set to treated and control)
  data_trt <- data_con <- data
  data_trt$corticobl = 'yes'
  data_con$corticobl = 'no'
  
  # Predicted outcomes for ATE
  pred1_hf28 <- predict(hf28_model, newdata = data_trt, type = "response")
  pred0_hf28 <- predict(hf28_model, newdata = data_con, type = "response")
  ATE_hf28 <- mean(pred1_hf28 - pred0_hf28)
  
  # Data for ATT (only treated individuals set to control)
  data_att_trt <- data[data$corticobl == 'yes', ]
  data_att_con <- data_att_trt
  data_att_con$corticobl = 'no'
  
  # Predicted outcomes for ATT
  pred1_att_hf28 <- predict(hf28_model, newdata = data_att_trt, type = "response")
  pred0_att_hf28 <- predict(hf28_model, newdata = data_att_con, type = "response")
  ATT_hf28 <- mean(pred1_att_hf28 - pred0_att_hf28)
  
  # Bootstrap replication for SE
  ate_hf28_boot <- c()
  att_hf28_boot <- c()
  
  for (i in 1:num_bootstraps) {
    data_boot <- data[sample(nrow(data), size = nrow(data), replace = TRUE), ]
    hf28_model_boot <- lm(model_formula, data = data_boot)
    
    # Bootstrap for ATE
    data_trt_boot <- data_con_boot <- data_boot
    data_trt_boot$corticobl = 'yes'
    data_con_boot$corticobl = 'no'
    pred1_hf28_boot <- predict(hf28_model_boot, newdata = data_trt_boot, type = "response")
    pred0_hf28_boot <- predict(hf28_model_boot, newdata = data_con_boot, type = "response")
    ate_hf28_boot <- c(ate_hf28_boot, mean(pred1_hf28_boot - pred0_hf28_boot))
    
    # Bootstrap for ATT
    data_att_boot_trt <- data_boot[data_boot$corticobl == 'yes', ]
    data_att_boot_con <- data_att_boot_trt
    data_att_boot_con$corticobl = 'no'
    pred1_att_hf28_boot <- predict(hf28_model_boot, newdata = data_att_boot_trt, type = "response")
    pred0_att_hf28_boot <- predict(hf28_model_boot, newdata = data_att_boot_con, type = "response")
    att_hf28_boot <- c(att_hf28_boot, mean(pred1_att_hf28_boot - pred0_att_hf28_boot))
  }
  
  # Calculate standard errors and confidence intervals
  SE_ATE <- sd(ate_hf28_boot)
  SE_ATT <- sd(att_hf28_boot)
  CI_ATE_lower <- ATE_hf28 - qnorm(0.975) * SE_ATE
  CI_ATE_upper <- ATE_hf28 + qnorm(0.975) * SE_ATE
  CI_ATT_lower <- ATT_hf28 - qnorm(0.975) * SE_ATT
  CI_ATT_upper <- ATT_hf28 + qnorm(0.975) * SE_ATT
  
  # Return results as separate CI bounds
  return(list(
    ATE = round(ATE_hf28, 3), SE_ATE = round(SE_ATE, 3), 
    CI_ATE_lower = round(CI_ATE_lower, 3), CI_ATE_upper = round(CI_ATE_upper, 3),
    ATT = round(ATT_hf28, 3), SE_ATT = round(SE_ATT, 3), 
    CI_ATT_lower = round(CI_ATT_lower, 3), CI_ATT_upper = round(CI_ATT_upper, 3)
  ))
}


full_model <-  paste('hospital_free_28 ~ corticobl * (', paste(names(tico_7485)[1:19], collapse = ' + '), ')')
full_hf28 <- lm(as.formula(full_model), data=tico_7485) 

# do stepwise selection
stepwise_hf28 <- stepAIC(full_hf28, direction = "both")
# summary(stepwise_hf28)

select_hf28 <- hospital_free_28 ~ corticobl + gender + age + copdhx + malignhx + immuno_supp + 
  bmi + newsbl + creatbl + crpbl + lymphbl + ordpulmbl + vl_rslt_log10 + corticobl:bmi +
  corticobl:lymphbl 
result_hf28 <- calculate_ATE_ATT(model_formula = select_hf28, data = tico_7485, num_bootstraps = 100)
# print(result_hf28)
```


```{r, echo=F}
output2 <- data.frame(
  Type = c("ATE", "ATT"),
  Estimate = c(result_hf28$ATE, result_hf28$ATT),
  SE = c(result_hf28$SE_ATE, result_hf28$SE_ATT),
  CI_Lower = c(result_hf28$CI_ATE_lower, result_hf28$CI_ATT_lower),
  CI_Upper = c(result_hf28$CI_ATE_upper, result_hf28$CI_ATT_upper)
)


kable(output2, caption = "ATE and ATT Estimates from Regression Model")
```

Note that I was using the regression model from stepwise model selection; the results may be slightly different when using other models. 

## Question 4 (35 points total; 5 points for a, 10 points for b, c, d)

a) 

```{r,include=F}
# a) Find the IPW standardized mean difference among treatment groups
ps_model <- glm(corticobl ~ gender + age + copdhx + malignhx + immuno_supp + 
  bmi + newsbl + creatbl + crpbl + lymphbl + ordpulmbl + vl_rslt_log10, family = binomial(), data = tico_7485)
tico_7485$ps <- predict(ps_model, type = "response")
tico_7485$weight <- with(tico_7485, ifelse(corticobl == "yes", 1 / ps, 1 / (1 - ps)))

dataSvy <- svydesign(ids = ~ 1, data = tico_7485, weights = ~ weight)

tabWeighted <- svyCreateTableOne(vars = names(tico_7485)[2:19], strata = "corticobl", data = dataSvy, test = FALSE)
t2 <- print(tabWeighted, smd = TRUE, showAllLevels = TRUE, varLabels = TRUE)
```

```{r,echo=F}
kable(t2, caption = "Summary Statistics Stratified by the use of corticosteroids using IPW")
```

b) The ATE and ATT estimates using PSS are presented below. The differences are not significant since both confidence intervals contain 0. 

```{r,echo=F}
# b) ATE and ATT with PSS
# ATE estimate
tico_7485$corticobl_numeric <- as.numeric(tico_7485$corticobl == "yes")
ps <- predict(ps_model, type = "response")

ps_quintile_ate <- cut(ps, breaks = quantile(ps, probs = seq(0, 1, by = 0.2)), 
                       include.lowest = TRUE, labels = FALSE)
te_quintile <- tapply(tico_7485$hospital_free_28[tico_7485$corticobl == 'yes'], 
                      ps_quintile_ate[tico_7485$corticobl == 'yes'], mean) -
               tapply(tico_7485$hospital_free_28[tico_7485$corticobl == 'no'], 
                      ps_quintile_ate[tico_7485$corticobl == 'no'], mean)
ATE_PSS <- sum(te_quintile * table(ps_quintile_ate) / nrow(tico_7485), na.rm = TRUE)

# ATT estimate
ps_quintile_att <- cut(ps, 
	breaks = c(0, quantile(ps[tico_7485$corticobl == 'yes'], p = c(0.2, 0.4, 0.6, 0.8)), 1), labels = 1:5)
nA <- nrow(tico_7485[tico_7485$corticobl == 'yes', ])
nAj <- table(ps_quintile_att[tico_7485$corticobl == 'yes'])
te_quintile_att <- tapply(tico_7485$hospital_free_28[tico_7485$corticobl == 'yes'], 
                         ps_quintile_att[tico_7485$corticobl == 'yes'], mean) -
	tapply(tico_7485$hospital_free_28[tico_7485$corticobl == 'no'], 
	       ps_quintile_att[tico_7485$corticobl == 'no'], mean)
ATT_PSS <- sum(te_quintile_att * nAj / nA, na.rm = TRUE)

# bootstrap to get the SE and CI
B <- 100
ATE_PSS.boot <- NULL
ATT_PSS.boot <- NULL
n <- nrow(tico_7485)

for(i in 1:B) {

  tico.boot <- tico_7485[sample(1:n, n, replace = TRUE), ]
  ps_model.boot <- glm(corticobl ~ gender + age + copdhx + malignhx + immuno_supp + 
  bmi + newsbl + creatbl + crpbl + lymphbl + ordpulmbl + vl_rslt_log10,
                 data = tico.boot, family = "binomial")
  
  ps.boot <- predict(ps_model.boot, type = "response")
  
  # ATE
  ps_quintile.boot <- cut(ps.boot,
                          breaks = quantile(ps.boot, probs = seq(0, 1, by = 0.2), na.rm = TRUE),
                          include.lowest = TRUE, labels = FALSE)
  
  te_quintile.boot <- tapply(tico.boot$hospital_free_28[tico.boot$corticobl == 'yes'],
                             ps_quintile.boot[tico.boot$corticobl == 'yes'], mean) -
    tapply(tico.boot$hospital_free_28[tico.boot$corticobl == 'no'],
           ps_quintile.boot[tico.boot$corticobl == 'no'], mean)
  
  nj.boot <- table(ps_quintile.boot)
  ATE.boot <- sum(te_quintile.boot * nj.boot / n, na.rm = TRUE)
  
  ATE_PSS.boot <- c(ATE_PSS.boot, ATE.boot)
  
  # ATT
  ps_quintile.boot_att <- cut(ps.boot,
                              breaks = c(0, quantile(ps.boot[tico.boot$corticobl == 'yes'], 
                                                     probs = c(0.2, 0.4, 0.6, 0.8), na.rm = TRUE), 1),
                              include.lowest = TRUE, labels = FALSE)
  
  nA.boot <- sum(tico.boot$corticobl == 'yes')
  nAj.boot <- table(ps_quintile.boot_att[tico.boot$corticobl == 'yes'])
  
  te_quintile_att.boot <- tapply(tico.boot$hospital_free_28[tico.boot$corticobl == 'yes'],
                                 ps_quintile.boot_att[tico.boot$corticobl == 'yes'], mean) -
    tapply(tico.boot$hospital_free_28[tico.boot$corticobl == 'no'],
           ps_quintile.boot_att[tico.boot$corticobl == 'no'], mean)
  
  ATT.boot <- sum(te_quintile_att.boot * nAj.boot / nA.boot, na.rm = TRUE)
  
  ATT_PSS.boot <- c(ATT_PSS.boot, ATT.boot)
}

SE_ATE_PSS <- sd(ATE_PSS.boot)
SE_ATT_PSS <- sd(ATT_PSS.boot)

CI_ATE_PSS <- ATE_PSS + c(-1, 1) * qnorm(0.975) * SE_ATE_PSS
CI_ATT_PSS <- ATT_PSS + c(-1, 1) * qnorm(0.975) * SE_ATT_PSS

output3 <- data.frame(
  Type = c("ATE", "ATT"),
  Estimate = c(ATE_PSS, ATT_PSS),
  SE = c(SE_ATE_PSS, SE_ATT_PSS),
  CI_Lower = c(CI_ATE_PSS[1], CI_ATT_PSS[1]),
  CI_Upper = c(CI_ATE_PSS[2], CI_ATT_PSS[2])
)

kable(output3, caption = "ATE and ATT Estimates using PSS")
```

c) The ATE and ATT estimates using IPW are presented below. The differences are not significant since both confidence intervals contain 0. 

```{r, echo=F}
# ATE and ATT estimates using IPW
# ATE
w1 <- as.numeric(tico_7485$corticobl_numeric)/ps
w0 <- (1-tico_7485$corticobl_numeric)/(1-ps)
ATE_IPW <- mean(tico_7485$hospital_free_28*w1) - mean(tico_7485$hospital_free_28*w0)

# ATT
w1 <- tico_7485$corticobl_numeric
w0 <- (1-tico_7485$corticobl_numeric)/(1-ps)*ps
ATT_IPW <- weighted.mean(tico_7485$hospital_free_28,w=w1) - 
	weighted.mean(tico_7485$hospital_free_28,w=w0)

# bootstrap
B <- 100
n <- nrow(tico_7485)
ATE_IPW.boot <- NULL
ATT_IPW.boot <- NULL

for(i in 1:B) {

  tico.boot <- tico_7485[sample(1:n, n, replace = TRUE), ]
  
  p1.boot <- glm(corticobl ~ gender + age + copdhx + malignhx + immuno_supp + bmi + newsbl + 
                   creatbl + crpbl + lymphbl + ordpulmbl + vl_rslt_log10, data = tico.boot, family = "binomial")
  ps.boot <- predict(p1.boot, type = "response")
  
  w1.boot <- as.numeric(tico.boot$corticobl_numeric) / ps.boot
  w0.boot <- (1 - tico.boot$corticobl_numeric) / (1 - ps.boot)
  
  ATE_IPW.boot <- c(ATE_IPW.boot, 
                    mean(tico.boot$hospital_free_28 * w1.boot) - 
                    mean(tico.boot$hospital_free_28 * w0.boot))
  
  w1_att.boot <- tico.boot$corticobl_numeric
  w0_att.boot <- (1 - tico.boot$corticobl_numeric) / (1 - ps.boot) * ps.boot
  
  ATT_IPW.boot <- c(ATT_IPW.boot, 
                    weighted.mean(tico.boot$hospital_free_28, w = w1_att.boot) - 
                    weighted.mean(tico.boot$hospital_free_28, w = w0_att.boot))
}

SE_ATE_IPW <- sd(ATE_IPW.boot)
CI_ATE_IPW <- ATE_IPW + c(-1, 1) * qnorm(0.975) * SE_ATE_IPW

SE_ATT_IPW <- sd(ATT_IPW.boot)
CI_ATT_IPW <- ATT_IPW + c(-1, 1) * qnorm(0.975) * SE_ATT_IPW

output4 <- data.frame(
  Type = c("ATE", "ATT"),
  Estimate = c(ATE_IPW, ATT_IPW),
  SE = c(SE_ATE_IPW, SE_ATT_IPW),
  CI_Lower = c(CI_ATE_IPW[1], CI_ATT_IPW[1]),
  CI_Upper = c(CI_ATE_IPW[2], CI_ATT_IPW[2])
)

kable(output4, caption = "ATE and ATT Estimates using IPW")

```

d) i.
```{r, include=F}
# d) 1:1 matching 
# i.Find the standardized mean difference among the matched cohort.
rownames(tico_7485)<-1:nrow(tico_7485)
mod_match<-matchit(corticobl ~ gender + age + copdhx + malignhx + immuno_supp + bmi + newsbl + creatbl + crpbl + lymphbl +
                     ordpulmbl + vl_rslt_log10,distance="logit",method="nearest",data=tico_7485,ratio=1,replace=TRUE)
#print(tail(mod_match$match.matrix))

matched_pairs <- na.omit(mod_match$match.matrix)  # to check if controls re used or not
n_matched <- nrow(matched_pairs)
print(n_matched) 

tico_match <- data.frame(rbind(tico_7485[tico_7485$corticobl == 'yes', ],
	tico_7485[as.numeric(mod_match$match.matrix), ]))

tico_match <- set_variable_labels(tico_match, asthmahx = "Asthma", 
                                 copdhx = 'COPD',
                                 diabeteshx = 'Diabetes requiring medication' ,
                                 chfhx = 'CHF',
                                 hypertenshx = 'Hypertension requiring medication', 
                                 malignhx = 'Malignancy (active or receiving treatment)',
                                 renalhx= 'Renal impairment',
                                 immuno_supp = 'Immunosuppressed',
                                 newsbl = 'New Score',
                                 creatbl = 'Creatinine (mg/dL)',
                                 crpbl='C-reactive protein',
                                 lymphbl = 'Lymphocytes (10^9/L)',
                                 ordpulmbl = 'Pulm ordinal', 
                                 vl_rslt_log10 = 'Log 10 of Viral load copies/mL')

tabmatched1.1 <- CreateTableOne(vars = names(tico_7485)[2:19], strata = "corticobl",
	data = tico_match, test = FALSE)
t3 <- print(tabmatched1.1, smd = TRUE, showAllLevels = TRUE, 
						varLabels = TRUE)
```


```{r, echo=F}
kable(t3, caption = "Summary Statistics Stratified by the use of corticosteroids using 1:1 Matching")
```

ii. The ATT estimate using 1:1 matching are presented below. The difference is not significant since the confidence interval contains 0. 

```{r, echo=F}
# ii.Estimate the ATT using 1:1 matching
p1 <- mean(tico_7485$hospital_free_28[tico_7485$corticobl == 'yes'])
p0 <- mean(tico_7485$hospital_free_28[as.numeric(mod_match$match.matrix)],na.rm=TRUE)
n1 <- table(tico_7485$corticobl)[2]
n0 <- length(as.numeric(mod_match$match.matrix))

ATT1match <- p1 -  p0
SE1match <- sqrt(-1*(p1*(1-p1)/n1 + p0*(1-p0)/n0))
CI1match <- p1 - p0 + c(-1, 1)*qnorm(0.975)*SE1match

output5 <- data.frame(
  Type = 'ATT',
  Estimate = round(ATT1match, 3),
  SE = round(unname(SE1match), 3),
  CI_Lower = round(CI1match[1], 3),
  CI_Upper = round(CI1match[2], 3)
)

kable(output5, caption = "ATT Estimate using 1:1 Matching")
```


## Question 5 (10 points)

The analyses above indicate that the adjusted ATE and ATT estimates using various causal methods were not significant, while the unadjusted ATE was significant. This highlights the importance of carefully selecting causal methods in non-randomized studies (and even in randomized studies when the factor of interest is not the primary treatment).


# Appendix - R code

## Question 1 
```{r, eval=F}
# load and impute dataset
tico_7485 <- read.csv(file.choose(), header=T)

tico_7485$active <- factor(tico_7485$active, levels = c(0,1), labels = c('placebo', 'active'))
tico_7485$gender <- factor(tico_7485$gender, levels = c(1,2), labels = c('male', 'female'))
tico_7485$asthmahx <- factor(tico_7485$asthmahx, levels = c(0,1), labels = c('no', 'yes'))
tico_7485$copdhx <- factor(tico_7485$copdhx, levels = c(0,1), labels = c('no', 'yes'))
tico_7485$diabeteshx <- factor(tico_7485$diabeteshx, levels = c(0,1), labels = c('no', 'yes'))
tico_7485$chfhx <- factor(tico_7485$chfhx, levels = c(0,1), labels = c('no', 'yes'))
tico_7485$hypertenshx <- factor(tico_7485$hypertenshx, levels = c(0,1), labels = c('no', 'yes'))
tico_7485$malignhx <- factor(tico_7485$malignhx, levels = c(0,1), labels = c('no', 'yes'))
tico_7485$renalhx <- factor(tico_7485$renalhx, levels = c(0,1), labels = c('no', 'yes'))
tico_7485$immuno_supp <- factor(tico_7485$immuno_supp, levels = c(0,1), labels = c('no', 'yes'))
tico_7485$ordpulmbl <- factor(tico_7485$ordpulmbl, levels = c(2,3,4), labels = c('no O2', 'low O2', 'high O2'))
tico_7485$corticobl <- factor(tico_7485$corticobl, levels = c(0,1), labels = c('no', 'yes'))

tico_7485 <- set_variable_labels(tico_7485, asthmahx = "Asthma", 
                                 copdhx = 'COPD',
                                 diabeteshx = 'Diabetes requiring medication' ,
                                 chfhx = 'CHF',
                                 hypertenshx = 'Hypertension requiring medication', 
                                 malignhx = 'Malignancy (active or receiving treatment)',
                                 renalhx= 'Renal impairment',
                                 immuno_supp = 'Immunosuppressed',
                                 newsbl = 'New Score',
                                 creatbl = 'Creatinine (mg/dL)',
                                 crpbl='C-reactive protein',
                                 lymphbl = 'Lymphocytes (10^9/L)',
                                 ordpulmbl = 'Pulm ordinal', 
                                 vl_rslt_log10 = 'Log 10 of Viral load copies/mL')

# sapply(tico_7485, function(x) sum(is.na(x)))

mode_function <- function(x) {
  x <- as.character(x)
  tbl <- table(x, useNA = "ifany")
  as.character(names(tbl)[which.max(tbl)])
}

for (var in names(tico_7485)) {
  if (is.numeric(tico_7485[[var]])) {
    tico_7485[[var]][is.na(tico_7485[[var]])] <- median(tico_7485[[var]], na.rm = TRUE)
    print(paste("Imputed numeric variable:", var))
    
  } else if (is.factor(tico_7485[[var]])) {
    tico_7485[[var]][is.na(tico_7485[[var]])] <- mode_function(tico_7485[[var]])
    print(paste("Imputed categorical variable:", var))
  }
}

t1 <- CreateTableOne(var = names(tico_7485)[2:19], data = tico_7485, strata = 'corticobl')
t11 <- print(t1, smd = TRUE, showAllLevels = TRUE, varLabels = TRUE)
kable(t11, caption = "Summary Statistics Stratified by the use of corticosteroids")
```

## Question 2

```{r,eval=F}
num_t1 <- table(tico_7485$corticobl)[2]
num_c1 <- table(tico_7485$corticobl)[1]

## for hospital_free_28
hf28_t <- tico_7485$hospital_free_28[tico_7485$corticobl == 'yes']
hf28_c <- tico_7485$hospital_free_28[tico_7485$corticobl == 'no']
et_hf28 <- mean(hf28_t, na.rm = T)
ec_hf28 <- mean(hf28_c, na.rm = T)

# Unadjusted ATE
ate_hf28 <- et_hf28 - ec_hf28
# Standard Error
se_hf28 <- sqrt(var(hf28_t, na.rm = T)/num_t1 + var(hf28_c, na.rm = T)/num_c1)
# Confidence Interval
conf_int_hf28 <- ate_hf28 + c(-1, 1)*qnorm(0.975)*se_hf28

output1 <- data.frame(
  Type = 'ATE', 
  Estimate = round(ate_hf28, 3),
  SE = round(unname(se_hf28), 3),
  CI_Lower = round(conf_int_hf28[1], 3),
  CI_Upper = round(conf_int_hf28[2], 3)
)

kable(output1, caption = "Unadjusted ATE for Hospital-Free Days")
```


## Question 3
```{r,eval=F}
calculate_ATE_ATT <- function(model_formula, data, num_bootstraps = 100) {
 
  # Fit initial outcome model
  hf28_model <- lm(model_formula, data = data)
  
  # Data for ATE (all individuals set to treated and control)
  data_trt <- data_con <- data
  data_trt$corticobl = 'yes'
  data_con$corticobl = 'no'
  
  # Predicted outcomes for ATE
  pred1_hf28 <- predict(hf28_model, newdata = data_trt, type = "response")
  pred0_hf28 <- predict(hf28_model, newdata = data_con, type = "response")
  ATE_hf28 <- mean(pred1_hf28 - pred0_hf28)
  
  # Data for ATT (only treated individuals set to control)
  data_att_trt <- data[data$corticobl == 'yes', ]
  data_att_con <- data_att_trt
  data_att_con$corticobl = 'no'
  
  # Predicted outcomes for ATT
  pred1_att_hf28 <- predict(hf28_model, newdata = data_att_trt, type = "response")
  pred0_att_hf28 <- predict(hf28_model, newdata = data_att_con, type = "response")
  ATT_hf28 <- mean(pred1_att_hf28 - pred0_att_hf28)
  
  # Bootstrap replication for SE
  ate_hf28_boot <- c()
  att_hf28_boot <- c()
  
  for (i in 1:num_bootstraps) {
    data_boot <- data[sample(nrow(data), size = nrow(data), replace = TRUE), ]
    hf28_model_boot <- lm(model_formula, data = data_boot)
    
    # Bootstrap for ATE
    data_trt_boot <- data_con_boot <- data_boot
    data_trt_boot$corticobl = 'yes'
    data_con_boot$corticobl = 'no'
    pred1_hf28_boot <- predict(hf28_model_boot, newdata = data_trt_boot, type = "response")
    pred0_hf28_boot <- predict(hf28_model_boot, newdata = data_con_boot, type = "response")
    ate_hf28_boot <- c(ate_hf28_boot, mean(pred1_hf28_boot - pred0_hf28_boot))
    
    # Bootstrap for ATT
    data_att_boot_trt <- data_boot[data_boot$corticobl == 'yes', ]
    data_att_boot_con <- data_att_boot_trt
    data_att_boot_con$corticobl = 'no'
    pred1_att_hf28_boot <- predict(hf28_model_boot, newdata = data_att_boot_trt, type = "response")
    pred0_att_hf28_boot <- predict(hf28_model_boot, newdata = data_att_boot_con, type = "response")
    att_hf28_boot <- c(att_hf28_boot, mean(pred1_att_hf28_boot - pred0_att_hf28_boot))
  }
  
  # Calculate standard errors and confidence intervals
  SE_ATE <- sd(ate_hf28_boot)
  SE_ATT <- sd(att_hf28_boot)
  CI_ATE_lower <- ATE_hf28 - qnorm(0.975) * SE_ATE
  CI_ATE_upper <- ATE_hf28 + qnorm(0.975) * SE_ATE
  CI_ATT_lower <- ATT_hf28 - qnorm(0.975) * SE_ATT
  CI_ATT_upper <- ATT_hf28 + qnorm(0.975) * SE_ATT
  
  # Return results as separate CI bounds
  return(list(
    ATE = round(ATE_hf28, 3), SE_ATE = round(SE_ATE, 3), 
    CI_ATE_lower = round(CI_ATE_lower, 3), CI_ATE_upper = round(CI_ATE_upper, 3),
    ATT = round(ATT_hf28, 3), SE_ATT = round(SE_ATT, 3), 
    CI_ATT_lower = round(CI_ATT_lower, 3), CI_ATT_upper = round(CI_ATT_upper, 3)
  ))
}


full_model <-  paste('hospital_free_28 ~ corticobl * (', paste(names(tico_7485)[1:19], collapse = ' + '), ')')
full_hf28 <- lm(as.formula(full_model), data=tico_7485) 

# do stepwise selection
stepwise_hf28 <- stepAIC(full_hf28, direction = "both")
# summary(stepwise_hf28)

select_hf28 <- hospital_free_28 ~ corticobl + gender + age + copdhx + malignhx + immuno_supp + 
  bmi + newsbl + creatbl + crpbl + lymphbl + ordpulmbl + vl_rslt_log10 + corticobl:bmi +
  corticobl:lymphbl 
result_hf28 <- calculate_ATE_ATT(model_formula = select_hf28, data = tico_7485, num_bootstraps = 100)
# print(result_hf28)

output2 <- data.frame(
  Type = c("ATE", "ATT"),
  Estimate = c(result_hf28$ATE, result_hf28$ATT),
  SE = c(result_hf28$SE_ATE, result_hf28$SE_ATT),
  CI_Lower = c(result_hf28$CI_ATE_lower, result_hf28$CI_ATT_lower),
  CI_Upper = c(result_hf28$CI_ATE_upper, result_hf28$CI_ATT_upper)
)


kable(output2, caption = "ATE and ATT Estimates from Regression Model")
```

## Question 4
```{r,eval=F}
# a) Find the IPW standardized mean difference among treatment groups
ps_model <- glm(corticobl ~ gender + age + copdhx + malignhx + immuno_supp + 
  bmi + newsbl + creatbl + crpbl + lymphbl + ordpulmbl + vl_rslt_log10, family = binomial(), data = tico_7485)
tico_7485$ps <- predict(ps_model, type = "response")
tico_7485$weight <- with(tico_7485, ifelse(corticobl == "yes", 1 / ps, 1 / (1 - ps)))

dataSvy <- svydesign(ids = ~ 1, data = tico_7485, weights = ~ weight)

tabWeighted <- svyCreateTableOne(vars = names(tico_7485)[2:19], strata = "corticobl", data = dataSvy, test = FALSE)
t2 <- print(tabWeighted, smd = TRUE, showAllLevels = TRUE, varLabels = TRUE)

kable(t2, caption = "Summary Statistics Stratified by the use of corticosteroids using IPW")
```


```{r,eval=F}
# b) ATE and ATT with PSS
# ATE estimate
tico_7485$corticobl_numeric <- as.numeric(tico_7485$corticobl == "yes")
ps <- predict(ps_model, type = "response")

ps_quintile_ate <- cut(ps, breaks = quantile(ps, probs = seq(0, 1, by = 0.2)), 
                       include.lowest = TRUE, labels = FALSE)
te_quintile <- tapply(tico_7485$hospital_free_28[tico_7485$corticobl == 'yes'], 
                      ps_quintile_ate[tico_7485$corticobl == 'yes'], mean) -
               tapply(tico_7485$hospital_free_28[tico_7485$corticobl == 'no'], 
                      ps_quintile_ate[tico_7485$corticobl == 'no'], mean)
ATE_PSS <- sum(te_quintile * table(ps_quintile_ate) / nrow(tico_7485), na.rm = TRUE)

# ATT estimate
ps_quintile_att <- cut(ps, 
	breaks = c(0, quantile(ps[tico_7485$corticobl == 'yes'], p = c(0.2, 0.4, 0.6, 0.8)), 1), labels = 1:5)
nA <- nrow(tico_7485[tico_7485$corticobl == 'yes', ])
nAj <- table(ps_quintile_att[tico_7485$corticobl == 'yes'])
te_quintile_att <- tapply(tico_7485$hospital_free_28[tico_7485$corticobl == 'yes'], 
                         ps_quintile_att[tico_7485$corticobl == 'yes'], mean) -
	tapply(tico_7485$hospital_free_28[tico_7485$corticobl == 'no'], 
	       ps_quintile_att[tico_7485$corticobl == 'no'], mean)
ATT_PSS <- sum(te_quintile_att * nAj / nA, na.rm = TRUE)

# bootstrap to get the SE and CI
B <- 100
ATE_PSS.boot <- NULL
ATT_PSS.boot <- NULL
n <- nrow(tico_7485)

for(i in 1:B) {

  tico.boot <- tico_7485[sample(1:n, n, replace = TRUE), ]
  ps_model.boot <- glm(corticobl ~ gender + age + copdhx + malignhx + immuno_supp + 
  bmi + newsbl + creatbl + crpbl + lymphbl + ordpulmbl + vl_rslt_log10,
                 data = tico.boot, family = "binomial")
  
  ps.boot <- predict(ps_model.boot, type = "response")
  
  # ATE
  ps_quintile.boot <- cut(ps.boot,
                          breaks = quantile(ps.boot, probs = seq(0, 1, by = 0.2), na.rm = TRUE),
                          include.lowest = TRUE, labels = FALSE)
  
  te_quintile.boot <- tapply(tico.boot$hospital_free_28[tico.boot$corticobl == 'yes'],
                             ps_quintile.boot[tico.boot$corticobl == 'yes'], mean) -
    tapply(tico.boot$hospital_free_28[tico.boot$corticobl == 'no'],
           ps_quintile.boot[tico.boot$corticobl == 'no'], mean)
  
  nj.boot <- table(ps_quintile.boot)
  ATE.boot <- sum(te_quintile.boot * nj.boot / n, na.rm = TRUE)
  
  ATE_PSS.boot <- c(ATE_PSS.boot, ATE.boot)
  
  # ATT
  ps_quintile.boot_att <- cut(ps.boot,
                              breaks = c(0, quantile(ps.boot[tico.boot$corticobl == 'yes'], 
                                                     probs = c(0.2, 0.4, 0.6, 0.8), na.rm = TRUE), 1),
                              include.lowest = TRUE, labels = FALSE)
  
  nA.boot <- sum(tico.boot$corticobl == 'yes')
  nAj.boot <- table(ps_quintile.boot_att[tico.boot$corticobl == 'yes'])
  
  te_quintile_att.boot <- tapply(tico.boot$hospital_free_28[tico.boot$corticobl == 'yes'],
                                 ps_quintile.boot_att[tico.boot$corticobl == 'yes'], mean) -
    tapply(tico.boot$hospital_free_28[tico.boot$corticobl == 'no'],
           ps_quintile.boot_att[tico.boot$corticobl == 'no'], mean)
  
  ATT.boot <- sum(te_quintile_att.boot * nAj.boot / nA.boot, na.rm = TRUE)
  
  ATT_PSS.boot <- c(ATT_PSS.boot, ATT.boot)
}

SE_ATE_PSS <- sd(ATE_PSS.boot)
SE_ATT_PSS <- sd(ATT_PSS.boot)

CI_ATE_PSS <- ATE_PSS + c(-1, 1) * qnorm(0.975) * SE_ATE_PSS
CI_ATT_PSS <- ATT_PSS + c(-1, 1) * qnorm(0.975) * SE_ATT_PSS

output3 <- data.frame(
  Type = c("ATE", "ATT"),
  Estimate = c(ATE_PSS, ATT_PSS),
  SE = c(SE_ATE_PSS, SE_ATT_PSS),
  CI_Lower = c(CI_ATE_PSS[1], CI_ATT_PSS[1]),
  CI_Upper = c(CI_ATE_PSS[2], CI_ATT_PSS[2])
)

kable(output3, caption = "ATE and ATT Estimates using PSS")
```


```{r,eval=F}
# c) ATE and ATT estimates using IPW
# ATE
w1 <- as.numeric(tico_7485$corticobl_numeric)/ps
w0 <- (1-tico_7485$corticobl_numeric)/(1-ps)
ATE_IPW <- mean(tico_7485$hospital_free_28*w1) - mean(tico_7485$hospital_free_28*w0)

# ATT
w1 <- tico_7485$corticobl_numeric
w0 <- (1-tico_7485$corticobl_numeric)/(1-ps)*ps
ATT_IPW <- weighted.mean(tico_7485$hospital_free_28,w=w1) - 
	weighted.mean(tico_7485$hospital_free_28,w=w0)

# bootstrap
B <- 100
n <- nrow(tico_7485)
ATE_IPW.boot <- NULL
ATT_IPW.boot <- NULL

for(i in 1:B) {

  tico.boot <- tico_7485[sample(1:n, n, replace = TRUE), ]
  
  p1.boot <- glm(corticobl ~ gender + age + copdhx + malignhx + immuno_supp + bmi + newsbl + 
                   creatbl + crpbl + lymphbl + ordpulmbl + vl_rslt_log10, data = tico.boot, family = "binomial")
  ps.boot <- predict(p1.boot, type = "response")
  
  w1.boot <- as.numeric(tico.boot$corticobl_numeric) / ps.boot
  w0.boot <- (1 - tico.boot$corticobl_numeric) / (1 - ps.boot)
  
  ATE_IPW.boot <- c(ATE_IPW.boot, 
                    mean(tico.boot$hospital_free_28 * w1.boot) - 
                    mean(tico.boot$hospital_free_28 * w0.boot))
  
  w1_att.boot <- tico.boot$corticobl_numeric
  w0_att.boot <- (1 - tico.boot$corticobl_numeric) / (1 - ps.boot) * ps.boot
  
  ATT_IPW.boot <- c(ATT_IPW.boot, 
                    weighted.mean(tico.boot$hospital_free_28, w = w1_att.boot) - 
                    weighted.mean(tico.boot$hospital_free_28, w = w0_att.boot))
}

SE_ATE_IPW <- sd(ATE_IPW.boot)
CI_ATE_IPW <- ATE_IPW + c(-1, 1) * qnorm(0.975) * SE_ATE_IPW

SE_ATT_IPW <- sd(ATT_IPW.boot)
CI_ATT_IPW <- ATT_IPW + c(-1, 1) * qnorm(0.975) * SE_ATT_IPW

output4 <- data.frame(
  Type = c("ATE", "ATT"),
  Estimate = c(ATE_IPW, ATT_IPW),
  SE = c(SE_ATE_IPW, SE_ATT_IPW),
  CI_Lower = c(CI_ATE_IPW[1], CI_ATT_IPW[1]),
  CI_Upper = c(CI_ATE_IPW[2], CI_ATT_IPW[2])
)

kable(output4, caption = "ATE and ATT Estimates using IPW")
```


```{r,eval=F}
# d) 1:1 matching 
# i.Find the standardized mean difference among the matched cohort.
rownames(tico_7485)<-1:nrow(tico_7485)
mod_match<-matchit(corticobl ~ gender + age + copdhx + malignhx + immuno_supp + bmi + newsbl + creatbl + crpbl + lymphbl +
                     ordpulmbl + vl_rslt_log10,distance="logit",method="nearest",data=tico_7485,ratio=1,replace=TRUE)
#print(tail(mod_match$match.matrix))

matched_pairs <- na.omit(mod_match$match.matrix)  # to check if controls re used or not
n_matched <- nrow(matched_pairs)
print(n_matched) 

tico_match <- data.frame(rbind(tico_7485[tico_7485$corticobl == 'yes', ],
	tico_7485[as.numeric(mod_match$match.matrix), ]))

tico_match <- set_variable_labels(tico_match, asthmahx = "Asthma", 
                                 copdhx = 'COPD',
                                 diabeteshx = 'Diabetes requiring medication' ,
                                 chfhx = 'CHF',
                                 hypertenshx = 'Hypertension requiring medication', 
                                 malignhx = 'Malignancy (active or receiving treatment)',
                                 renalhx= 'Renal impairment',
                                 immuno_supp = 'Immunosuppressed',
                                 newsbl = 'New Score',
                                 creatbl = 'Creatinine (mg/dL)',
                                 crpbl='C-reactive protein',
                                 lymphbl = 'Lymphocytes (10^9/L)',
                                 ordpulmbl = 'Pulm ordinal', 
                                 vl_rslt_log10 = 'Log 10 of Viral load copies/mL')

tabmatched1.1 <- CreateTableOne(vars = names(tico_7485)[2:19], strata = "corticobl",
	data = tico_match, test = FALSE)
t3 <- print(tabmatched1.1, smd = TRUE, showAllLevels = TRUE, 
						varLabels = TRUE)

kable(t3, caption = "Summary Statistics Stratified by the use of corticosteroids using 1:1 Matching")

# ii.Estimate the ATT using 1:1 matching
p1 <- mean(tico_7485$hospital_free_28[tico_7485$corticobl == 'yes'])
p0 <- mean(tico_7485$hospital_free_28[as.numeric(mod_match$match.matrix)],na.rm=TRUE)
n1 <- table(tico_7485$corticobl)[2]
n0 <- length(as.numeric(mod_match$match.matrix))

ATT1match <- p1 -  p0
SE1match <- sqrt(-1*(p1*(1-p1)/n1 + p0*(1-p0)/n0))
CI1match <- p1 - p0 + c(-1, 1)*qnorm(0.975)*SE1match

output5 <- data.frame(
  Type = 'ATT',
  Estimate = round(ATT1match, 3),
  SE = round(unname(SE1match), 3),
  CI_Lower = round(CI1match[1], 3),
  CI_Upper = round(CI1match[2], 3)
)

kable(output5, caption = "ATT Estimate using 1:1 Matching")
```



