---
title: "Homework 6"
author: "Alex Ojemann"
date: "2024-11-22"
output: 
  pdf_document:
    toc: true       # Enable Table of Contents
    toc_depth: 2    # Set depth of TOC (e.g., 2 for sections and subsections)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

\newpage

\section{Problem 1}

\subsection{Summary}

The following model is a SARIMA(1,1,0)x(2,1,2)[10] on $X_t$.

```{r}
load('hw6dat.rsav')

# Load the necessary library
library(astsa)

# Fit the SARIMA(1,1,0)(4,1,0)[10] model on dat1 without generating plots
sarima_fit <- sarima(dat1, 
                     p = 1, d = 1, q = 0,  # Non-seasonal components
                     P = 2, D = 1, Q = 2, S = 10,  # Seasonal components
                     details = FALSE, quiet = TRUE)

# Extract and print the model summary
summary(sarima_fit$fit)
```

\subsection{Explanation}

Looking at the initial plot of this series, $X_t$ very clearly isn't stationary. The plot of the first difference of the series makes it clearer that there is also expanding variance as time increases, so a Box-Cox transformation was performed. Of the transformations plotted, the one in which $\lambda = 0.125$ looks the most stationary, so it was used for further analysis. However, when models were fit using this transformation, the models all showed extremely heavy tails in the normal QQ plot, so the original data without transformation was used. The ACF of $X_t$ shows clear seasonality with period 10. After taking a seasonal difference with s=10, the ACF appears to tail off both seasonally and non seasonally, and the PACF appears to cut off abruptly after non seasonal lad 1 and seasonal lag 1, so the initial model fit was a SARIMA(1,1,0)x(1,1,0)[10]. This model showed three significant lags h = 10, 20, and 30 in the ACF of the residuals and the LBP showed cumulative autocorrelation that was unaccounted for at all lags, so a different model was fit with three additional terms. None of these four models had any concerningly significant lags in the ACFs of their residuals, and their LBPs had 2-3 lags below the threshold but were much better than that of the initial SARIMA(1,1,0)x(1,1,0)[10] model. Of these four models, the SARIMA(1,1,0)x(1,1,3)[10] and the SARIMA(1,1,0)x(2,1,2)[10] had the best AIC, and the SARIMA(1,1,0)x(2,1,2)[10] had fewer insignificant parameters at the p<0.01 level so it was used as the final model.

\subsection{Output}

Plots of unadjusted and differenced series $X_t$.

```{r}
plot(dat1, main = "Time Series - dat1", xlab = "Time", ylab = "Value", type = "l", col = "blue")
# Difference the dat2 series to make it stationary
dat1_diff <- diff(dat1)

# Plot the differenced dat2 time series
plot(dat1_diff, main = "Time Series - dat1 (Differenced)", xlab = "Time", ylab = "Differenced Value", type = "l", col = "red")
```
Plot of Box-Cox transformations of the differenced $X_t$ with varying $\lambda$ (no transformations ended up being used due to heavy tails in Normal QQ plot of models using transformed data):

```{r}
# Load the necessary library
library(forecast)

# Define lambda values to try for Box-Cox transformation
lambda_values <- c(-0.125, 0, 0.125,0.25, 0.375,0.5)  # 0 corresponds to log transformation

# Set up the plotting area
par(mfrow = c(2, 3))  # Arrange plots in a grid (2 rows, 3 columns)

# Loop through lambda values, apply the Box-Cox transformation, and plot
for (lambda in lambda_values) {
  # Apply the Box-Cox transformation
  if (lambda == 0) {
    transformed_dat1_diff <- log(dat1_diff)  # Log transformation
    title <- "Log Transformation"
  } else {
    transformed_dat1_diff <- (dat1_diff^lambda - 1) / lambda  # Box-Cox formula
    title <- paste("Lambda =", lambda)
  }
  
  # Plot the transformed series
  plot(transformed_dat1_diff, type = "l", main = title, xlab = "Time", ylab = "Transformed Value")
}

# Reset plotting area to default
par(mfrow = c(1, 1))

```
ACF and PACF of $X_t$

```{r}
# Load the necessary library
library(forecast)

# Define lambda for the Box-Cox transformation
lambda <- 0.125

# Shift dat1_diff to ensure all values are positive
dat1_diff <- dat1_diff - min(dat1_diff) + 1  # Add 1 to avoid zeros

# Apply the Box-Cox transformation to dat1_diff
dat1_diff_transformed <- (dat1_diff^lambda - 1) / lambda

# Plot the ACF of the transformed series
acf(dat1_diff_transformed, main = "ACF of dat1_diff (Lambda = 0.125)")

# Plot the PACF of the transformed series
pacf(dat1_diff_transformed, main = "PACF of dat1_diff (Lambda = 0.125)")

```

ACF and PACF of seasonally differenced series:

```{r}
# Perform seasonal differencing with s = 10
dat1_seasonal_diff <- diff(dat1_diff, lag = 10)

# Plot the seasonally differenced series
plot(dat1_seasonal_diff, main = "Seasonally Differenced Series (s = 10)", 
     xlab = "Time", ylab = "Differenced Value", type = "l", col = "blue")

acf(dat1_seasonal_diff, main = "ACF of Seasonally Differenced Series (s = 10)")

# Plot the PACF of the seasonally differenced series
pacf(dat1_seasonal_diff, main = "PACF of Seasonally Differenced Series (s = 10)")

```

First model attempt (SARIMA(1,1,0)x(1,1,0)[10])

```{r}
# Load the necessary libraries
library(astsa)
library(forecast)

# Fit the SARIMA(1, 0, 0) x (1, 0, 0)[10] model to dat1_seasonal_diff
sarima_fit <- sarima(dat1_seasonal_diff, p = 1, d = 0, q = 0, 
                     P = 1, D = 0, Q = 0, S = 10)

# Diagnostic plots and model summary are automatically generated

```

Additional models fit with three more seasonal terms than the SARIMA(1,1,0)x(1,1,0)[10]:

```{r}
# Load necessary library
library(astsa)

# Initialize a list to store AIC values
model_aic <- list()

# Loop through combinations where added seasonal terms sum to 3
for (P_added in 0:3) {  # Added seasonal AR terms
  Q <- 3 - P_added      # Seasonal MA terms (to make P_added + Q = 3)
  
  # Model parameters
  P <- 1 + P_added  # Total seasonal AR terms
  model_name <- paste0("SARIMA(1,0,0)(", P, ",0,", Q, ")[10]")
  cat("\nFitting model: ", model_name, "\n")
  
  # Fit the SARIMA model
  sarima_fit <- sarima(dat1_seasonal_diff, 
                       p = 1, d = 0, q = 0, 
                       P = P, D = 0, Q = Q, S = 10)
  
  # Calculate AIC manually
  loglik <- sarima_fit$fit$loglik  # Log-likelihood from the fit
  n_params <- length(sarima_fit$fit$coef)  # Number of estimated parameters
  aic_value <- -2 * loglik + 2 * n_params  # AIC formula
  
  # Store the AIC value
  model_aic[[model_name]] <- aic_value
  cat("AIC for", model_name, ":", aic_value, "\n")
}

# Print the AIC values for all models
cat("\nAIC values for all models:\n")
print(model_aic)

```


