---
title: "PubH7485 Homework3 Key"
date: "2024-10-05"
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(7485)
library(dplyr)
library(labelled)
library(causaldata)
library(knitr)
library(formattable)
library(MASS)
library(reshape2)
library(ggplot2)
library(boot) 
library(randomForest)
library(survey)
library(tableone)

```

```{r, include=F}
data('nhefs')

# treatment
nhefs$qsmk 
# outcomes of interest
nhefs$wt82_71
nhefs$sbp
nhefs$dbp

all_covars <- setdiff(names(nhefs), c('dadth', 'death', 'hightax82', 'modth', 'price71_82', 
                                      'price82', 'seqn', 'yrdth', 'wt82', 'tax71_82',
                                      'tax82', 'smokeyrs', 'smkintensity82_71',
                                      'birthplace','id','censored','older'))

adj_vars <- c('sex', 'race', 'age', 'education', 'smokeintensity', 'smokeyrs', 'exercise', 'active', 'wt71', 'hbp')


# Convert categorical variables to factors and set levels for them in nhefs data frame
nhefs$active <- factor(nhefs$active, levels = c(0, 1, 2), labels = c("Very active", "Moderately active", "Inactive"))
nhefs$alcoholfreq <- factor(nhefs$alcoholfreq, levels = c(0, 1, 2, 3, 4, 5), labels = c("Almost every day", "2-3 times/week", "1-4 times/month", "< 12 times/year", "No alcohol last year", "Unknown"))
nhefs$alcoholpy <- factor(nhefs$alcoholpy, levels = c(0, 1, 2), labels = c("Never", "Ever", "Missing"))
nhefs$alcoholtype <- factor(nhefs$alcoholtype, levels = c(1, 2, 3, 4), labels = c("Beer", "Wine", "Liquor", "Other/Unknown"))
nhefs$allergies <- factor(nhefs$allergies, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$asthma <- factor(nhefs$asthma, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$birthcontrol <- factor(nhefs$birthcontrol, levels = c(0, 1, 2), labels = c("No", "Yes", "Missing"))
nhefs$boweltrouble <- factor(nhefs$boweltrouble, levels = c(0, 1, 2), labels = c("Never", "Ever", "Missing"))
nhefs$bronch <- factor(nhefs$bronch, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$chroniccough <- factor(nhefs$chroniccough, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$colitis <- factor(nhefs$colitis, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$diabetes <- factor(nhefs$diabetes, levels = c(0, 1, 2), labels = c("Never", "Ever", "Missing"))
nhefs$education <- factor(nhefs$education, levels = c(1, 2, 3, 4, 5), labels = c("8th grade or less", "HS dropout", "HS", "College dropout", "College or more"))
nhefs$exercise <- factor(nhefs$exercise, levels = c(0, 1, 2), labels = c("Little or no exercise", "Moderate exercise", "Much exercise"))
nhefs$hayfever <- factor(nhefs$hayfever, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$hbp <- factor(nhefs$hbp, levels = c(0, 1, 2), labels = c("Never", "Ever", "Missing"))
nhefs$hbpmed <- factor(nhefs$hbpmed, levels = c(0, 1, 2), labels = c("Never", "Ever", "Missing"))
nhefs$headache <- factor(nhefs$headache, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$hepatitis <- factor(nhefs$hepatitis, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$hf <- factor(nhefs$hf, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$marital <- factor(nhefs$marital, levels = c(1, 2, 3, 4, 5, 6, 8), labels = c("Under 17", "Married", "Widowed", "Never married", "Divorced", "Separated", "Unknown"))
nhefs$nerves <- factor(nhefs$nerves, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$nervousbreak <- factor(nhefs$nervousbreak, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$otherpain <- factor(nhefs$otherpain, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$pepticulcer <- factor(nhefs$pepticulcer, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$pica <- factor(nhefs$pica, levels = c(0, 1, 2), labels = c("Never", "Ever", "Missing"))
nhefs$polio <- factor(nhefs$polio, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$qsmk <- factor(nhefs$qsmk, levels = c(0, 1), labels = c("No", "Yes"))
nhefs$race <- factor(nhefs$race, levels = c(0, 1), labels = c("White", "Black or Other"))
nhefs$sex <- factor(nhefs$sex, levels = c(0, 1), labels = c("Male", "Female"))
nhefs$tb <- factor(nhefs$tb, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$tumor <- factor(nhefs$tumor, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$weakheart <- factor(nhefs$weakheart, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$wtloss <- factor(nhefs$wtloss, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$lackpep <- factor(nhefs$lackpep, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$infection <- factor(nhefs$infection, levels = c(0, 1), labels = c("Never", "Ever"))


mode_function <- function(x) {
  x <- as.character(x)
  tbl <- table(x, useNA = "ifany")
  as.character(names(tbl)[which.max(tbl)])
}

for (var in all_covars) {
  if (is.numeric(nhefs[[var]])) {
    # Impute missing values for numeric variables with the median
    nhefs[[var]][is.na(nhefs[[var]])] <- median(nhefs[[var]], na.rm = TRUE)
    print(paste("Imputed numeric variable:", var))
    
  } else if (is.factor(nhefs[[var]]) || is.character(nhefs[[var]])) {
    # Impute missing values for categorical variables with the mode
    nhefs[[var]][is.na(nhefs[[var]])] <- mode_function(nhefs[[var]])
    # nhefs[[var]] <- as.factor(nhefs[[var]])  # Ensure it's still a factor
    print(paste("Imputed categorical variable:", var))
  }
}

```



## Question 1 (30 points total, 20 points for the plot, 10 points for the choice of model and justification)

From the plot, we can see that the propensity score weighting is useful in achieving balance. Generally, Model b  with the nonlinear terms seems to work the best, with all SMDs less than 0.1 and having a smallest average SMD of 0.0164. 

(The key is using random forest to select the set of covariates for Model d, however, any optimal method is acceptable.)

```{r,echo=F}
# (a) The logistic regression model with just main effects.
# Logistic regression with main effects
ps_model_a <- glm(qsmk ~ sex + race + age + education + smokeintensity + smokeyrs + exercise + active + wt71 + hbp, family = binomial(), data = nhefs)
nhefs$ps_a <- predict(ps_model_a, type = "response")
nhefs$weight_a <- with(nhefs, ifelse(qsmk == "Yes", 1 / ps_a, 1 / (1 - ps_a)))

# Calculate weighted and unweighted SMDs
dataSvy_a <- svydesign(ids = ~ 1, data = nhefs, weights = ~ weight_a)
tabUnweighted_a <- CreateTableOne(vars = adj_vars, strata = "qsmk", data = nhefs, test = FALSE)
tabWeighted_a <- svyCreateTableOne(vars = adj_vars, strata = "qsmk", data = dataSvy_a, test = FALSE)

# Extracting SMD values
dataPlot_a <- data.frame(variable  = rownames(ExtractSmd(tabUnweighted_a)),
                       Unweighted = as.numeric(ExtractSmd(tabUnweighted_a)),
                       Weighted_a  = as.numeric(ExtractSmd(tabWeighted_a)))

# Convert to long format for ggplot
dataPlotMelt_a <- melt(data = dataPlot_a,
                     id.vars = c("variable"),
                     variable.name = "Method",
                     value.name = "SMD")

# (b) The same set of covariates as in (a) but allow for nonlinear terms in the continuous covariates. 
# Logistic regression with main effects and nonlinear terms in the continuous covariates
ps_model_b <- glm(qsmk ~ sex + race + poly(age, 2) + education + poly(smokeintensity, 2) + poly(smokeyrs, 2) + exercise + active + poly(wt71, 2) + hbp, family = binomial(), data = nhefs)
nhefs$ps_b <- predict(ps_model_b, type = "response")
nhefs$weight_b <- with(nhefs, ifelse(qsmk == "Yes", 1 / ps_b, 1 / (1 - ps_b)))

# Calculate weighted and unweighted SMDs
dataSvy_b <- svydesign(ids = ~ 1, data = nhefs, weights = ~ weight_b)
tabUnweighted_b <- CreateTableOne(vars = adj_vars, strata = "qsmk", data = nhefs, test = FALSE)
tabWeighted_b <- svyCreateTableOne(vars = adj_vars, strata = "qsmk", data = dataSvy_b, test = FALSE)

# Extracting SMD values
dataPlot_b <- data.frame(variable  = rownames(ExtractSmd(tabUnweighted_b)),
                       Unweighted = as.numeric(ExtractSmd(tabUnweighted_b)),
                       Weighted_b  = as.numeric(ExtractSmd(tabWeighted_b)))

# Convert to long format for ggplot
dataPlotMelt_b <- melt(data = dataPlot_b,
                     id.vars = c("variable"),
                     variable.name = "Method",
                     value.name = "SMD")



# (c) The same set of covariates as in (a) but allowing for all pairwise interactions and performing some form of variable selection (e.g., forward selection using AIC).
# Allowing for all pairwise interactions and performing some form of variable selection (e.g., forward selection using AIC).
full_model_c <- glm(qsmk ~ (sex + race + age + education + smokeintensity + smokeyrs + exercise + active + wt71 + hbp)^2, family = binomial(), data = nhefs)
ps_model_c <- step(full_model_c, trace=0, direction="forward", criteria="AIC")
nhefs$ps_c <- predict(ps_model_c, type="response")
nhefs$weight_c <- with(nhefs, ifelse(qsmk == "Yes", 1 / ps_c, 1 / (1 - ps_c)))

# Calculate weighted and unweighted SMDs
dataSvy_c <- svydesign(ids = ~ 1, data = nhefs, weights = ~ weight_c)
tabUnweighted_c <- CreateTableOne(vars = adj_vars, strata = "qsmk", data = nhefs, test = FALSE)
tabWeighted_c <- svyCreateTableOne(vars = adj_vars, strata = "qsmk", data = dataSvy_c, test = FALSE)

# Extracting SMD values
dataPlot_c <- data.frame(variable  = rownames(ExtractSmd(tabUnweighted_c)),
                       Unweighted = as.numeric(ExtractSmd(tabUnweighted_c)),
                       Weighted_c = as.numeric(ExtractSmd(tabWeighted_c)))

# Convert to long format for ggplot
dataPlotMelt_c <- melt(data = dataPlot_c,
                     id.vars = c("variable"),
                     variable.name = "Method",
                     value.name = "SMD")


# (d) A logistic regression model considering all available baseline covariates and some form of variable selection.
# Fit a random forest model and select top 10 important covariates
rf_model <- randomForest(as.formula(paste("qsmk ~", paste(all_covars[-c(1:3,13)], collapse = " + "))), data = nhefs)
important_vars <- importance(rf_model)[order(importance(rf_model)[, "MeanDecreaseGini"], decreasing = TRUE), ]
top_10_vars <- row.names(as.data.frame(important_vars[1:10]))

formula_d <- as.formula(paste("qsmk ~", paste(top_10_vars, collapse=" + ")))

ps_model_d <- glm(formula_d, family = binomial(), data = nhefs)
nhefs$ps_d <- predict(ps_model_d, type = "response")
nhefs$weight_d <- with(nhefs, ifelse(qsmk == "Yes", 1 / ps_d, 1 / (1 - ps_d)))

# Calculate weighted and unweighted SMDs
dataSvy_d <- svydesign(ids = ~ 1, data = nhefs, weights = ~ weight_d)
tabUnweighted_d <- CreateTableOne(vars = adj_vars, strata = "qsmk", data = nhefs, test = FALSE)
tabWeighted_d <- svyCreateTableOne(vars = adj_vars, strata = "qsmk", data = dataSvy_d, test = FALSE)

# Extracting SMD values
dataPlot_d <- data.frame(variable  = rownames(ExtractSmd(tabUnweighted_d)),
                       Unweighted = as.numeric(ExtractSmd(tabUnweighted_d)),
                       Weighted_d  = as.numeric(ExtractSmd(tabWeighted_d)))

# Convert to long format for ggplot
dataPlotMelt_d <- melt(data = dataPlot_d,
                     id.vars = c("variable"),
                     variable.name = "Method",
                     value.name = "SMD")
dataPlotMelt <- rbind(dataPlotMelt_a, dataPlotMelt_b[11:20, ], dataPlotMelt_c[11:20, ], dataPlotMelt_d[11:20, ])
dataPlotMelt %>%
  arrange(desc(SMD), variable) %>%
  mutate(variable_ordered = factor(variable, levels = unique(variable))) %>%
  ggplot(aes(x = variable_ordered, y = SMD, group = Method, color = Method)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0.1, color = "black", linewidth = 0.1) +
  coord_flip() +
  theme_bw() +
  theme(legend.key = element_blank()) +  xlab('Variables') + 
  labs(title="Standardized Mean Differences for all Models")
```


```{r,echo=F}
#(e) Select one of these models to use in question 2 and justify your choice.
# Calculating average SMD for each model
average_smd_a <- mean(dataPlot_a$Weighted)
average_smd_b <- mean(dataPlot_b$Weighted)
average_smd_c <- mean(dataPlot_c$Weighted)
average_smd_d <- mean(dataPlot_d$Weighted)

# Combining the averages into a single data frame
smd_averages <- data.frame(
  Model = c("Model A", "Model B", "Model C", "Model D"),
  Average_SMD = c(average_smd_a, average_smd_b, average_smd_c, average_smd_d)
)

kable(smd_averages)

```


## Question 2 (30 points total, 10 points for each outcome)

The calculated results are summarized in the following tables: 

```{r, echo=F, warning=F}
nhefs$qsmk_numeric <- as.numeric(nhefs$qsmk == "Yes")

calculate_effects <- function(data, outcome, formula) {
  # Fit propensity score model
  ps_model <- glm(formula, family = binomial(), data = data)
  data$ps <- predict(ps_model, type = "response")
  
  # pss
  data$ps_quintile <- cut(data$ps, breaks = quantile(data$ps, probs = seq(0, 1, by = 0.2)), 
                          include.lowest = TRUE, labels = FALSE)
  
  te_quintile <- tapply(data[[outcome]][data$qsmk_numeric == 1], 
                        data$ps_quintile[data$qsmk_numeric == 1], mean, na.rm = TRUE) -
                 tapply(data[[outcome]][data$qsmk_numeric == 0], 
                        data$ps_quintile[data$qsmk_numeric == 0], mean, na.rm = TRUE)
  PSS <- sum(te_quintile * table(data$ps_quintile) / nrow(data), na.rm = TRUE)
  # ATE_PSS_wt82_71 <- sum(effects_wt82_71 * nj / nrow(nhefs), na.rm = TRUE)
  
  # IPW calculations
  data$w1 <- data$qsmk_numeric / data$ps
  data$w0 <- (1 - data$qsmk_numeric) / (1 - data$ps)
  IPW1 <- mean(data[[outcome]] * data$w1) - mean(data[[outcome]] * data$w0)
  IPW2 <- weighted.mean(data[[outcome]], data$w1) - weighted.mean(data[[outcome]], data$w0)
  
  B <- 1000
  nj <- table(nhefs$ps_quintile)
  ATE_PSS.boot <- NULL
  ATE_IPW.boot <- NULL
  ATE_IPW2.boot <- NULL
  n <- nrow(nhefs)
  
  for(i in 1:B) {
    
    data.boot <- data[sample(1:n, n, replace = TRUE), ]
    p1.boot <- glm(formula, data = data.boot, family = "binomial")
    ps.boot <- predict(p1.boot, type = "response")
    
    ps_quintile.boot <- cut(ps.boot, 
  		breaks = c(0, quantile(ps.boot, p = c(0.2, 0.4, 0.6, 0.8)), 1), labels = 1:5)
    
    nj.boot <- table(ps_quintile.boot)
    
  	te_quintile.boot <- tapply(data.boot[[outcome]][data.boot$qsmk_numeric == 1], 
  		ps_quintile.boot[data.boot$qsmk_numeric == 1], mean) -
  	tapply(data.boot[[outcome]][data.boot$qsmk_numeric == 0], 
  	       ps_quintile.boot[data.boot$qsmk_numeric == 0], mean)
  	
  	ATE.boot <- sum(te_quintile.boot * nj.boot/ nrow(data.boot), na.rm = TRUE)
    ATE_PSS.boot <- c(ATE_PSS.boot, ATE.boot)

    w1.boot <- data.boot$qsmk_numeric/ps.boot
    w0.boot <- (1-data.boot$qsmk_numeric)/(1-ps.boot)
    
    ATE_IPW.boot <- c(ATE_IPW.boot, 
      mean(data.boot[[outcome]]*w1.boot) - mean(data.boot[[outcome]]*w0.boot))
    ATE_IPW2.boot <- c(ATE_IPW2.boot, 
      weighted.mean(data.boot[[outcome]], w1.boot) - weighted.mean(data.boot[[outcome]], w0.boot))
    
  }
  
  results <- data.frame(
    ATE = round(c(PSS, IPW1, IPW2), 3),  # Round ATE values to 3 decimal places
  `Standard Error` = round(c(sd(ATE_PSS.boot), sd(ATE_IPW.boot), sd(ATE_IPW2.boot)), 3),  
  `95% CI` = c(
    sprintf("%.3f, %.3f", PSS - qnorm(0.975) * sd(ATE_PSS.boot), PSS + qnorm(0.975) * sd(ATE_PSS.boot)),
    sprintf("%.3f, %.3f", IPW1 - qnorm(0.975) * sd(ATE_IPW.boot), IPW1 + qnorm(0.975) * sd(ATE_IPW.boot)),
    sprintf("%.3f, %.3f", IPW2 - qnorm(0.975) * sd(ATE_IPW2.boot), IPW2 + qnorm(0.975) * sd(ATE_IPW2.boot))
  )
  )
  rownames(results) <- c("PSS", "IPW1", "IPW2")
  colnames(results)[3] <- "95% CI"
  colnames(results)[2] <- "Standard Error"
  
  return(results)
}

formula_test <- "qsmk ~ sex + race + poly(age, 2) + education + poly(smokeintensity, 2) + poly(smokeyrs, 2) + exercise + active + poly(wt71, 2) + hbp"
kable(calculate_effects(nhefs, "wt82_71",  formula_test),caption = 'Results for Weight Change')
kable(calculate_effects(nhefs, "sbp",  formula_test),caption = 'Results for SBP')
kable(calculate_effects(nhefs, "dbp",  formula_test),caption = 'Results for DBP')
```

Note that depend on different models you are using, the results may be a little different. 


# Appendix - R code

## Question 1 

```{r, eval=F}
### Logistic regression with main effects
ps_model_a <- glm(qsmk ~ sex + race + age + education + smokeintensity + smokeyrs + exercise + active + wt71 + hbp, family = binomial(), data = nhefs)
nhefs$ps_a <- predict(ps_model_a, type = "response")
nhefs$weight_a <- with(nhefs, ifelse(qsmk == "Yes", 1 / ps_a, 1 / (1 - ps_a)))

# Calculate weighted and unweighted SMDs
dataSvy_a <- svydesign(ids = ~ 1, data = nhefs, weights = ~ weight_a)
tabUnweighted_a <- CreateTableOne(vars = adj_vars, strata = "qsmk", data = nhefs, test = FALSE)
tabWeighted_a <- svyCreateTableOne(vars = adj_vars, strata = "qsmk", data = dataSvy_a, test = FALSE)

dataPlot_a <- data.frame(variable  = rownames(ExtractSmd(tabUnweighted_a)),
                       Unweighted = as.numeric(ExtractSmd(tabUnweighted_a)),
                       Weighted_a  = as.numeric(ExtractSmd(tabWeighted_a)))

# Convert to long format for ggplot
dataPlotMelt_a <- melt(data = dataPlot_a,
                     id.vars = c("variable"),
                     variable.name = "Method",
                     value.name = "SMD")

### Logistic regression with main effects and nonlinear terms in the continuous covariates
ps_model_b <- glm(qsmk ~ sex + race + poly(age, 2) + education + poly(smokeintensity, 2) + poly(smokeyrs, 2) + exercise + active + poly(wt71, 2) + hbp, family = binomial(), data = nhefs)
nhefs$ps_b <- predict(ps_model_b, type = "response")
nhefs$weight_b <- with(nhefs, ifelse(qsmk == "Yes", 1 / ps_b, 1 / (1 - ps_b)))

dataSvy_b <- svydesign(ids = ~ 1, data = nhefs, weights = ~ weight_b)
tabUnweighted_b <- CreateTableOne(vars = adj_vars, strata = "qsmk", data = nhefs, test = FALSE)
tabWeighted_b <- svyCreateTableOne(vars = adj_vars, strata = "qsmk", data = dataSvy_b, test = FALSE)

dataPlot_b <- data.frame(variable  = rownames(ExtractSmd(tabUnweighted_b)),
                       Unweighted = as.numeric(ExtractSmd(tabUnweighted_b)),
                       Weighted_b  = as.numeric(ExtractSmd(tabWeighted_b)))

dataPlotMelt_b <- melt(data = dataPlot_b,
                     id.vars = c("variable"),
                     variable.name = "Method",
                     value.name = "SMD")

### Allowing for all pairwise interactions and performing some form of variable selection (e.g., forward selection using AIC).
full_model_c <- glm(qsmk ~ (sex + race + age + education + smokeintensity + smokeyrs + exercise + active + wt71 + hbp)^2, family = binomial(), data = nhefs)
ps_model_c <- step(full_model_c, trace=0, direction="forward", criteria="AIC")
nhefs$ps_c <- predict(ps_model_c, type="response")
nhefs$weight_c <- with(nhefs, ifelse(qsmk == "Yes", 1 / ps_c, 1 / (1 - ps_c)))

dataSvy_c <- svydesign(ids = ~ 1, data = nhefs, weights = ~ weight_c)
tabUnweighted_c <- CreateTableOne(vars = adj_vars, strata = "qsmk", data = nhefs, test = FALSE)
tabWeighted_c <- svyCreateTableOne(vars = adj_vars, strata = "qsmk", data = dataSvy_c, test = FALSE)

dataPlot_c <- data.frame(variable  = rownames(ExtractSmd(tabUnweighted_c)),
                       Unweighted = as.numeric(ExtractSmd(tabUnweighted_c)),
                       Weighted_c = as.numeric(ExtractSmd(tabWeighted_c)))

dataPlotMelt_c <- melt(data = dataPlot_c,
                     id.vars = c("variable"),
                     variable.name = "Method",
                     value.name = "SMD")


### A logistic regression model considering all available baseline covariates and fit a random forest model to select top 10 important covariates
rf_model <- randomForest(as.formula(paste("qsmk ~", paste(all_covars[-c(1:3,13)], collapse = " + "))), data = nhefs)
important_vars <- importance(rf_model)[order(importance(rf_model)[, "MeanDecreaseGini"], decreasing = TRUE), ]
top_10_vars <- row.names(as.data.frame(important_vars[1:10]))

formula_d <- as.formula(paste("qsmk ~", paste(top_10_vars, collapse=" + ")))

ps_model_d <- glm(formula_d, family = binomial(), data = nhefs)
nhefs$ps_d <- predict(ps_model_d, type = "response")
nhefs$weight_d <- with(nhefs, ifelse(qsmk == "Yes", 1 / ps_d, 1 / (1 - ps_d)))

dataSvy_d <- svydesign(ids = ~ 1, data = nhefs, weights = ~ weight_d)
tabUnweighted_d <- CreateTableOne(vars = adj_vars, strata = "qsmk", data = nhefs, test = FALSE)
tabWeighted_d <- svyCreateTableOne(vars = adj_vars, strata = "qsmk", data = dataSvy_d, test = FALSE)

dataPlot_d <- data.frame(variable  = rownames(ExtractSmd(tabUnweighted_d)),
                       Unweighted = as.numeric(ExtractSmd(tabUnweighted_d)),
                       Weighted_d  = as.numeric(ExtractSmd(tabWeighted_d)))

dataPlotMelt_d <- melt(data = dataPlot_d,
                     id.vars = c("variable"),
                     variable.name = "Method",
                     value.name = "SMD")

# Prepare dataset for ggplot
dataPlotMelt <- rbind(dataPlotMelt_a, dataPlotMelt_b[11:20, ], dataPlotMelt_c[11:20, ], dataPlotMelt_d[11:20, ])
dataPlotMelt %>%
  arrange(desc(SMD), variable) %>%
  mutate(variable_ordered = factor(variable, levels = unique(variable))) %>%
  ggplot(aes(x = variable_ordered, y = SMD, group = Method, color = Method)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0.1, color = "black", linewidth = 0.1) +
  coord_flip() +
  theme_bw() +
  theme(legend.key = element_blank()) +  xlab('Variables') + 
  labs(title="Standardized Mean Differences for all Models")
```

## Question 2

```{r, eval=F}
nhefs$qsmk_numeric <- as.numeric(nhefs$qsmk == "Yes")

calculate_effects <- function(data, outcome, formula) {
  
  # Fit propensity score model
  ps_model <- glm(formula, family = binomial(), data = data)
  data$ps <- predict(ps_model, type = "response")
  
  # pss
  data$ps_quintile <- cut(data$ps, breaks = quantile(data$ps, probs = seq(0, 1, by = 0.2)), 
                          include.lowest = TRUE, labels = FALSE)
  
  te_quintile <- tapply(data[[outcome]][data$qsmk_numeric == 1], 
                        data$ps_quintile[data$qsmk_numeric == 1], mean, na.rm = TRUE) -
                 tapply(data[[outcome]][data$qsmk_numeric == 0], 
                        data$ps_quintile[data$qsmk_numeric == 0], mean, na.rm = TRUE)
  PSS <- sum(te_quintile * table(data$ps_quintile) / nrow(data), na.rm = TRUE)
  
  # IPW
  data$w1 <- data$qsmk_numeric / data$ps
  data$w0 <- (1 - data$qsmk_numeric) / (1 - data$ps)
  IPW1 <- mean(data[[outcome]] * data$w1) - mean(data[[outcome]] * data$w0)
  IPW2 <- weighted.mean(data[[outcome]], data$w1) - weighted.mean(data[[outcome]], data$w0)
  
  # Perform bootstrap 
  B <- 1000
  nj <- table(nhefs$ps_quintile)
  ATE_PSS.boot <- NULL
  ATE_IPW.boot <- NULL
  ATE_IPW2.boot <- NULL
  n <- nrow(nhefs)
  
  for(i in 1:B) {
    
    data.boot <- data[sample(1:n, n, replace = TRUE), ]
    p1.boot <- glm(formula, data = data.boot, family = "binomial")
    ps.boot <- predict(p1.boot, type = "response")
    
    ps_quintile.boot <- cut(ps.boot, 
  		breaks = c(0, quantile(ps.boot, p = c(0.2, 0.4, 0.6, 0.8)), 1), labels = 1:5)
    
    nj.boot <- table(ps_quintile.boot)
    
  	te_quintile.boot <- tapply(data.boot[[outcome]][data.boot$qsmk_numeric == 1], 
  		ps_quintile.boot[data.boot$qsmk_numeric == 1], mean) -
  	tapply(data.boot[[outcome]][data.boot$qsmk_numeric == 0], 
  	       ps_quintile.boot[data.boot$qsmk_numeric == 0], mean)
  	
  	ATE.boot <- sum(te_quintile.boot * nj.boot/ nrow(data.boot), na.rm = TRUE)
    ATE_PSS.boot <- c(ATE_PSS.boot, ATE.boot)

    w1.boot <- data.boot$qsmk_numeric/ps.boot
    w0.boot <- (1-data.boot$qsmk_numeric)/(1-ps.boot)
    
    ATE_IPW.boot <- c(ATE_IPW.boot, 
      mean(data.boot[[outcome]]*w1.boot) - mean(data.boot[[outcome]]*w0.boot))
    ATE_IPW2.boot <- c(ATE_IPW2.boot, 
      weighted.mean(data.boot[[outcome]], w1.boot) - weighted.mean(data.boot[[outcome]], w0.boot))
    
  }
  
  results <- data.frame(
    ATE = round(c(PSS, IPW1, IPW2), 3),  # Round ATE values to 3 decimal places
  `Standard Error` = round(c(sd(ATE_PSS.boot), sd(ATE_IPW.boot), sd(ATE_IPW2.boot)), 3),  
  `95% CI` = c(
    sprintf("%.3f, %.3f", PSS - qnorm(0.975) * sd(ATE_PSS.boot), PSS + qnorm(0.975) * sd(ATE_PSS.boot)),
    sprintf("%.3f, %.3f", IPW1 - qnorm(0.975) * sd(ATE_IPW.boot), IPW1 + qnorm(0.975) * sd(ATE_IPW.boot)),
    sprintf("%.3f, %.3f", IPW2 - qnorm(0.975) * sd(ATE_IPW2.boot), IPW2 + qnorm(0.975) * sd(ATE_IPW2.boot))
  )
  )
  
  rownames(results) <- c("PSS", "IPW1", "IPW2")
  colnames(results)[3] <- "95% CI"
  colnames(results)[2] <- "Standard Error"
  
  return(results)
}

formula_test <- "qsmk ~ sex + race + poly(age, 2) + education + poly(smokeintensity, 2) + poly(smokeyrs, 2) + exercise + active + poly(wt71, 2) + hbp"
kable(calculate_effects(nhefs, "wt82_71",  formula_test),caption = 'Result for Weight Change')
kable(calculate_effects(nhefs, "sbp",  formula_test),caption = 'Result for SBP')
kable(calculate_effects(nhefs, "dbp",  formula_test),caption = 'Result for DBP')
```
