---
title: "Homework 1"
author: "Alex Ojemann"
date: "2024-09-30"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(tinytex)
```

## Problem 1

$$\gamma(0) = 6 * \sigma^2$$
$$\gamma(1) = \gamma(-1) = 4*\sigma^2$$
$$\rho(0) = 6 * \sigma^2 / 6 * \sigma^2 = 1$$

$$\rho(1) = \rho(-1) = 4*\sigma^2 / 6 * \sigma^2 = 2/3$$  
```{r,echo=FALSE}
# Set the seed for reproducibility
set.seed(123)

# Generate n = 100 observations from white noise Wt ~ N(0,1)
n <- 100
W <- rnorm(n, mean = 0, sd = 1)

# Compute Xt = Wt+1 + 2Wt + Wt-1
X <- rep(0, n)  # Preallocate space for Xt

for (t in 2:(n-1)) {
  X[t] <- W[t-1] + 2*W[t] + W[t+1]
}

# Function to compute the ACF "by hand" for lags from -max_lag to max_lag
compute_acf_by_hand <- function(X, max_lag) {
  n <- length(X)
  X_mean <- mean(X)
  acf_vals <- numeric(2 * max_lag + 1)  # Preallocate space for ACF values
  
  # Compute ACF for lags from -max_lag to max_lag
  for (h in -max_lag:max_lag) {
    if (h >= 0) {
      cov_h <- sum((X[1:(n-h)] - X_mean) * (X[(h+1):n] - X_mean)) / (n - h)
    } else {
      cov_h <- sum((X[(1-h):n] - X_mean) * (X[1:(n+h)] - X_mean)) / (n + h)
    }
    var_X <- sum((X - X_mean)^2) / n
    acf_vals[h + max_lag + 1] <- cov_h / var_X
  }
  
  return(acf_vals)
}


# Theoretical ACF based on earlier work
theoretical_acf <- function(h) {
  if (h == 0) {
    return(1)  # ACF at lag 0 is 1
  } else if (abs(h) == 1) {
    return(2/3)  # ACF at lag ±1 is 2/3
  } else {
    return(0)  # ACF is 0 for other lags
  }
}

# Compute the sample ACF "by hand" for Xt
max_lag <- 10  # Set the maximum lag to 10 for the plot
sample_acf <- compute_acf_by_hand(X, max_lag)  # Now computes ACF for -10 to 10

# Theoretical ACF values for lags from -10 to 10
lags <- -max_lag:max_lag  # Set the x-axis from -10 to 10
true_acf <- sapply(abs(lags), theoretical_acf)  # Reflect true ACF to negative lags

# Adjust figure margins using the mar parameter
par(mar = c(4, 4, 2, 2))

# Plot the true ACF
par(mfrow = c(2, 1))  # Two plots stacked vertically

plot(lags, true_acf, type = "h", lwd = 2, col = "blue", ylim = c(0, 1), xlim = c(-10, 10),
     main = "True ACF of X_t", xlab = "Lag", ylab = "ACF", cex.lab = 1.5, cex.axis = 1.2)
points(lags, true_acf, col = "blue", pch = 16)

# Plot the sample ACF
plot(lags, sample_acf, type = "h", lwd = 2, col = "red", ylim = c(0, 1), xlim = c(-10, 10),
     main = "Sample ACF of X_t", xlab = "Lag", ylab = "ACF", cex.lab = 1.5, cex.axis = 1.2)
points(lags, sample_acf, col = "red", pch = 16)

# Restore default margin settings for the combined plot
par(mfrow = c(1, 1))  # Single plot again
par(mar = c(4, 4, 2, 2))  # Adjust margins again for the combined plot

# Plot both ACFs on the same plot for comparison
plot(lags, true_acf, type = "h", lwd = 2, col = "blue", ylim = c(-0.2, 1), xlim = c(-10, 10),
     main = "True ACF vs Sample ACF", xlab = "Lag", ylab = "ACF", cex.lab = 1.5, cex.axis = 1.2)
points(lags, true_acf, col = "blue", pch = 16)

# Add sample ACF on the same plot
lines(lags, sample_acf, type = "h", lwd = 2, col = "red")
points(lags, sample_acf, col = "red", pch = 16)

# Add a legend to differentiate the true ACF and the sample ACF
legend("topright", legend = c("True ACF", "Sample ACF"), col = c("blue", "red"), lwd = 2, pch = 16)

```

## Problem 2

### Part a

```{r,echo=FALSE,message=FALSE, warning=FALSE}
# Load necessary libraries
library(forecast)

# Load the Johnson & Johnson data (assuming you have it loaded as 'jj')
# If using built-in R dataset
data("JohnsonJohnson")
jj <- JohnsonJohnson

# Step 1: Log-transform the data
log_jj <- log(jj)

# Step 2: Create a time variable for the trend
time <- time(jj)  # Extract the time index from the data

# Step 3: Create quarter dummy variables for seasonality
Q1 <- ifelse(cycle(jj) == 1, 1, 0)  # Q1 dummy variable
Q2 <- ifelse(cycle(jj) == 2, 1, 0)  # Q2 dummy variable
Q3 <- ifelse(cycle(jj) == 3, 1, 0)  # Q3 dummy variable
Q4 <- ifelse(cycle(jj) == 4, 1, 0)  # Q4 dummy variable

# Step 4: Fit the regression model with trend and seasonal components
fit <- lm(log_jj ~ 0 + time + Q1 + Q2 + Q3 + Q4)

# Step 5: Display the summary of the fitted model
summary(fit)



```

### Part b

The average annual increase in the logged earnings per share if the model is correct is $0.1672 * 4 = 0.6688$ as shown in the summary of the regression model in part a multiplied by 4 to reflect the average yearly increase instead of the average quarterly increase.

### Part c

The logged earnings per share decreases from the third to the fourth quarter if the model is correct because the estimate for the Q4 indicator variable is lower than that of the Q3 indicator variable.

### Part d

If you attempt to include an intercept term in the model from part a, you can no longer use all of Q1, Q2, Q3, and Q4 as indicator variables. One of these will be redundant when fitting the model because one of them must be the baseline quarter from which the other quarter variables reflect the average relative difference from the baseline quarter in logged earnings per share.

### Part e

```{r,echo=FALSE}
# Step 6: Check residuals to see if they resemble white noise
residuals <- resid(fit)
acf(residuals)  # Plot ACF of residuals to check for white noise

# Step 7: Plot the original data and the fitted trend + seasonal components

# Ensure that 'time' and 'fitted_values' are time series objects
# 'log_jj' is already a time series, so convert 'fitted_values' to match its structure
fitted_values <- ts(fitted(fit), start = start(log_jj), frequency = frequency(log_jj))

# Plot the original log-transformed data (log_jj) with the fitted values
plot(log_jj, type = "l", col = "blue", main = "Log Johnson & Johnson Data with Fitted Values",
     ylab = "Log Transformed Data", xlab = "Time")

# Add the fitted values as a red line to the plot
lines(fitted_values, col = "red", lwd = 2)

# Add a legend to the plot to differentiate the two lines
legend("topleft", legend = c("Data", "Fitted Values"), col = c("blue", "red"), lwd = 2)
```

## Problem 3

### Part a

Yes, this is stationary because the mean is constant and the autocovariance only depends on the lag.

$\mu = 0$

$\gamma(h) = 2*\sigma^2$ when h = 0, $\gamma(h) = -\sigma^2$ when h = 3, and $\gamma(h) = 0$ otherwise.

### Part b

This is not stationary because the autocovariance depends on the time rather than the lag.

### Part c

This is not stationary because the mean is not constant.

### Part d

Yes, this is stationary because the mean is constant and the autocovariance only depends on the lag.

$\mu = E[W_t^2] = Var(W_t) + E[W_t]^2 = Var(W_t) + 0 = \sigma^2$

$\gamma(h) = E[W_t^4] - E[W_t^2]^2 = E[W_t^4] - \sigma^4$ when h = 0 and 0 otherwise.

### Part e

Yes, this is stationary because the mean is constant and the autocovariance only depends on the lag.

$\mu = E[W_tW_{t-2}] = E[W_t]E[W_{t-2}] = 0$

$\gamma(h) = E[W_t^2W_{t-2}^2] - E[W_tW_{t-2}]^2 = E[W_t^2W_{t-2}^2] - 0 = E[W_t^2W_{t-2}^2]$ when h = 0 and $E[W_tW_{t-2}W_{t+2}W_t] = E[W_t]^2E[W_{t-2}W_{t+2}] = \sigma^2 * 0 = 0$ when h = 2 and 0 otherwise.

## Problem 4 

X is white noise because:

$\mu = E[X] = 1/2 * 0 + 1/2 * E[(1-W_{t-1})Z_t] = 1/2 * 0 + 1/4 * 0 + 1/4 * Z_t = 0$

$Var(X_t) = E[X_t^2] - E[X_t]^2 = E[X_t^2] = 1/4* E[Z_t^2]  = 1/4$

$\gamma(h) = 0$ for all $h \neq 0$ because $\gamma(1) = E[W_tW_{t-1}(1-W_t)(1-W_{t-1})Z_tZ_{t+1}] = E[W_tW_{t-1}(1-W_t)(1-W_{t-1})]E[Z_t]E[Z_{t+1}] = 0$

X is not iid because it depends in values from the past ($W_{t-1}$})

## Problem 5

```{r,echo=FALSE}
# Load necessary libraries
library(forecast)  # for plotting ACF
library(ggplot2)

# Load the data from mystery.csv (assuming the file is in the working directory)
df <- read.csv("mystery.csv")

# Extract the data column (assuming it's the first column)
data <- ts(df[, 1])  # Convert the data into a time series object

# (a) Plot the sample ACF
acf(data, main = "Sample ACF of the Mystery Data", lag.max = 40)

# (b) Time series plot
plot.ts(data, main = "Time Series Plot of the Mystery Data", 
        xlab = "Time", ylab = "Value")

# (b) Lag-1 plot: plot X_t against X_{t-1}
lag_data <- data[-length(data)]  # X_{t-1}
lag1_data <- data[-1]  # X_t

# Use ggplot for a cleaner scatter plot
lag_df <- data.frame(Xt_minus_1 = lag_data, Xt = lag1_data)

ggplot(lag_df, aes(x = Xt_minus_1, y = Xt)) + 
  geom_point(alpha = 0.6) +
  labs(title = "Lag-1 Plot of the Mystery Data", 
       x = expression(X[t-1]), 
       y = expression(X[t])) +
  theme_minimal()


```

### Part a

Looking at the ACF and the series itself, it looks like it could be modeled as white noise.

## Part b

Seeing the lag 1 plot, it is clear that there is a relationship between $X_t$ and $X_{t-1}$, specifically a negative quadratic, so that rules out the possibility that it's white noise.



