---
title: "Homework 2"
author: "Alex Ojemann"
date: "2024-09-27"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem 1

### Part a

The following regression adjustments estimate the causal effect of quitting smoking on weight change, SBP, and DBP using sex, race, age, education, smoking intensity per day, number of years having smoked, exercise level, activity level, weight at the beginning of the study period, and whether the subject had high blood pressure at the beginning of the study period.

```{r,echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(broom)

library(causaldata)
data("nhefs", package = "causaldata")

# Label categorical variables properly
nhefs <- nhefs %>%
  mutate(
    qsmk = as.factor(qsmk), 
    race = as.factor(race), 
    education = as.factor(education),  
    hbp = as.factor(hbp), 
    exercise = as.factor(exercise),  
    active = as.factor(active)  
  )

# Fit linear regression models for each outcome, adjusting for covariates

# 1. Weight change (wt82_71)
model_wt82_71 <- lm(wt82_71 ~ qsmk + race + age + education + smokeintensity + smokeyrs +
                    exercise + active + wt71 + hbp, data = nhefs)

# 2. Systolic blood pressure (sbp)
model_sbp <- lm(sbp ~ qsmk + race + age + education + smokeintensity + smokeyrs +
                exercise + active + wt71 + hbp, data = nhefs)

# 3. Diastolic blood pressure (dbp)
model_dbp <- lm(dbp ~ qsmk + race + age + education + smokeintensity + smokeyrs +
                exercise + active + wt71 + hbp, data = nhefs)

format_results <- function(model, outcome_label) {
  # Extract the relevant statistics for qsmk (treatment effect)
  result <- tidy(model) %>% filter(term == "qsmk1")

  estimate <- round(result$estimate, 2)
  std_error <- round(result$std.error, 2)
  lower_CI <- round(result$estimate - 1.96 * result$std.error, 2)
  upper_CI <- round(result$estimate + 1.96 * result$std.error, 2)
  
  # Create a formatted, readable summary
  paste(
    outcome_label, ":\n",
    "- Estimated effect of quitting smoking: ", estimate, "\n",
    "- Standard Error: ", std_error, "\n",
    "- 95% Confidence Interval: [", lower_CI, ", ", upper_CI, "]\n",
    "\n"
  )
}

# Generate formatted results for each outcome
output_wt82_71 <- format_results(model_wt82_71, "Weight Change (1971-1982)")
output_sbp <- format_results(model_sbp, "Systolic Blood Pressure")
output_dbp <- format_results(model_dbp, "Diastolic Blood Pressure")


cat(output_wt82_71)
cat(output_sbp)
cat(output_dbp)

```
### Part b

The following regression adjustments estimate the causal effect of quitting smoking on weight change, SBP, and DBP using the same covariates as those in part a but with squared and logarithmic terms included for the continuous numeric covariates.

```{r,echo=FALSE}
data("nhefs", package = "causaldata")

# Properly label categorical variables
nhefs <- nhefs %>%
  mutate(
    qsmk = as.factor(qsmk), 
    race = as.factor(race), 
    education = as.factor(education),  
    hbp = as.factor(hbp), 
    exercise = as.factor(exercise),  
    active = as.factor(active),
    log_age = log(age),  # Add log transformation for age
    log_smokeintensity = log(smokeintensity + 1),  # Add log for smokeintensity (+1 to avoid log(0))
    log_smokeyrs = log(smokeyrs + 1),  # Add log for smokeyrs
    log_wt71 = log(wt71),  # Add log for wt71
    age_squared = age^2,  # Add quadratic term for age
    smokeintensity_squared = smokeintensity^2,  # Add quadratic term for smokeintensity
    smokeyrs_squared = smokeyrs^2,  # Add quadratic term for smokeyrs
    wt71_squared = wt71^2  # Add quadratic term for wt71
  )

# Fit linear regression models for each outcome, adjusting for covariates, including quadratic and log terms

# 1. Weight change (wt82_71)
model_wt82_71 <- lm(wt82_71 ~ qsmk + race + age + age_squared + log_age + 
                    education + smokeintensity + smokeintensity_squared + log_smokeintensity + 
                    smokeyrs + smokeyrs_squared + log_smokeyrs + exercise + active + 
                    wt71 + wt71_squared + log_wt71 + hbp, data = nhefs)

# 2. Systolic blood pressure (sbp)
model_sbp <- lm(sbp ~ qsmk + race + age + age_squared + log_age + 
                education + smokeintensity + smokeintensity_squared + log_smokeintensity + 
                smokeyrs + smokeyrs_squared + log_smokeyrs + exercise + active + 
                wt71 + wt71_squared + log_wt71 + hbp, data = nhefs)

# 3. Diastolic blood pressure (dbp)
model_dbp <- lm(dbp ~ qsmk + race + age + age_squared + log_age + 
                education + smokeintensity + smokeintensity_squared + log_smokeintensity + 
                smokeyrs + smokeyrs_squared + log_smokeyrs + exercise + active + 
                wt71 + wt71_squared + log_wt71 + hbp, data = nhefs)

# Function to format results
format_results <- function(model, outcome_label) {
  # Extract the relevant statistics for qsmk (treatment effect)
  result <- tidy(model) %>% filter(term == "qsmk1")

  estimate <- round(result$estimate, 2)
  std_error <- round(result$std.error, 2)
  lower_CI <- round(result$estimate - 1.96 * result$std.error, 2)
  upper_CI <- round(result$estimate + 1.96 * result$std.error, 2)
  
  # Create a formatted, readable summary
  paste(
    outcome_label, ":\n",
    "- Estimated effect of quitting smoking: ", estimate, "\n",
    "- Standard Error: ", std_error, "\n",
    "- 95% Confidence Interval: [", lower_CI, ", ", upper_CI, "]\n",
    "\n"
  )
}

# Generate formatted results for each outcome
output_wt82_71 <- format_results(model_wt82_71, "Weight Change (1971-1982)")
output_sbp <- format_results(model_sbp, "Systolic Blood Pressure")
output_dbp <- format_results(model_dbp, "Diastolic Blood Pressure")

# Print the formatted results
cat(output_wt82_71)
cat(output_sbp)
cat(output_dbp)


```

### Part c

The following regression adjustments estimate the causal effect of quitting smoking on weight change, SBP, and DBP using the same covariates as those in part a but with the product of the quit smoking covariate with each of the other covariates as additional features.

```{r,echo=FALSE}
# Load necessary packages
library(dplyr)
library(broom)

# Load the NHEFS dataset
library(causaldata)
data("nhefs", package = "causaldata")

# Label categorical variables properly
nhefs <- nhefs %>%
  mutate(
    qsmk = as.factor(qsmk), 
    race = as.factor(race), 
    education = as.factor(education),  
    hbp = as.factor(hbp), 
    exercise = as.factor(exercise),  
    active = as.factor(active)  
  )

# Fit linear regression models with interaction terms

# 1. Weight change (wt82_71) with interaction terms
model_wt82_71 <- lm(wt82_71 ~ qsmk * age + qsmk * race + qsmk * education + qsmk * exercise + 
                    smokeintensity + smokeyrs + active + wt71 + hbp, data = nhefs)

# 2. Systolic blood pressure (sbp) with interaction terms
model_sbp <- lm(sbp ~ qsmk * age + qsmk * race + qsmk * education + qsmk * exercise + 
                smokeintensity + smokeyrs + active + wt71 + hbp, data = nhefs)

# 3. Diastolic blood pressure (dbp) with interaction terms
model_dbp <- lm(dbp ~ qsmk * age + qsmk * race + qsmk * education + qsmk * exercise + 
                smokeintensity + smokeyrs + active + wt71 + hbp, data = nhefs)

# Function to format and display only the main qsmk term
format_qsmk_results <- function(model, outcome_label) {
  # Extract the relevant statistics for qsmk (main effect only)
  result <- tidy(model) %>% filter(term == "qsmk1")

  # Format the output for readability
  output <- paste0(outcome_label, ":\n",
                   "- Estimated effect of quitting smoking: ", round(result$estimate, 2), "\n",
                   "- Standard Error: ", round(result$std.error, 2), "\n",
                   "- 95% CI: [", round(result$estimate - 1.96 * result$std.error, 2), 
                   ", ", round(result$estimate + 1.96 * result$std.error, 2), "]\n\n")
  
  return(output)
}

# Generate formatted results for each outcome (only qsmk term)
output_wt82_71 <- format_qsmk_results(model_wt82_71, "Weight Change (1971-1982)")
output_sbp <- format_qsmk_results(model_sbp, "Systolic Blood Pressure")
output_dbp <- format_qsmk_results(model_dbp, "Diastolic Blood Pressure")

# Print the formatted results
cat(output_wt82_71)
cat(output_sbp)
cat(output_dbp)



```
### Part d

The following regression adjustments estimate the causal effect of quitting smoking on weight change, SBP, and DBP using the same covariates as those in parts a, b, and c but with the final models having features selected using stepwise feature selection with BIC as the information criterion.

Note: when initially using stepwise feature selection, the model dropped the quit smoking variable in the SBP model. It was added back in manually after stepwise feature selection because it is the variable for which we are trying to evaluate causal relationships.

```{r,echo=FALSE, message=FALSE, warning=FALSE}
# Load necessary packages
library(dplyr)
library(broom)
library(MASS)  # For stepwise feature selection

# Load the NHEFS dataset
library(causaldata)
data("nhefs", package = "causaldata")

# Label categorical variables properly and create squared and log terms
nhefs <- nhefs %>%
  mutate(
    qsmk = as.factor(qsmk), 
    race = as.factor(race), 
    education = as.factor(education),  
    hbp = as.factor(hbp), 
    exercise = as.factor(exercise),  
    active = as.factor(active),
    # Create squared and log terms
    log_age = log(age),  # Log of age
    age_squared = age^2,  # Age squared
    log_smokeintensity = log(smokeintensity + 1),  # Log of smokeintensity (+1 to avoid log(0))
    smokeintensity_squared = smokeintensity^2,  # Smokeintensity squared
    log_smokeyrs = log(smokeyrs + 1),  # Log of smoking years
    smokeyrs_squared = smokeyrs^2,  # Smoking years squared
    log_wt71 = log(wt71),  # Log of weight in 1971
    wt71_squared = wt71^2  # Weight in 1971 squared
  )

# Build initial models with all predictors, their squared and log terms, and interaction terms

# 1. Weight change (wt82_71)
model_wt82_71 <- lm(wt82_71 ~ qsmk * (age + race + education + exercise) +
                    log_age + age_squared + log_smokeintensity + smokeintensity_squared + 
                    log_smokeyrs + smokeyrs_squared + log_wt71 + wt71_squared + 
                    active + wt71 + hbp, data = nhefs)

# 2. Systolic blood pressure (sbp)
model_sbp <- lm(sbp ~ qsmk * (age + race + education + exercise) +
                log_age + age_squared + log_smokeintensity + smokeintensity_squared + 
                log_smokeyrs + smokeyrs_squared + log_wt71 + wt71_squared + 
                active + wt71 + hbp, data = nhefs)

# 3. Diastolic blood pressure (dbp)
model_dbp <- lm(dbp ~ qsmk * (age + race + education + exercise) +
                log_age + age_squared + log_smokeintensity + smokeintensity_squared + 
                log_smokeyrs + smokeyrs_squared + log_wt71 + wt71_squared + 
                active + wt71 + hbp, data = nhefs)

# Stepwise feature selection for each model

n <- nrow(nhefs)
# 1. Stepwise feature selection for weight change using BIC
final_model_wt82_71 <- step(model_wt82_71, direction = "both", trace = FALSE, k = log(n))

# 2. Stepwise feature selection for systolic blood pressure using BIC
final_model_sbp <- step(model_sbp, direction = "both", trace = FALSE, k = log(n))

# Amend the final model by adding back the qsmk variable
final_model_sbp <- update(final_model_sbp, . ~ . + qsmk)

# 3. Stepwise feature selection for diastolic blood pressure using BIC
final_model_dbp <- step(model_dbp, direction = "both", trace = FALSE, k = log(n))

# Function to format and display only the qsmk term from the final models
format_qsmk_final_model <- function(final_model, outcome_label) {
  # Extract the relevant statistics for qsmk (main effect only)
  result <- tidy(final_model) %>% filter(term == "qsmk1")
  
  # Format the output for readability
  output <- paste0(outcome_label, " (Selected Model):\n",
                   "- Estimated effect of quitting smoking: ", round(result$estimate, 2), "\n",
                   "- Standard Error: ", round(result$std.error, 2), "\n",
                   "- 95% CI: [", round(result$estimate - 1.96 * result$std.error, 2), 
                   ", ", round(result$estimate + 1.96 * result$std.error, 2), "]\n\n")
  
  return(output)
}

# Generate formatted results for each final model (only qsmk term)
output_wt82_71 <- format_qsmk_final_model(final_model_wt82_71, "Weight Change (1971-1982)")
output_sbp <- format_qsmk_final_model(final_model_sbp, "Systolic Blood Pressure")
output_dbp <- format_qsmk_final_model(final_model_dbp, "Diastolic Blood Pressure")

# Print the formatted results
cat(output_wt82_71)
cat(output_sbp)
cat(output_dbp)

```

Below is the summary for the model used to predict weight change:

```{r}
summary(final_model_wt82_71)
```

Below is the summary for the model used to predict SBP:

```{r}
summary(final_model_sbp)
```

Below is the summary for the model used to predict DBP:

```{r}
summary(final_model_dbp)
```


