---
title: "PubH7485 Homework2 Key"
date: "2024-09-27"
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(7485)
library(dplyr)
library(labelled)
library(causaldata)
library(knitr)
library(formattable)
library(MASS)
```

```{r, include=F}
data('nhefs')

# treatment
nhefs$qsmk 
# outcomes of interest
nhefs$wt82_71
nhefs$sbp
nhefs$dbp

adj_vars <- c('sex', 'race', 'age', 'education', 'smokeintensity', 'smokeyrs', 'exercise', 'active', 'wt71', 'hbp')

# imputation
cont_vars <- c('age', 'smokeintensity', 'smokeyrs', 'wt71', 'wt82_71', 'sbp', 'dbp')
cat_vars <- c('sex', 'race', 'education', 'exercise', 'active', 'hbp')

for (var in cont_vars) {
  nhefs[[var]][is.na(nhefs[[var]])] <- median(nhefs[[var]], na.rm = TRUE)
}

mode_function <- function(x) {
  x <- as.character(x)
  tbl <- table(x, useNA = "ifany")
  as.character(names(tbl)[which.max(tbl)])
}

for (var in cat_vars) {
  nhefs[[var]][is.na(nhefs[[var]])] <- mode_function(nhefs[[var]])
  nhefs[[var]] <- as.factor(nhefs[[var]]) 
}
```


```{r, include=F}
# function for calculating the ATE
calculate_ATE <- function(model_formula, data, num_bootstraps = 100) {
 
  wc_model <- lm(model_formula, data = data)
  
  nhefs_trt <- nhefs_con <- data
  nhefs_trt$qsmk = 1
  nhefs_con$qsmk = 0
  
  pred1_wc <- predict(wc_model, newdata = nhefs_trt, type = "response")
  pred0_wc <- predict(wc_model, newdata = nhefs_con, type = "response")
  ATE_wc <- mean(pred1_wc - pred0_wc)
  
  # Bootstrap replication for SE
  ate_wc_boot <- c()
  for (i in 1:num_bootstraps) {
    nhefs_boot <- data[sample(nrow(data), size = nrow(data), replace = TRUE), ]
    wc_model_boot <- lm(model_formula, data = nhefs_boot)
    
    nhefs_trt_boot <- nhefs_con_boot <- nhefs_boot
    nhefs_trt_boot$qsmk = 1
    nhefs_con_boot$qsmk = 0
    
    pred1_wc_boot <- predict(wc_model_boot, newdata = nhefs_trt_boot, type = "response")
    pred0_wc_boot <- predict(wc_model_boot, newdata = nhefs_con_boot, type = "response")
    ate_wc_boot <- c(ate_wc_boot, mean(pred1_wc_boot - pred0_wc_boot))
  }
  
  # Calculate standard error and confidence interval
  SE <- sd(ate_wc_boot)
  CI <- ATE_wc + c(-1, 1) * qnorm(0.975) * SE
  CI_str <- paste0(round(CI[1], 3), ", ", round(CI[2], 3))
  
  return(list(ATE = round(ATE_wc, 3), SE = round(SE, 3), CI = CI_str))
}

## for weight change
# linear regression
model_wc1 <- wt82_71 ~ qsmk + sex + race + age + education + smokeintensity + smokeyrs + 
                 exercise + active + wt71 + hbp
result_wc1 <- calculate_ATE(model_wc1, nhefs, num_bootstraps = 100)

print(result_wc1)

# add non-linear term
model_wc2 <- wt82_71 ~ qsmk + sex + race + poly(age, 2) + education + poly(smokeintensity, 2) + 
  poly(smokeyrs, 2) + exercise + active + poly(wt71, 2) + hbp
result_wc2 <- calculate_ATE(model_wc2, nhefs, num_bootstraps = 100)

print(result_wc2)

# add interaction term
model_wc3 <- wt82_71 ~ qsmk * (sex + race + age + education + smokeintensity + smokeyrs + 
                 exercise + active + wt71 + hbp)
result_wc3 <- calculate_ATE(model_wc3, nhefs, num_bootstraps = 100)

print(result_wc3)

## for sbp
# linear regression
model_sbp1 <- sbp ~ qsmk + sex + race + age + education + smokeintensity + smokeyrs + 
                 exercise + active + wt71 + hbp
result_sbp1 <- calculate_ATE(model_sbp1, nhefs, num_bootstraps = 100)
print(result_sbp1)

# add non-linear term
model_sbp2 <- sbp ~ qsmk + sex + race + poly(age, 2) + education + poly(smokeintensity, 2) + 
  poly(smokeyrs, 2) + exercise + active + poly(wt71, 2) + hbp
result_sbp2 <- calculate_ATE(model_sbp2, nhefs, num_bootstraps = 100)
print(result_sbp2)

# add interaction term
model_sbp3 <- sbp ~ qsmk * (sex + race + age + education + smokeintensity + smokeyrs + 
                 exercise + active + wt71 + hbp)
result_sbp3 <- calculate_ATE(model_sbp3, nhefs, num_bootstraps = 100)
print(result_sbp3)

## for dbp
# linear regression
model_dbp1 <- dbp ~ qsmk + sex + race + age + education + smokeintensity + smokeyrs + 
                 exercise + active + wt71 + hbp
result_dbp1 <- calculate_ATE(model_dbp1, nhefs, num_bootstraps = 100)
print(result_dbp1)

# add non-linear term
model_dbp2 <- dbp ~ qsmk + sex + race + poly(age, 2) + education + poly(smokeintensity, 2) + 
  poly(smokeyrs, 2) + exercise + active + poly(wt71, 2) + hbp
result_dbp2 <- calculate_ATE(model_dbp2, nhefs, num_bootstraps = 100)
print(result_dbp2)

# add interaction term
model_dbp3 <- dbp ~ qsmk * (sex + race + age + education + smokeintensity + smokeyrs + 
                 exercise + active + wt71 + hbp)
result_dbp3 <- calculate_ATE(model_dbp3, nhefs, num_bootstraps = 100)
print(result_dbp3)

```

## Question 1 (20 points for each outcome, 60 points total)

The calculated results are summarized in the following tables: 

```{r,echo=F}
rows <- c("ATE", "Standard Error", "95% CI")
columns <- c("Linear Regression", "Non-linear Term", "Interaction Term")

table_wc <- rbind(result_wc1, result_wc2, result_wc3)
colnames(table_wc) <- rows
rownames(table_wc) <- columns
kable(table_wc, caption = "Outcome Table for Weight Change")


table_sbp <- rbind(result_sbp1, result_sbp2, result_sbp3)
colnames(table_sbp) <- rows
rownames(table_sbp) <- columns
kable(table_sbp, caption = "Outcome Table for SBP")

table_dbp <- rbind(result_dbp1, result_dbp2, result_dbp3)
colnames(table_dbp) <- rows
rownames(table_dbp) <- columns
kable(table_dbp, caption = "Outcome Table for DBP")
```

Note that exact numbers may vary for students depending on the model fitting.


## Bonus point (5 points for each outcome, 15 points total)

```{r,include=F}
## for weight change
# specify full model
full_wc <- lm (wt82_71 ~ qsmk * (sex + race + poly(age,2) + education + poly(smokeintensity,2) + poly(smokeyrs,2) +
                              exercise + active + poly(wt71,2) + hbp), data=nhefs) 

# do stepwise selection
stepwise_wc <- stepAIC(full_wc, direction = "both")
summary(stepwise_wc)

select_wc <- wt82_71 ~ qsmk + sex + poly(age,2) + education + active + poly(wt71,2) + qsmk:education +
  qsmk:poly(age,2) + qsmk:poly(wt71,2)
result_wc4 <- calculate_ATE(select_wc, nhefs, num_bootstraps = 100)
print(result_wc4)

## for sbp
# specify full model
full_sbp <- lm (sbp ~ qsmk * (sex + race + poly(age,2) + education + poly(smokeintensity,2) + poly(smokeyrs,2) +
                              exercise + active + poly(wt71,2) + hbp), data=nhefs) 

# do stepwise selection
stepwise_sbp <- stepAIC(full_sbp, direction = "both")
summary(stepwise_sbp)

select_sbp <- sbp ~ qsmk + sex + race + poly(age,2) + education + poly(smokeintensity,2) + poly(smokeyrs,2)  +
  exercise + poly(wt71,2) + qsmk:race + qsmk:exercise + qsmk:poly(smokeintensity,2) + qsmk:poly(smokeyrs,2)
result_sbp4 <- calculate_ATE(select_sbp, nhefs, num_bootstraps = 100)
print(result_sbp4)


## for dbp
# specify full model
full_dbp <- lm (dbp ~ qsmk * (sex + race + poly(age,2) + education + poly(smokeintensity,2) + poly(smokeyrs,2) +
                              exercise + active + poly(wt71,2) + hbp),data=nhefs) 

# do stepwise selection
stepwise_dbp <- stepAIC(full_dbp, direction = "both")
summary(stepwise_dbp)

select_dbp <- dbp ~ qsmk + sex + race + poly(age,2) + poly(smokeyrs,2) + exercise + 
  active + poly(wt71,2) + qsmk:active
result_dbp4 <- calculate_ATE(select_dbp, nhefs, num_bootstraps = 100)
print(result_dbp4)
```


The calculated results are summarized in the following tables: 

```{r,echo=F}
rows_bonus <- c("ATE", "Standard Error", "95% CI")
columns_bonus <- c("Weight Change", "SBP", "DBP")

table_bonus <- rbind(result_wc4, result_sbp4, result_dbp4)
colnames(table_bonus) <- rows_bonus
rownames(table_bonus) <- columns_bonus

kable(table_bonus, caption = "Outcome Table by Stepwise Selection")
```

The key is using stepwise selection but you can perform other variable selection methods like lasso,  etc. 

# Appendix - R Code

1. Impute Missing Data

```{r,eval=F}
set.seed(7485)
data('nhefs')

# treatment
nhefs$qsmk 
# outcomes of interest
nhefs$wt82_71
nhefs$sbp
nhefs$dbp

adj_vars <- c('sex', 'race', 'age', 'education', 'smokeintensity', 'smokeyrs', 'exercise', 'active', 'wt71', 'hbp')

# imputation
cont_vars <- c('age', 'smokeintensity', 'smokeyrs', 'wt71', 'wt82_71', 'sbp', 'dbp')
cat_vars <- c('sex', 'race', 'education', 'exercise', 'active', 'hbp')

for (var in cont_vars) {
  nhefs[[var]][is.na(nhefs[[var]])] <- median(nhefs[[var]], na.rm = TRUE)
}

mode_function <- function(x) {
  x <- as.character(x)
  tbl <- table(x, useNA = "ifany")
  as.character(names(tbl)[which.max(tbl)])
}

for (var in cat_vars) {
  nhefs[[var]][is.na(nhefs[[var]])] <- mode_function(nhefs[[var]])
  nhefs[[var]] <- as.factor(nhefs[[var]]) 
}
```

2. Calculate ATE, standard error and 95% CI for each outcome

```{r, eval=F}
# function for calculating the ATE
calculate_ATE <- function(model_formula, data, num_bootstraps = 100) {
 
  wc_model <- lm(model_formula, data = data)
  
  nhefs_trt <- nhefs_con <- data
  nhefs_trt$qsmk = 1
  nhefs_con$qsmk = 0
  
  pred1_wc <- predict(wc_model, newdata = nhefs_trt, type = "response")
  pred0_wc <- predict(wc_model, newdata = nhefs_con, type = "response")
  ATE_wc <- mean(pred1_wc - pred0_wc)
  
  # Bootstrap replication for SE
  ate_wc_boot <- c()
  for (i in 1:num_bootstraps) {
    nhefs_boot <- data[sample(nrow(data), size = nrow(data), replace = TRUE), ]
    wc_model_boot <- lm(model_formula, data = nhefs_boot)
    
    nhefs_trt_boot <- nhefs_con_boot <- nhefs_boot
    nhefs_trt_boot$qsmk = 1
    nhefs_con_boot$qsmk = 0
    
    pred1_wc_boot <- predict(wc_model_boot, newdata = nhefs_trt_boot, type = "response")
    pred0_wc_boot <- predict(wc_model_boot, newdata = nhefs_con_boot, type = "response")
    ate_wc_boot <- c(ate_wc_boot, mean(pred1_wc_boot - pred0_wc_boot))
  }
  
  # Calculate standard error and confidence interval
  SE <- sd(ate_wc_boot)
  CI <- ATE_wc + c(-1, 1) * qnorm(0.975) * SE
  CI_str <- paste0(round(CI[1], 3), ", ", round(CI[2], 3))
  
  return(list(ATE = round(ATE_wc, 3), SE = round(SE, 3), CI = CI_str))
}

## for weight change
# linear regression
model_wc1 <- wt82_71 ~ qsmk + sex + race + age + education + smokeintensity + smokeyrs + 
                 exercise + active + wt71 + hbp
result_wc1 <- calculate_ATE(model_wc1, nhefs, num_bootstraps = 100)

print(result_wc1)

# add non-linear term
model_wc2 <- qsmk + sex + race + poly(age, 2) + education + poly(smokeintensity, 2) + 
  poly(smokeyrs, 2) + exercise + active + poly(wt71, 2) + hbp
result_wc2 <- calculate_ATE(model_wc2, nhefs, num_bootstraps = 100)

print(result_wc2)

# add interaction term
model_wc3 <- wt82_71 ~ qsmk * (sex + race + age + education + smokeintensity + smokeyrs + 
                 exercise + active + wt71 + hbp)
result_wc3 <- calculate_ATE(model_wc3, nhefs, num_bootstraps = 100)

print(result_wc3)

## for sbp
# linear regression
model_sbp1 <- sbp ~ qsmk + sex + race + age + education + smokeintensity + smokeyrs + 
                 exercise + active + wt71 + hbp
result_sbp1 <- calculate_ATE(model_sbp1, nhefs, num_bootstraps = 100)
print(result_sbp1)

# add non-linear term
model_sbp2 <- sbp ~ qsmk + sex + race + poly(age, 2) + education + poly(smokeintensity, 2) + 
  poly(smokeyrs, 2) + exercise + active + poly(wt71, 2) + hbp
result_sbp2 <- calculate_ATE(model_sbp2, nhefs, num_bootstraps = 100)
print(result_sbp2)

# add interaction term
model_sbp3 <- sbp ~ qsmk * (sex + race + age + education + smokeintensity + smokeyrs + 
                 exercise + active + wt71 + hbp)
result_sbp3 <- calculate_ATE(model_sbp3, nhefs, num_bootstraps = 100)
print(result_sbp3)

## for dbp
# linear regression
model_dbp1 <- dbp ~ qsmk + sex + race + age + education + smokeintensity + smokeyrs + 
                 exercise + active + wt71 + hbp
result_dbp1 <- calculate_ATE(model_dbp1, nhefs, num_bootstraps = 100)
print(result_dbp1)

# add non-linear term
model_dbp2 <- dbp ~ qsmk + sex + race + poly(age, 2) + education + poly(smokeintensity, 2) + 
  poly(smokeyrs, 2) + exercise + active + poly(wt71, 2) + hbp
result_dbp2 <- calculate_ATE(model_dbp2, nhefs, num_bootstraps = 100)
print(result_dbp2)

# add interaction term
model_dbp3 <- dbp ~ qsmk * (sex + race + age + education + smokeintensity + smokeyrs + 
                 exercise + active + wt71 + hbp)
result_dbp3 <- calculate_ATE(model_dbp3, nhefs, num_bootstraps = 100)
print(result_dbp3)

```

3. Create summary table for each outcomes

```{r, eval=F}
rows <- c("ATE", "Standard Error", "95% CI")
columns <- c("Linear Regression", "Non-linear Term", "Interaction Term")

table_wc <- rbind(result_wc1, result_wc2, result_wc3)
colnames(table_wc) <- rows
rownames(table_wc) <- columns
kable(table_wc, caption = "Outcome Table for Weight Change")


table_sbp <- rbind(result_sbp1, result_sbp2, result_sbp3)
colnames(table_sbp) <- rows
rownames(table_sbp) <- columns
kable(table_sbp, caption = "Outcome Table for SBP")

table_dbp <- rbind(result_dbp1, result_dbp2, result_dbp3)
colnames(table_dbp) <- rows
rownames(table_dbp) <- columns
kable(table_dbp, caption = "Outcome Table for DBP")
```

4. Perform stepwise selection for each outcome (or you can perform lasso or other variable selection method)

```{r,eval=F}
## for weight change
# specify full model
full_wc <- lm (wt82_71 ~ qsmk * (sex + race + poly(age,2) + education + poly(smokeintensity,2) + poly(smokeyrs,2) +
                              exercise + active + poly(wt71,2) + hbp), data=nhefs) 

# do stepwise selection
stepwise_wc <- stepAIC(full_wc, direction = "both")
summary(stepwise_wc)

select_wc <- wt82_71 ~ qsmk + sex + age + education + smokeyrs + active + wt71 + qsmk:education
result_wc4 <- calculate_ATE(select_wc, nhefs, num_bootstraps = 100)
print(result_wc4)

## for sbp
# specify full model
full_sbp <- lm (sbp ~ qsmk * (sex + race + poly(age,2) + education + poly(smokeintensity,2) + poly(smokeyrs,2) +
                              exercise + active + poly(wt71,2) + hbp), data=nhefs) 

# do stepwise selection
stepwise_sbp <- stepAIC(full_sbp, direction = "both")
summary(stepwise_sbp)

select_sbp <- sbp ~ qsmk + sex + race + age + education + wt71 + hbp + qsmk:sex + qsmk:race + qsmk:hbp
result_sbp4 <- calculate_ATE(select_sbp, nhefs, num_bootstraps = 100)
print(result_sbp4)


## for dbp
# specify full model
full_dbp <- lm (dbp ~ qsmk * (sex + race + poly(age,2) + education + poly(smokeintensity,2) + poly(smokeyrs,2) +
                              exercise + active + poly(wt71,2) + hbp),data=nhefs) 

# do stepwise selection
stepwise_dbp <- stepAIC(full_dbp, direction = "both")
summary(stepwise_dbp)

select_dbp <- dbp ~ qsmk + sex + race + smokeyrs + exercise + active + wt71 + hbp + qsmk : active
result_dbp4 <- calculate_ATE(select_dbp, nhefs, num_bootstraps = 100)
print(result_dbp4)

rows_bonus <- c("ATE", "Standard Error", "95% CI")
columns_bonus <- c("Weight Change", "SBP", "DBP")

table_bonus <- rbind(result_wc4, result_sbp4, result_dbp4)
colnames(table_bonus) <- rows_bonus
rownames(table_bonus) <- columns_bonus

kable(table_bonus, caption = "Outcome Table by Stepwise Selection")
```
