---
title: "Final"
author: "Alex Ojemann"
date: "2024-12-17"
output: 
  pdf_document:
    toc: true       # Enable Table of Contents
    toc_depth: 2    # Set depth of TOC (e.g., 2 for sections and subsections)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage

## Problem 1

### Summary

For the series $Y_t = X_t$ I chose a SARIMA(4,0,1)x(0,1,2)[12] model excluding the intercept term. The coefficients shown below are from a model fit on $X_t$ after having been non seasonally differenced and seasonally differenced with lag 12.

Coefficients: 

     Estimate     SE  t.value p.value
     
ar1    0.2667 0.1446   1.8447  0.0656

ar2    0.6089 0.0459  13.2693  0.0000

ar3   -0.1343 0.0811  -1.6558  0.0983

ar4   -0.2486 0.0442  -5.6252  0.0000

ma1   -0.1801 0.1487  -1.2109  0.2264

sma1  -0.5948 0.0435 -13.6705  0.0000

sma2  -0.2039 0.0469  -4.3443  0.0000

### Explanation

The initial plot and ACF and PACF of $X_t$ suggest that the series is stationary so we don't need to take a non-seasonal difference but there is a seasonality of 12 so we should take a seasonal difference at lag 12. After the seasonal difference, the significant non seasonal lags of the ACF cut off for the most part after lag 2 and those of the PACF cut off after lag 4, while the significant seasonal lags of the ACF cut off for the most part after lag 2 and those of the PACF cut off after lag 2, so we fit different models on the first seasonally differenced data with a maximum p of 4, a maximum q of 2, a maximum P of 2 and a maximum Q of 2. Only the seven models with an AICc multiplied by the length of the series within 5 of the minimum and a BIC multiplied by the length of the series within 7 of the minimum were considered. Of these, the SARMA(4,1)x(0,2) was chosen because it's very close to the best AICc and it has the highest LBP values. So the final model on the original series $X_t$ is SARIMA(4,0,1)x(0,1,2)[12]. The intercept was not included because it was not significant at any level. It's worth noting that there are some notably heavy tails in the Normal QQ plots of the residuals for all models.

### Output

```{r}
load("final-data.rsav")
xx <- dat1
plot(xx)
```
```{r}
library(astsa)
invisible(acf2(xx))
```
```{r}
Dxx <- diff(xx,lag=12)
plot(Dxx)
```

```{r}
invisible(acf2(Dxx))
```



```{r}
maxAR <- 4+1
maxMA <- 2+1
maxSAR <- 2+1
maxSMA <- 2+1
fits_all <- array(list(), dim = c(maxAR, maxMA, maxSAR, maxSMA))
AICmin <- BICmin <- Inf
for (ii in 1:maxAR) {
for (jj in 1:maxMA) {
for (kk in 1:maxSAR) {
for (ll in 1:maxSMA) {
#if(!(ii==3 && jj == 3 && kk == 3 && ll == 2)){ #Condition because these models caused an error
print(ii)
print(jj)
print(kk)
print(ll)
fits_all[[ii, jj, kk, ll]] <- astsa::sarima(Dxx, p = ii - 1, d = 0,
q = jj - 1, P = kk - 1, D = 0, Q = ll - 1, S = 12, no.constant = F,
details = F)
if (fits_all[[ii, jj, kk, ll]]$ICs["AICc"] < AICmin)
AICmin <- fits_all[[ii, jj, kk, ll]]$ICs["AICc"]
if (fits_all[[ii, jj, kk, ll]]$ICs["BIC"] < BICmin)
BICmin <- fits_all[[ii, jj, kk, ll]]$ICs["BIC"]
}
}
}
}
#}
```

```{r}
nn <- length(Dxx)
print("AICcs")
## [1] "AICcs"
for (ii in 1:maxAR) {
for (jj in 1:maxMA) {
for (kk in 1:maxSAR) {
for (ll in 1:maxSMA) {
#if(!(ii==4 && jj == 5 && kk == 2 && ll == 3)){ #Condition because (3,4)x(1,2) caused an error
AICdelta <- ((fits_all[[ii, jj, kk, ll]])$ICs["AICc"] - AICmin) * nn
if (AICdelta < 5) {
print(paste0("(", ii - 1, ", ", jj - 1, ") x (", kk - 1, ", ",
ll - 1, ")"))
print(AICdelta)
}
}
}
}
}
#}
```

```{r}
print("BICs")
## [1] "BICs"
for (ii in 1:maxAR) {
for (jj in 1:maxMA) {
for (kk in 1:maxSAR) {
for (ll in 1:maxSMA) {
BICdelta <- ((fits_all[[ii, jj, kk, ll]])$ICs["BIC"] - BICmin) * nn
if (BICdelta < 7) {
print(paste0("(", ii - 1, ", ", jj - 1, ") x (", kk - 1, ", ",
ll - 1, ")"))
print(BICdelta)
}
}
}
}
}
```

```{r}
sarima(Dxx, p = 1, d = 0, q = 2, P = 0, D = 0,
Q = 2, S = 12,no.constant=F,details=T)
```

```{r}
sarima(Dxx, p = 2, d = 0, q = 2, P = 0, D = 0,
Q = 2, S = 12,no.constant=F,details=T)
```

```{r}
sarima(Dxx, p = 2, d = 0, q = 2, P = 1, D = 0,
Q = 1, S = 12,no.constant=F,details=T)
```

```{r}
sarima(Dxx, p = 4, d = 0, q = 0, P = 0, D = 0,
Q = 2, S = 12,no.constant=F,details=T)
```
```{r}
sarima(Dxx, p = 4, d = 0, q = 0, P = 1, D = 0,
Q = 1, S = 12,no.constant=F,details=T)
```

```{r}
sarima(Dxx, p = 0, d = 0, q = 2, P = 0, D = 0,
Q = 2, S = 12,no.constant=F,details=T)
```

```{r}
sarima(Dxx, p = 0, d = 0, q = 2, P = 1, D = 0,
Q = 1, S = 12,no.constant=F,details=T)
```

```{r}
sarima(Dxx, p = 4, d = 0, q = 1, P = 0, D = 0,
Q = 2, S = 12,no.constant=F,details=T)
```

```{r}
sarima(Dxx, p = 4, d = 0, q = 0, P = 1, D = 0,
Q = 2, S = 12,no.constant=F,details=T)
```

Final model:
```{r}
sarima(Dxx, p = 4, d = 0, q = 1, P = 0, D = 0,
Q = 2, S = 12,no.constant=T,details=T)
```

## Problem 2

### Summary

a. 

Frequency 0.027: AR(1) Spectral Density = 8.58566 | CI = [3.51401, 13.45976] | In CI: YES
Frequency 0.049: AR(1) Spectral Density = 5.49633 | CI = [6.53474, 25.03013] | In CI: NO
Frequency 0.250: AR(1) Spectral Density = 0.52240 | CI = [0.92678, 3.54986] | In CI: NO

b. 

Frequency 0.027: AR(1) Spectral Density = 8.46003 | CI = [3.75334, 14.37649] | In CI: YES

### Explanation

For dat2a, of the smooth periodograms tested, the modified Daniell kernel (4,4) was chosen because it was not as jagged as some of the others but still maintained the steepness of the steep portions of the original periodogram as well as any of the others.

The same kernel was selected for dat2b because the plot of different smoothed periodograms is extremely similar.

The added value in part b results in a slightly different confidence interval than that of part a for $\omega = 0.027$ but the true value of the baseline AR(1) is still in this interval. This true AR(1) value for dat2b is slightly different from that of dat2a because the closest index to 0.027 from the smooth periodogram is slightly different for dat2b.


### Output

Baseline AR(1):

```{r}
library(TSA) ## ARMAspec

phi <- .7
trueARMAspec <- TSA::ARMAspec(model = list(ar=phi),
                              lty=2) ## function for plotting true
 
```

Dat2a:

```{r}
xx <- dat2a

# Reset plot parameters and set smaller margins
par(mfrow = c(1, 1))              # Single plot layout
par(mai = c(1, 1, 1, 1))          # Margins: B, L, T, R (reasonable size)

# Start the plot
plot.new()

# Plot the spectral density with multiple spans
spg <- spec.pgram(x = xx, taper = 0, log = "no", detrend = FALSE,
                  add = FALSE, lwd = 0.5, col = "gray")  # First plot (not add=TRUE)

# Overlay additional spectral densities with smoothing
spec.pgram(x = xx, spans = 15, taper = 0, log = "no", detrend = FALSE,
           add = TRUE, col = "red")

spec.pgram(x = xx, spans = 33, taper = 0, log = "no", detrend = FALSE,
           add = TRUE, col = "red")

spec.pgram(x = xx, spans = 71, taper = 0, log = "no", detrend = FALSE,
           add = TRUE, col = "blue")

spec.pgram(x = xx, spans = 171, taper = 0, log = "no", detrend = FALSE,
           add = TRUE, col = "orange")

(myk <- kernel("modified.daniell", m=c(30,30)))
## plot(myk)
spec.pgram(x=xx, kernel=myk,
           taper=0, log="no", detrend=FALSE,
           add=TRUE, col="green")

(myk <- kernel("modified.daniell", m=c(4,4)))
## plot(myk)
spec.pgram(x=xx, kernel=myk,
           taper=0, log="no", detrend=FALSE,
           add=TRUE, col="purple")

(myk <- kernel("modified.daniell", m=c(10, 10)))
## plot(myk)
spec.pgram(x=xx, kernel=myk,
           taper=0, log="no", detrend=FALSE,
           add=TRUE, col="darkgreen") 


# Add a title and axis labels
title("Spectral Density Estimates with Varying Smoothing Spans")
mtext("Frequency", side = 1, line = 2.5)
mtext("Spectral Density", side = 2, line = 2.5)

```
Modified Daniell Kernel (4,4)

```{r}
myk <- kernel("modified.daniell", m=c(4, 4))
spec.pgram(x=xx, kernel=myk,
           taper=0, log="yes", detrend=FALSE,
           lwd=.5, col="gray")

```

Confidence intervals and tests:

```{r}
# Frequencies of interest
omega <- c(0.027, 0.049, 0.25)

# Spectral density estimate using modified Daniell kernel (m = 4, 4 --> LL = 9)
myk <- kernel("modified.daniell", m = c(4, 4))
spec_est <- spec.pgram(x = xx, kernel = myk, taper = 0, log = "no", detrend = FALSE, plot = FALSE)

# Extract the spectral density estimates at the closest frequencies
closest_indices <- sapply(omega, function(w) which.min(abs(spec_est$freq - w)))
f_hat <- spec_est$spec[closest_indices]  

LL <- 9  
df <- 2 * LL  

# Chi-squared quantiles for 95% CI
q_lower <- qchisq(0.025, df)
q_upper <- qchisq(0.975, df)

# Construct confidence intervals
ci_lower <- f_hat * (df / q_upper)
ci_upper <- f_hat * (df / q_lower)

# Extract true spectral density at the chosen frequencies
true_spectrum <- trueARMAspec$spec[closest_indices]

# Combine results into a data frame
results <- data.frame(
  Frequency = omega,
  Empirical_Spec = f_hat,
  CI_Lower = ci_lower,
  CI_Upper = ci_upper,
  True_Spec = true_spectrum,
  In_CI = true_spectrum >= ci_lower & true_spectrum <= ci_upper
)

# Print results
print(results)

# Display whether the true spectrum falls within the confidence intervals
cat("\nResults:\n")
for (i in 1:nrow(results)) {
  cat(sprintf("Frequency %.3f: AR(1) Spectral Density = %.5f | CI = [%.5f, %.5f] | In CI: %s\n",
              results$Frequency[i], results$True_Spec[i],
              results$CI_Lower[i], results$CI_Upper[i],
              ifelse(results$In_CI[i], "YES", "NO")))
}


```

dat2b:

```{r}
xx <- dat2b

# Reset plot parameters and set smaller margins
par(mfrow = c(1, 1))              # Single plot layout
par(mai = c(1, 1, 1, 1))          # Margins: B, L, T, R (reasonable size)

# Start the plot
plot.new()

# Plot the spectral density with multiple spans
spg <- spec.pgram(x = xx, taper = 0, log = "no", detrend = FALSE,
                  add = FALSE, lwd = 0.5, col = "gray")  # First plot (not add=TRUE)

# Overlay additional spectral densities with smoothing
spec.pgram(x = xx, spans = 15, taper = 0, log = "no", detrend = FALSE,
           add = TRUE, col = "red")

spec.pgram(x = xx, spans = 33, taper = 0, log = "no", detrend = FALSE,
           add = TRUE, col = "red")

spec.pgram(x = xx, spans = 71, taper = 0, log = "no", detrend = FALSE,
           add = TRUE, col = "blue")

spec.pgram(x = xx, spans = 171, taper = 0, log = "no", detrend = FALSE,
           add = TRUE, col = "orange")

(myk <- kernel("modified.daniell", m=c(30,30)))
## plot(myk)
spec.pgram(x=xx, kernel=myk,
           taper=0, log="no", detrend=FALSE,
           add=TRUE, col="green")

(myk <- kernel("modified.daniell", m=c(4,4)))
## plot(myk)
spec.pgram(x=xx, kernel=myk,
           taper=0, log="no", detrend=FALSE,
           add=TRUE, col="purple")

(myk <- kernel("modified.daniell", m=c(10, 10)))
## plot(myk)
spec.pgram(x=xx, kernel=myk,
           taper=0, log="no", detrend=FALSE,
           add=TRUE, col="darkgreen") 


# Add a title and axis labels
title("Spectral Density Estimates with Varying Smoothing Spans")
mtext("Frequency", side = 1, line = 2.5)
mtext("Spectral Density", side = 2, line = 2.5)

```

Modified Daniell Kernel (4,4)

```{r}
myk <- kernel("modified.daniell", m=c(4, 4))
spec.pgram(x=xx, kernel=myk,
           taper=0, log="yes", detrend=FALSE,
           lwd=.5, col="gray")
```

Confidence intervals and tests:

```{r}
# Frequencies of interest
omega <- c(0.027)

# Spectral density estimate using modified Daniell kernel (m = 4, 4 --> LL = 9)
myk <- kernel("modified.daniell", m = c(4, 4))
spec_est <- spec.pgram(x = xx, kernel = myk, taper = 0, log = "no", detrend = FALSE, plot = FALSE)

# Extract the spectral density estimates at the closest frequencies
closest_indices <- sapply(omega, function(w) which.min(abs(spec_est$freq - w)))
f_hat <- spec_est$spec[closest_indices]  

LL <- 9  
df <- 2 * LL  

# Chi-squared quantiles for 95% CI
q_lower <- qchisq(0.025, df)
q_upper <- qchisq(0.975, df)

# Construct confidence intervals
ci_lower <- f_hat * (df / q_upper)
ci_upper <- f_hat * (df / q_lower)

# Extract true spectral density at the chosen frequencies
true_spectrum <- trueARMAspec$spec[closest_indices]

# Combine results into a data frame
results <- data.frame(
  Frequency = omega,
  Empirical_Spec = f_hat,
  CI_Lower = ci_lower,
  CI_Upper = ci_upper,
  True_Spec = true_spectrum,
  In_CI = true_spectrum >= ci_lower & true_spectrum <= ci_upper
)

# Print results
print(results)

# Display whether the true spectrum falls within the confidence intervals
cat("\nResults:\n")
for (i in 1:nrow(results)) {
  cat(sprintf("Frequency %.3f: AR(1) Spectral Density = %.5f | CI = [%.5f, %.5f] | In CI: %s\n",
              results$Frequency[i], results$True_Spec[i],
              results$CI_Lower[i], results$CI_Upper[i],
              ifelse(results$In_CI[i], "YES", "NO")))
}


```



## Problem 3

### Summary

For the series $Y_t$ I chose an ARMA(1,1) model including the intercept with exogenous regressors $X_{t+8}$ and X_{t+11}.

Coefficients: 

          Estimate     SE t.value p.value
          
ar1         0.6878 0.0308 22.3412  0.0000

ma1         0.7379 0.0298 24.7469  0.0000

intercept  94.3697 2.5859 36.4939  0.0000

XL8        -0.2148 0.0990 -2.1701  0.0304

XL11        0.2368 0.0958  2.4711  0.0137


### Explanation

Both $X_t$ and $Y_t$ are stationary and do not have changing variance over time or seasonality. The ACF and PACF of $X_t$ show 1 significant lag in the ACF and 5 significant lags in the PACF, so models with a maximum p of 1 and a maximum q of 5 were considered. The ARMA(2,1) model had an AICc and BIC very close to the best and had by far the best LBP values so it was chosen for prewhitening. The intercept was included because it was significant at all levels.

After prewhiteneing both $X_t$ and $Y_t$ with this model, the ccf shows significant correlations at -8, -11, -14, -17, -20, and -23. When testing combinations of these lags, it was found that only -8 and -11 were significant in a regression model. Thus $X_{t+8}$ and $X_{t+11}$ were used in the design matrix with $Y_t$.

For the model of $Y_t$ including the exogenous regressors $X_{t+8}$ and $X_{t+11}$, the ACF and PACF show 3 significant lags in the ACF and 7 significant lags in the PACF, so models with a maximum p of 7 and a maximum q of 3 were considered. Of the ones with an AICc and BIC within 10 of the minimum, the ARMA(1,1) had the best BIC and strong diagnostic plots, so it was the chosen model. The intercept was included because it was significant at all levels.

### Output

```{r}
# Access the 'xx' column
xx <- dat3[, "xx"]
```

```{r}
plot(dat3)

# Compute and plot the ACF and PACF of 'xx'
invisible(acf2(xx, main = "ACF and PACF of xx"))
```

```{r}
maxAR<-5+1
maxMA<-1+1
fits_all<-vector("list",length=maxAR)
AICmin<-BICmin<-Inf
for (ii in 1:maxAR){
fits_all[[ii]]<-vector("list",length=maxMA)
for (jj in 1:maxMA){
fits_all[[ii]][[jj]]<-invisible(sarima(xx,p=ii-1,d=0,q=jj-1,no.constant=FALSE, details=FALSE))
if(fits_all[[ii]][[jj]]$ICs["AICc"]<AICmin) AICmin<-fits_all[[ii]][[jj]]$ICs["AICc"]
if(fits_all[[ii]][[jj]]$ICs["BIC"]<BICmin) BICmin<-fits_all[[ii]][[jj]]$ICs["BIC"]
}
}
nn <- length(xx)
#look at AICs and BICs
print("check AICcs")
## [1] "check AICcs"
for (ii in 1:maxAR){
for(jj in 1:maxMA){
print(paste0(ii-1,"",jj-1))
print(((fits_all[[ii]][[jj]])$ICs["AICc"] - AICmin) * nn)
}
}
print("check BICs")
## [1] "check BICs"
for (ii in 1:maxAR){
for(jj in 1:maxMA){
print(paste0(ii-1,"",jj-1))
print(((fits_all[[ii]][[jj]])$ICs["BIC"] - BICmin) * nn)
}
}
```

```{r}
sarima(xx,p=0,d=0,q=1,no.constant=FALSE, details=TRUE)
```
```{r}
sarima(xx,p=1,d=0,q=1,no.constant=FALSE, details=TRUE)
```
```{r}
sarima(xx,p=2,d=0,q=1,no.constant=FALSE, details=TRUE)
```

```{r}
sarima(xx,p=3,d=0,q=1,no.constant=FALSE, details=TRUE)
```

```{r}
library(astsa)  # For sarima
library(stats)  # For filtering

# Fit an ARMA(2,1) model to xx using sarima
sarima_fit <- sarima(xx, p = 2, d = 0, q = 1, details = FALSE)

# Pre-whiten xx: Extract residuals from the SARIMA fit
xx_prewhitened <- ts(sarima_fit$fit$residuals)

# Get AR(2) coefficients and MA(1) coefficient
ar_coeffs <- coef(sarima_fit$fit)[c("ar1", "ar2")]  # AR(2) coefficients
ma_coeff  <- coef(sarima_fit$fit)["ma1"]            # MA(1) coefficient

# Step 1: Pre-whiten yy with the AR(2) part
yy <- dat3[, "yy"]

# Apply AR(2) filter: Use (1 - ar1*B - ar2*B^2)
yy_ar_filtered <- stats::filter(yy, filter = c(1, -ar_coeffs), method = "convolution", sides = 1)

# Step 2: Remove NAs caused by AR filtering
yy_ar_filtered <- na.omit(yy_ar_filtered)
xx_prewhitened <- xx_prewhitened[-c(1, 2)]

# Step 3: Apply MA(1) filter to the AR-filtered series
# Use (1 + ma1*B) where B is the lag operator
yy_prewhitened <- stats::filter(yy_ar_filtered, filter = c(1, -ma_coeff), method = "recursive", sides = 1)

# Plot the cross-correlation function (CCF) of the pre-whitened series
ccf(xx_prewhitened, yy_prewhitened, main = "CCF of Pre-whitened xx and yy",
    ylab = "CCF", na.action = na.omit)

```

```{r}
XL8 <- lag(xx, -8)
XL11 <- lag(xx, -11)
design_matrix <- ts.intersect(yy, XL8,XL11)

fit1 <- lm(yy ~ XL8+XL11, data = design_matrix)
summary(fit1)
invisible(acf2(resid(fit1)))
```

```{r}
maxAR<-7+1
maxMA<-3+1
fits_all<-vector("list",length=maxAR)
AICmin<-BICmin<-Inf
for (ii in 1:maxAR){
fits_all[[ii]]<-vector("list",length=maxMA)
for (jj in 1:maxMA){
fits_all[[ii]][[jj]]<-invisible(sarima(design_matrix[, 1],p=ii-1,d=0,q=jj-1,xreg = design_matrix[, 2:3],no.constant=FALSE, details=FALSE))
if(fits_all[[ii]][[jj]]$ICs["AICc"]<AICmin) AICmin<-fits_all[[ii]][[jj]]$ICs["AICc"]
if(fits_all[[ii]][[jj]]$ICs["BIC"]<BICmin) BICmin<-fits_all[[ii]][[jj]]$ICs["BIC"]
}
}
nn <- length(xx)
#look at AICs and BICs
print("check AICcs")
## [1] "check AICcs"
for (ii in 1:maxAR){
for(jj in 1:maxMA){
print(paste0(ii-1,"",jj-1))
print(((fits_all[[ii]][[jj]])$ICs["AICc"] - AICmin) * nn)
}
}
print("check BICs")
## [1] "check BICs"
for (ii in 1:maxAR){
for(jj in 1:maxMA){
print(paste0(ii-1,"",jj-1))
print(((fits_all[[ii]][[jj]])$ICs["BIC"] - BICmin) * nn)
}
}
```

```{r}
sarima(design_matrix[, 1],p=1,d=0,q=1,xreg = design_matrix[, 2:3],no.constant=FALSE, details=TRUE)
```

```{r}
sarima(design_matrix[, 1],p=2,d=0,q=1,xreg = design_matrix[, 2:3],no.constant=FALSE, details=TRUE)
```

```{r}
sarima(design_matrix[, 1],p=1,d=0,q=2,xreg = design_matrix[, 2:3],no.constant=FALSE, details=TRUE)
```

```{r}
sarima(design_matrix[, 1],p=5,d=0,q=3,xreg = design_matrix[, 2:3],no.constant=FALSE, details=TRUE)
```

