---
title: "PubH7485 Homework4 Key"
date: "2024-10-18"
output: 
  pdf_document: 
    latex_engine: xelatex
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

set.seed(7485)

library(dplyr)
library(labelled)
library(causaldata)
library(knitr)
library(MASS)
library(reshape2)
library(ggplot2)
library(survey)
library(tableone)

library(formattable)
library(rms)
library(forestplot)
library(Matching)
library(MatchIt)
```

```{r, include=F}
data('nhefs')

# treatment
nhefs$qsmk 
# outcomes of interest
nhefs$wt82_71
nhefs$sbp
nhefs$dbp

all_covars <- setdiff(names(nhefs), c('dadth', 'death', 'hightax82', 'modth', 'price71_82', 
                                      'price82', 'seqn', 'yrdth', 'wt82', 'tax71_82',
                                      'tax82', 'smokeyrs', 'smkintensity82_71',
                                      'birthplace','id','censored','older'))

adj_vars <- c('sex', 'race', 'age', 'education', 'smokeintensity', 'smokeyrs', 'exercise', 'active', 'wt71', 'hbp')


# Convert categorical variables to factors and set levels for them in nhefs data frame
nhefs$active <- factor(nhefs$active, levels = c(0, 1, 2), labels = c("Very active", "Moderately active", "Inactive"))
nhefs$alcoholfreq <- factor(nhefs$alcoholfreq, levels = c(0, 1, 2, 3, 4, 5), labels = c("Almost every day", "2-3 times/week", "1-4 times/month", "< 12 times/year", "No alcohol last year", "Unknown"))
nhefs$alcoholpy <- factor(nhefs$alcoholpy, levels = c(0, 1, 2), labels = c("Never", "Ever", "Missing"))
nhefs$alcoholtype <- factor(nhefs$alcoholtype, levels = c(1, 2, 3, 4), labels = c("Beer", "Wine", "Liquor", "Other/Unknown"))
nhefs$allergies <- factor(nhefs$allergies, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$asthma <- factor(nhefs$asthma, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$birthcontrol <- factor(nhefs$birthcontrol, levels = c(0, 1, 2), labels = c("No", "Yes", "Missing"))
nhefs$boweltrouble <- factor(nhefs$boweltrouble, levels = c(0, 1, 2), labels = c("Never", "Ever", "Missing"))
nhefs$bronch <- factor(nhefs$bronch, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$chroniccough <- factor(nhefs$chroniccough, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$colitis <- factor(nhefs$colitis, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$diabetes <- factor(nhefs$diabetes, levels = c(0, 1, 2), labels = c("Never", "Ever", "Missing"))
nhefs$education <- factor(nhefs$education, levels = c(1, 2, 3, 4, 5), labels = c("8th grade or less", "HS dropout", "HS", "College dropout", "College or more"))
nhefs$exercise <- factor(nhefs$exercise, levels = c(0, 1, 2), labels = c("Little or no exercise", "Moderate exercise", "Much exercise"))
nhefs$hayfever <- factor(nhefs$hayfever, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$hbp <- factor(nhefs$hbp, levels = c(0, 1, 2), labels = c("Never", "Ever", "Missing"))
nhefs$hbpmed <- factor(nhefs$hbpmed, levels = c(0, 1, 2), labels = c("Never", "Ever", "Missing"))
nhefs$headache <- factor(nhefs$headache, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$hepatitis <- factor(nhefs$hepatitis, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$hf <- factor(nhefs$hf, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$marital <- factor(nhefs$marital, levels = c(1, 2, 3, 4, 5, 6, 8), labels = c("Under 17", "Married", "Widowed", "Never married", "Divorced", "Separated", "Unknown"))
nhefs$nerves <- factor(nhefs$nerves, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$nervousbreak <- factor(nhefs$nervousbreak, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$otherpain <- factor(nhefs$otherpain, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$pepticulcer <- factor(nhefs$pepticulcer, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$pica <- factor(nhefs$pica, levels = c(0, 1, 2), labels = c("Never", "Ever", "Missing"))
nhefs$polio <- factor(nhefs$polio, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$qsmk <- factor(nhefs$qsmk, levels = c(0, 1), labels = c("No", "Yes"))
nhefs$race <- factor(nhefs$race, levels = c(0, 1), labels = c("White", "Black or Other"))
nhefs$sex <- factor(nhefs$sex, levels = c(0, 1), labels = c("Male", "Female"))
nhefs$tb <- factor(nhefs$tb, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$tumor <- factor(nhefs$tumor, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$weakheart <- factor(nhefs$weakheart, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$wtloss <- factor(nhefs$wtloss, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$lackpep <- factor(nhefs$lackpep, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$infection <- factor(nhefs$infection, levels = c(0, 1), labels = c("Never", "Ever"))


mode_function <- function(x) {
  x <- as.character(x)
  tbl <- table(x, useNA = "ifany")
  as.character(names(tbl)[which.max(tbl)])
}

for (var in all_covars) {
  if (is.numeric(nhefs[[var]])) {
    # Impute missing values for numeric variables with the median
    nhefs[[var]][is.na(nhefs[[var]])] <- median(nhefs[[var]], na.rm = TRUE)
    print(paste("Imputed numeric variable:", var))
    
  } else if (is.factor(nhefs[[var]]) || is.character(nhefs[[var]])) {
    # Impute missing values for categorical variables with the mode
    nhefs[[var]][is.na(nhefs[[var]])] <- mode_function(nhefs[[var]])
    # nhefs[[var]] <- as.factor(nhefs[[var]])  # Ensure it's still a factor
    print(paste("Imputed categorical variable:", var))
  }
}

vars_to_include1 <- setdiff(names(nhefs), c('qsmk','wt82_71','sbp','dbp'))
```


## Question 1 (30 points total, 5 for each estimate)

```{r, include=F}
m1 <- glm(wt82_71 ~ qsmk * (sex + race + age + education + smokeintensity + smokeyrs + 
                 exercise + active + wt71 + hbp), data = nhefs)

m2 <- glm(sbp ~ qsmk * (sex + race + age + education + smokeintensity + smokeyrs + 
                 exercise + active + wt71 + hbp), data = nhefs)

m3 <- glm(dbp ~ qsmk * (sex + race + age + education + smokeintensity + smokeyrs + 
                 exercise + active + wt71 + hbp), data = nhefs)

p1 <- glm(qsmk ~ sex + race + poly(age, 2) + education + poly(smokeintensity, 2) + poly(smokeyrs, 2) + 
            exercise + active + poly(wt71, 2) + hbp, family = binomial(), data = nhefs)

# Regression
data_trt <- data_ctr <- nhefs[nhefs$qsmk == 'Yes', ]

data_trt$qsmk = 'Yes'; data_trt$qsmk <- as.factor(data_trt$qsmk)
data_ctr$qsmk = 'No'; data_ctr$qsmk <- as.factor(data_ctr$qsmk)

pred1_wc <- predict(m1, newdata = data_trt, type = "response")
pred0_wc <- predict(m1, newdata = data_ctr, type = "response")
(ATT_wc <- mean(pred1_wc - pred0_wc))

pred1_sbp <- predict(m2, newdata = data_trt, type = "response")
pred0_sbp <- predict(m2, newdata = data_ctr, type = "response")
(ATT_sbp <- mean(pred1_sbp - pred0_sbp))

pred1_dbp <- predict(m3, newdata = data_trt, type = "response")
pred0_dbp <- predict(m3, newdata = data_ctr, type = "response")
(ATT_dbp <- mean(pred1_dbp - pred0_dbp))


# PS regression
nhefs$ps <- predict(p1, type = "response")

m1.ps <- glm(wt82_71 ~ qsmk*rcs(ps, 5), data = nhefs)
m2.ps <- glm(sbp ~ qsmk*rcs(ps, 5), data = nhefs)
m3.ps <- glm(dbp ~ qsmk*rcs(ps, 5), data = nhefs)

data_trt_alt <- data_ctr_alt <- nhefs
data_trt_alt$qsmk = 'Yes'
data_ctr_alt$qsmk = 'No'

pred1.ps_wc <- predict(m1.ps, newdata = data_trt_alt, type = "response")
pred0.ps_wc <- predict(m1.ps, newdata = data_ctr_alt, type = "response")
(ATT.ps_wc <- mean(pred1.ps_wc[nhefs$qsmk == 'Yes'] - pred0.ps_wc[nhefs$qsmk == 'Yes']))

pred1.ps_sbp <- predict(m2.ps, newdata = data_trt_alt, type = "response")
pred0.ps_sbp <- predict(m2.ps, newdata = data_ctr_alt, type = "response")
ATT.ps_sbp <- mean(pred1.ps_sbp[nhefs$qsmk == 'Yes'] - pred0.ps_sbp[nhefs$qsmk == 'Yes'])

pred1.ps_dbp <- predict(m3.ps, newdata = data_trt_alt, type = "response")
pred0.ps_dbp <- predict(m3.ps, newdata = data_ctr_alt, type = "response")
ATT.ps_dbp <- mean(pred1.ps_dbp[nhefs$qsmk == 'Yes'] - pred0.ps_dbp[nhefs$qsmk == 'Yes'])

# PS stratification
ps <- predict(p1, type = "response")

ps_quintile <- cut(ps, 
	breaks = c(0, quantile(ps[nhefs$qsmk == 'Yes'], p = c(0.2, 0.4, 0.6, 0.8)), 1), labels = 1:5)
nA <- nrow(nhefs[nhefs$qsmk == 'Yes', ])
nAj <- table(ps_quintile[nhefs$qsmk == 'Yes'])

te_quintile_wc <- tapply(nhefs$wt82_71[nhefs$qsmk == 'Yes'], ps_quintile[nhefs$qsmk == 'Yes'], mean) -
	tapply(nhefs$wt82_71[nhefs$qsmk == 'No'], ps_quintile[nhefs$qsmk == 'No'], mean)
(ATT_PSS_wc <- sum(te_quintile_wc *nAj/nA))

te_quintile_sbp <- tapply(nhefs$sbp[nhefs$qsmk == 'Yes'], ps_quintile[nhefs$qsmk == 'Yes'], mean) -
	tapply(nhefs$sbp[nhefs$qsmk == 'No'], ps_quintile[nhefs$qsmk == 'No'], mean)
(ATT_PSS_sbp <- sum(te_quintile_sbp *nAj/nA))

te_quintile_dbp <- tapply(nhefs$dbp[nhefs$qsmk == 'Yes'], ps_quintile[nhefs$qsmk == 'Yes'], mean) -
	tapply(nhefs$dbp[nhefs$qsmk == 'No'], ps_quintile[nhefs$qsmk == 'No'], mean)
(ATT_PSS_dbp <- sum(te_quintile_dbp *nAj/nA))

# IPW
w1 <- as.numeric(nhefs$qsmk)-1
w0 <- (1-(as.numeric(nhefs$qsmk)-1))/(1-ps)*ps
(ATT_IPW_wc <- weighted.mean(nhefs$wt82_71, w = w1) - weighted.mean(nhefs$wt82_71, w=w0))
(ATT_IPW_sbp <- weighted.mean(nhefs$sbp, w = w1) - weighted.mean(nhefs$sbp, w=w0))
(ATT_IPW_dbp <- weighted.mean(nhefs$dbp, w = w1) - weighted.mean(nhefs$dbp, w=w0))

```

```{r, include=F}
## matching
# 1:1
rownames(nhefs) <- 1:nrow(nhefs)
mod_match1 <- matchit(qsmk ~ sex + race + poly(age, 2) + education + poly(smokeintensity, 2) + 
                          poly(smokeyrs, 2) + exercise + active + poly(wt71, 2) + hbp, 
            distance = "logit", method = "nearest", data = nhefs, ratio = 1)
print(tail(mod_match1$match.matrix))

nhefs_match <- data.frame(rbind(nhefs[nhefs$qsmk == 'Yes', ], nhefs[as.numeric(mod_match1$match.matrix), ]))
tabmatched1.1 <- CreateTableOne(vars = adj_vars, strata = "qsmk",
	data = nhefs_match, test = FALSE)
t1.1 <- print(tabmatched1.1, smd = TRUE, showAllLevels = TRUE, varLabels = TRUE)

plot(mod_match1, type = "hist")
# plot(mod_match1, type = "QQ")

n1 <- table(nhefs$qsmk)[2]
n0 <- length(as.numeric(mod_match1$match.matrix))

p1_wc <- mean(nhefs$wt82_71[nhefs$qsmk == 'Yes'])
p0_wc <- mean(nhefs$wt82_71[as.numeric(mod_match1$match.matrix)])
p1_sbp <- mean(nhefs$sbp[nhefs$qsmk == 'Yes'])
p0_sbp <- mean(nhefs$sbp[as.numeric(mod_match1$match.matrix)])
p1_dbp <- mean(nhefs$dbp[nhefs$qsmk == 'Yes'])
p0_dbp <- mean(nhefs$dbp[as.numeric(mod_match1$match.matrix)])

(ATT1match_wc <- p1_wc -  p0_wc)
(ATT1match_sbp <- p1_sbp -  p0_sbp)
(ATT1match_dbp <- p1_dbp -  p0_dbp)

# 2:1
mod_match2 <- matchit(qsmk ~ sex + race + poly(age, 2) + education + poly(smokeintensity, 2) + 
                          poly(smokeyrs, 2) + exercise + active + poly(wt71, 2) + hbp, 
            distance = "logit", method = "nearest", data = nhefs, ratio = 2)
print(tail(mod_match2$match.matrix))

nhefs_match2 <- data.frame(rbind(nhefs[nhefs$qsmk == 'Yes', ], nhefs[as.numeric(mod_match2$match.matrix), ]))
tabmatched2.1 <- CreateTableOne(vars = adj_vars, strata = "qsmk",
	data = nhefs_match2, test = FALSE)
t2.1 <- print(tabmatched2.1, smd = TRUE, showAllLevels = TRUE, varLabels = TRUE)

plot(mod_match2, type = "hist")
# plot(mod_match2, type = "QQ")

n1 <- table(nhefs$qsmk)[2]
n0_2.1 <- length(as.numeric(mod_match2$match.matrix))

p1_wc_2.1 <- mean(nhefs$wt82_71[nhefs$qsmk == 'Yes'])
p0_wc_2.1 <- mean(nhefs$wt82_71[as.numeric(mod_match2$match.matrix)])
p1_sbp_2.1 <- mean(nhefs$sbp[nhefs$qsmk == 'Yes'])
p0_sbp_2.1 <- mean(nhefs$sbp[as.numeric(mod_match2$match.matrix)])
p1_dbp_2.1 <- mean(nhefs$dbp[nhefs$qsmk == 'Yes'])
p0_dbp_2.1 <- mean(nhefs$dbp[as.numeric(mod_match2$match.matrix)])

(ATT2match_wc <- p1_wc_2.1 -  p0_wc_2.1)
(ATT2match_sbp <- p1_sbp_2.1 -  p0_sbp_2.1)
(ATT2match_dbp <- p1_dbp_2.1 -  p0_dbp_2.1)

```
```{r,echo=F}
ATT_reg <- c(ATT_wc, ATT_sbp, ATT_dbp)
ATT_ps <- c(ATT.ps_wc, ATT.ps_sbp, ATT.ps_dbp)
ATT_pss <- c(ATT_PSS_wc, ATT_PSS_sbp, ATT_PSS_dbp)
ATT_ipw <- c(ATT_IPW_wc, ATT_IPW_sbp, ATT_IPW_dbp)
ATT_match_11 <- c(ATT1match_wc, ATT1match_sbp, ATT1match_dbp)
ATT_match_21 <- c(ATT2match_wc, ATT2match_sbp, ATT2match_dbp)


tabletext <- rbind(c("", "Weight Change", "SBP", "DBP"),
	c("Regression", round(ATT_reg, digits = 3)),
	c("PS Regression", round(ATT_ps, digits = 3)),
	c("PSS", round(ATT_pss, digits=3)),
	c("IPW", round(ATT_ipw, digits=3)),
	c("Match 1:1", round(ATT_match_11, digits=3)),
	c("Match 2:1", round(ATT_match_21, digits=3))
	)

ATT_table <- as.data.frame(tabletext)
colnames(ATT_table) <- c("", "", "", "") 
kable(ATT_table, caption = "ATT estimates using different causal methods")
```
The model used for obtaining ATTs could be different (based on the best model you chose from the previous homework), but the estimates should be similar. 


## Question 2 (20 points total, 10 for plots and 10 for tables)

I created histograms, SMD plot and tableones for diagnosis of the matching process. Other diagnosis (including QQ-plots) are also accepted. 

We can see the matching process (for both 1:1 and 2:1 matching) gives us balance of the adjusted covariates between the 'qsmk' groups. 

```{r,echo=F}
# table diagnosis
kable(t1.1)
kable(t2.1)

# plot diagnosis
print("For 1:1 Matching")
plot(mod_match1, type = "hist")
print("For 2:1 Matching")
plot(mod_match2, type = "hist")


nhefs$weight_a <- with(nhefs, ifelse(qsmk == "Yes", 1 / ps, 1 / (1 - ps)))
dataSvy_a <- svydesign(ids = ~ 1, data = nhefs, weights = ~ weight_a)
tabUnweighted_a <- CreateTableOne(vars = adj_vars, strata = "qsmk", data = nhefs, test = FALSE)
tabWeighted_a <- svyCreateTableOne(vars = adj_vars, strata = "qsmk", data = dataSvy_a, test = FALSE)

# Extracting SMD values
dataPlot_a <- data.frame(variable  = rownames(ExtractSmd(tabUnweighted_a)),
                       Unweighted = as.numeric(ExtractSmd(tabUnweighted_a)),
                       IPW  = as.numeric(ExtractSmd(tabWeighted_a)))

# Convert to long format for ggplot
dataPlotMelt_a <- melt(data = dataPlot_a,
                     id.vars = c("variable"),
                     variable.name = "Method",
                     value.name = "SMD")

# Extracting SMD values
dataPlot_b <- data.frame(variable  = rownames(ExtractSmd(tabmatched1.1)),
                       Unweighted = as.numeric(ExtractSmd(tabmatched1.1)),
                       match_1.1  = as.numeric(ExtractSmd(tabmatched1.1)))

# Convert to long format for ggplot
dataPlotMelt_b <- melt(data = dataPlot_b,
                     id.vars = c("variable"),
                     variable.name = "Method",
                     value.name = "SMD")

# Extracting SMD values
dataPlot_c <- data.frame(variable  = rownames(ExtractSmd(tabmatched2.1)),
                       Unweighted = as.numeric(ExtractSmd(tabmatched2.1)),
                       match_2.1  = as.numeric(ExtractSmd(tabmatched2.1)))

# Convert to long format for ggplot
dataPlotMelt_c <- melt(data = dataPlot_c,
                     id.vars = c("variable"),
                     variable.name = "Method",
                     value.name = "SMD")


dataPlotMelt <- rbind(dataPlotMelt_a, dataPlotMelt_b[11:20, ], dataPlotMelt_c[11:20, ])
dataPlotMelt %>%
  arrange(desc(SMD), variable) %>%
  mutate(variable_ordered = factor(variable, levels = unique(variable))) %>%
  ggplot(aes(x = variable_ordered, y = SMD, group = Method, color = Method)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0.1, color = "black", linewidth = 0.1) +
  coord_flip() +
  theme_bw() +
  theme(legend.key = element_blank()) +  xlab('Variables') + 
  labs(title="Standardized Mean Differences for all Models")
```



# Appendix - R code

## Question 1 

1. Load packages and impute the dataset

```{r, eval=F}
set.seed(7485)

library(dplyr)
library(labelled)
library(causaldata)
library(knitr)
library(MASS)
library(reshape2)
library(ggplot2)
library(survey)
library(tableone)

library(formattable)
library(rms)
library(forestplot)
library(Matching)
library(MatchIt)

data('nhefs')

# treatment
nhefs$qsmk 
# outcomes of interest
nhefs$wt82_71
nhefs$sbp
nhefs$dbp

all_covars <- setdiff(names(nhefs), c('dadth', 'death', 'hightax82', 'modth', 'price71_82', 
                                      'price82', 'seqn', 'yrdth', 'wt82', 'tax71_82',
                                      'tax82', 'smokeyrs', 'smkintensity82_71',
                                      'birthplace','id','censored','older'))

adj_vars <- c('sex', 'race', 'age', 'education', 'smokeintensity', 'smokeyrs', 'exercise', 'active', 'wt71', 'hbp')


# Convert categorical variables to factors and set levels for them in nhefs data frame
nhefs$active <- factor(nhefs$active, levels = c(0, 1, 2), labels = c("Very active", "Moderately active", "Inactive"))
nhefs$alcoholfreq <- factor(nhefs$alcoholfreq, levels = c(0, 1, 2, 3, 4, 5), labels = c("Almost every day", "2-3 times/week", "1-4 times/month", "< 12 times/year", "No alcohol last year", "Unknown"))
nhefs$alcoholpy <- factor(nhefs$alcoholpy, levels = c(0, 1, 2), labels = c("Never", "Ever", "Missing"))
nhefs$alcoholtype <- factor(nhefs$alcoholtype, levels = c(1, 2, 3, 4), labels = c("Beer", "Wine", "Liquor", "Other/Unknown"))
nhefs$allergies <- factor(nhefs$allergies, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$asthma <- factor(nhefs$asthma, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$birthcontrol <- factor(nhefs$birthcontrol, levels = c(0, 1, 2), labels = c("No", "Yes", "Missing"))
nhefs$boweltrouble <- factor(nhefs$boweltrouble, levels = c(0, 1, 2), labels = c("Never", "Ever", "Missing"))
nhefs$bronch <- factor(nhefs$bronch, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$chroniccough <- factor(nhefs$chroniccough, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$colitis <- factor(nhefs$colitis, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$diabetes <- factor(nhefs$diabetes, levels = c(0, 1, 2), labels = c("Never", "Ever", "Missing"))
nhefs$education <- factor(nhefs$education, levels = c(1, 2, 3, 4, 5), labels = c("8th grade or less", "HS dropout", "HS", "College dropout", "College or more"))
nhefs$exercise <- factor(nhefs$exercise, levels = c(0, 1, 2), labels = c("Little or no exercise", "Moderate exercise", "Much exercise"))
nhefs$hayfever <- factor(nhefs$hayfever, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$hbp <- factor(nhefs$hbp, levels = c(0, 1, 2), labels = c("Never", "Ever", "Missing"))
nhefs$hbpmed <- factor(nhefs$hbpmed, levels = c(0, 1, 2), labels = c("Never", "Ever", "Missing"))
nhefs$headache <- factor(nhefs$headache, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$hepatitis <- factor(nhefs$hepatitis, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$hf <- factor(nhefs$hf, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$marital <- factor(nhefs$marital, levels = c(1, 2, 3, 4, 5, 6, 8), labels = c("Under 17", "Married", "Widowed", "Never married", "Divorced", "Separated", "Unknown"))
nhefs$nerves <- factor(nhefs$nerves, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$nervousbreak <- factor(nhefs$nervousbreak, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$otherpain <- factor(nhefs$otherpain, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$pepticulcer <- factor(nhefs$pepticulcer, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$pica <- factor(nhefs$pica, levels = c(0, 1, 2), labels = c("Never", "Ever", "Missing"))
nhefs$polio <- factor(nhefs$polio, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$qsmk <- factor(nhefs$qsmk, levels = c(0, 1), labels = c("No", "Yes"))
nhefs$race <- factor(nhefs$race, levels = c(0, 1), labels = c("White", "Black or Other"))
nhefs$sex <- factor(nhefs$sex, levels = c(0, 1), labels = c("Male", "Female"))
nhefs$tb <- factor(nhefs$tb, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$tumor <- factor(nhefs$tumor, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$weakheart <- factor(nhefs$weakheart, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$wtloss <- factor(nhefs$wtloss, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$lackpep <- factor(nhefs$lackpep, levels = c(0, 1), labels = c("Never", "Ever"))
nhefs$infection <- factor(nhefs$infection, levels = c(0, 1), labels = c("Never", "Ever"))


mode_function <- function(x) {
  x <- as.character(x)
  tbl <- table(x, useNA = "ifany")
  as.character(names(tbl)[which.max(tbl)])
}

for (var in all_covars) {
  if (is.numeric(nhefs[[var]])) {
    # Impute missing values for numeric variables with the median
    nhefs[[var]][is.na(nhefs[[var]])] <- median(nhefs[[var]], na.rm = TRUE)
    print(paste("Imputed numeric variable:", var))
    
  } else if (is.factor(nhefs[[var]]) || is.character(nhefs[[var]])) {
    # Impute missing values for categorical variables with the mode
    nhefs[[var]][is.na(nhefs[[var]])] <- mode_function(nhefs[[var]])
    # nhefs[[var]] <- as.factor(nhefs[[var]])  # Ensure it's still a factor
    print(paste("Imputed categorical variable:", var))
  }
}

```

2. Obtain ATT estimates

```{r, eval=F}
m1 <- glm(wt82_71 ~ qsmk * (sex + race + age + education + smokeintensity + smokeyrs + 
                 exercise + active + wt71 + hbp), data = nhefs)

m2 <- glm(sbp ~ qsmk * (sex + race + age + education + smokeintensity + smokeyrs + 
                 exercise + active + wt71 + hbp), data = nhefs)

m3 <- glm(dbp ~ qsmk * (sex + race + age + education + smokeintensity + smokeyrs + 
                 exercise + active + wt71 + hbp), data = nhefs)

p1 <- glm(qsmk ~ sex + race + poly(age, 2) + education + poly(smokeintensity, 2) + poly(smokeyrs, 2) + 
            exercise + active + poly(wt71, 2) + hbp, family = binomial(), data = nhefs)

# Regression
data_trt <- data_ctr <- nhefs[nhefs$qsmk == 'Yes', ]

data_trt$qsmk = 'Yes'; data_trt$qsmk <- as.factor(data_trt$qsmk)
data_ctr$qsmk = 'No'; data_ctr$qsmk <- as.factor(data_ctr$qsmk)

pred1_wc <- predict(m1, newdata = data_trt, type = "response")
pred0_wc <- predict(m1, newdata = data_ctr, type = "response")
(ATT_wc <- mean(pred1_wc - pred0_wc))

pred1_sbp <- predict(m2, newdata = data_trt, type = "response")
pred0_sbp <- predict(m2, newdata = data_ctr, type = "response")
(ATT_sbp <- mean(pred1_sbp - pred0_sbp))

pred1_dbp <- predict(m3, newdata = data_trt, type = "response")
pred0_dbp <- predict(m3, newdata = data_ctr, type = "response")
(ATT_dbp <- mean(pred1_dbp - pred0_dbp))


# PS regression
nhefs$ps <- predict(p1, type = "response")

m1.ps <- glm(wt82_71 ~ qsmk*rcs(ps, 5), data = nhefs)
m2.ps <- glm(sbp ~ qsmk*rcs(ps, 5), data = nhefs)
m3.ps <- glm(dbp ~ qsmk*rcs(ps, 5), data = nhefs)

data_trt_alt <- data_ctr_alt <- nhefs
data_trt_alt$qsmk = 'Yes'
data_ctr_alt$qsmk = 'No'

pred1.ps_wc <- predict(m1.ps, newdata = data_trt_alt, type = "response")
pred0.ps_wc <- predict(m1.ps, newdata = data_ctr_alt, type = "response")
(ATT.ps_wc <- mean(pred1.ps_wc[nhefs$qsmk == 'Yes'] - pred0.ps_wc[nhefs$qsmk == 'Yes']))

pred1.ps_sbp <- predict(m2.ps, newdata = data_trt_alt, type = "response")
pred0.ps_sbp <- predict(m2.ps, newdata = data_ctr_alt, type = "response")
ATT.ps_sbp <- mean(pred1.ps_sbp[nhefs$qsmk == 'Yes'] - pred0.ps_sbp[nhefs$qsmk == 'Yes'])

pred1.ps_dbp <- predict(m3.ps, newdata = data_trt_alt, type = "response")
pred0.ps_dbp <- predict(m3.ps, newdata = data_ctr_alt, type = "response")
ATT.ps_dbp <- mean(pred1.ps_dbp[nhefs$qsmk == 'Yes'] - pred0.ps_dbp[nhefs$qsmk == 'Yes'])

# PS stratification
ps <- predict(p1, type = "response")

ps_quintile <- cut(ps, 
	breaks = c(0, quantile(ps[nhefs$qsmk == 'Yes'], p = c(0.2, 0.4, 0.6, 0.8)), 1), labels = 1:5)
nA <- nrow(nhefs[nhefs$qsmk == 'Yes', ])
nAj <- table(ps_quintile[nhefs$qsmk == 'Yes'])

te_quintile_wc <- tapply(nhefs$wt82_71[nhefs$qsmk == 'Yes'], ps_quintile[nhefs$qsmk == 'Yes'], mean) -
	tapply(nhefs$wt82_71[nhefs$qsmk == 'No'], ps_quintile[nhefs$qsmk == 'No'], mean)
(ATT_PSS_wc <- sum(te_quintile_wc *nAj/nA))

te_quintile_sbp <- tapply(nhefs$sbp[nhefs$qsmk == 'Yes'], ps_quintile[nhefs$qsmk == 'Yes'], mean) -
	tapply(nhefs$sbp[nhefs$qsmk == 'No'], ps_quintile[nhefs$qsmk == 'No'], mean)
(ATT_PSS_sbp <- sum(te_quintile_sbp *nAj/nA))

te_quintile_dbp <- tapply(nhefs$dbp[nhefs$qsmk == 'Yes'], ps_quintile[nhefs$qsmk == 'Yes'], mean) -
	tapply(nhefs$dbp[nhefs$qsmk == 'No'], ps_quintile[nhefs$qsmk == 'No'], mean)
(ATT_PSS_dbp <- sum(te_quintile_dbp *nAj/nA))

# IPW
w1 <- as.numeric(nhefs$qsmk)-1
w0 <- (1-(as.numeric(nhefs$qsmk)-1))/(1-ps)*ps
(ATT_IPW_wc <- weighted.mean(nhefs$wt82_71, w = w1) - weighted.mean(nhefs$wt82_71, w=w0))
(ATT_IPW_sbp <- weighted.mean(nhefs$sbp, w = w1) - weighted.mean(nhefs$sbp, w=w0))
(ATT_IPW_dbp <- weighted.mean(nhefs$dbp, w = w1) - weighted.mean(nhefs$dbp, w=w0))

## matching
# 1:1
rownames(nhefs) <- 1:nrow(nhefs)
mod_match1 <- matchit(qsmk ~ sex + race + poly(age, 2) + education + poly(smokeintensity, 2) + 
                          poly(smokeyrs, 2) + exercise + active + poly(wt71, 2) + hbp, 
            distance = "logit", method = "nearest", data = nhefs, ratio = 1)
print(tail(mod_match1$match.matrix))

nhefs_match <- data.frame(rbind(nhefs[nhefs$qsmk == 'Yes', ], nhefs[as.numeric(mod_match1$match.matrix), ]))
tabmatched1.1 <- CreateTableOne(vars = adj_vars, strata = "qsmk",
	data = nhefs_match, test = FALSE)
t1.1 <- print(tabmatched1.1, smd = TRUE, showAllLevels = TRUE, varLabels = TRUE)

# plot(mod_match1, type = "hist")
# plot(mod_match1, type = "QQ")

n1 <- table(nhefs$qsmk)[2]
n0 <- length(as.numeric(mod_match1$match.matrix))

p1_wc <- mean(nhefs$wt82_71[nhefs$qsmk == 'Yes'])
p0_wc <- mean(nhefs$wt82_71[as.numeric(mod_match1$match.matrix)])
p1_sbp <- mean(nhefs$sbp[nhefs$qsmk == 'Yes'])
p0_sbp <- mean(nhefs$sbp[as.numeric(mod_match1$match.matrix)])
p1_dbp <- mean(nhefs$dbp[nhefs$qsmk == 'Yes'])
p0_dbp <- mean(nhefs$dbp[as.numeric(mod_match1$match.matrix)])

(ATT1match_wc <- p1_wc -  p0_wc)
(ATT1match_sbp <- p1_sbp -  p0_sbp)
(ATT1match_dbp <- p1_dbp -  p0_dbp)

# 2:1
mod_match2 <- matchit(qsmk ~ sex + race + poly(age, 2) + education + poly(smokeintensity, 2) + 
                          poly(smokeyrs, 2) + exercise + active + poly(wt71, 2) + hbp, 
            distance = "logit", method = "nearest", data = nhefs, ratio = 2)
print(tail(mod_match2$match.matrix))

nhefs_match2 <- data.frame(rbind(nhefs[nhefs$qsmk == 'Yes', ], nhefs[as.numeric(mod_match2$match.matrix), ]))
tabmatched2.1 <- CreateTableOne(vars = adj_vars, strata = "qsmk",
	data = nhefs_match2, test = FALSE)
t2.1 <- print(tabmatched2.1, smd = TRUE, showAllLevels = TRUE, varLabels = TRUE)

# plot(mod_match2, type = "hist")
# plot(mod_match2, type = "QQ")

n1 <- table(nhefs$qsmk)[2]
n0_2.1 <- length(as.numeric(mod_match2$match.matrix))

p1_wc_2.1 <- mean(nhefs$wt82_71[nhefs$qsmk == 'Yes'])
p0_wc_2.1 <- mean(nhefs$wt82_71[as.numeric(mod_match2$match.matrix)])
p1_sbp_2.1 <- mean(nhefs$sbp[nhefs$qsmk == 'Yes'])
p0_sbp_2.1 <- mean(nhefs$sbp[as.numeric(mod_match2$match.matrix)])
p1_dbp_2.1 <- mean(nhefs$dbp[nhefs$qsmk == 'Yes'])
p0_dbp_2.1 <- mean(nhefs$dbp[as.numeric(mod_match2$match.matrix)])

(ATT2match_wc <- p1_wc_2.1 -  p0_wc_2.1)
(ATT2match_sbp <- p1_sbp_2.1 -  p0_sbp_2.1)
(ATT2match_dbp <- p1_dbp_2.1 -  p0_dbp_2.1)

ATT_reg <- c(ATT_wc, ATT_sbp, ATT_dbp)
ATT_ps <- c(ATT.ps_wc, ATT.ps_sbp, ATT.ps_dbp)
ATT_pss <- c(ATT_PSS_wc, ATT_PSS_sbp, ATT_PSS_dbp)
ATT_ipw <- c(ATT_IPW_wc, ATT_IPW_sbp, ATT_IPW_dbp)
ATT_match_11 <- c(ATT1match_wc, ATT1match_sbp, ATT1match_dbp)
ATT_match_21 <- c(ATT2match_wc, ATT2match_sbp, ATT2match_dbp)


tabletext <- rbind(c("", "Weight Change", "SBP", "DBP"),
	c("Regression", round(ATT_reg, digits = 3)),
	c("PS Regression", round(ATT_ps, digits = 3)),
	c("PSS", round(ATT_pss, digits=3)),
	c("IPW", round(ATT_ipw, digits=3)),
	c("Match 1:1", round(ATT_match_11, digits=3)),
	c("Match 2:1", round(ATT_match_21, digits=3))
	)

kable(as.data.frame(tabletext))
```

## Question 2 

1. Table and plot diagnosis for the matching process

```{r, eval=F}
# table diagnosis
kable(t1.1)
kable(t2.1)

# plot diagnosis
plot(mod_match1, type = "hist")
plot(mod_match2, type = "hist")


# SMD plot
nhefs$weight_a <- with(nhefs, ifelse(qsmk == "Yes", 1 / ps, 1 / (1 - ps)))
dataSvy_a <- svydesign(ids = ~ 1, data = nhefs, weights = ~ weight_a)
tabUnweighted_a <- CreateTableOne(vars = adj_vars, strata = "qsmk", data = nhefs, test = FALSE)
tabWeighted_a <- svyCreateTableOne(vars = adj_vars, strata = "qsmk", data = dataSvy_a, test = FALSE)

dataPlot_a <- data.frame(variable  = rownames(ExtractSmd(tabUnweighted_a)),
                       Unweighted = as.numeric(ExtractSmd(tabUnweighted_a)),
                       IPW  = as.numeric(ExtractSmd(tabWeighted_a)))

dataPlotMelt_a <- melt(data = dataPlot_a,
                     id.vars = c("variable"),
                     variable.name = "Method",
                     value.name = "SMD")

dataPlot_b <- data.frame(variable  = rownames(ExtractSmd(tabmatched1.1)),
                       Unweighted = as.numeric(ExtractSmd(tabmatched1.1)),
                       match_1.1  = as.numeric(ExtractSmd(tabmatched1.1)))

dataPlotMelt_b <- melt(data = dataPlot_b,
                     id.vars = c("variable"),
                     variable.name = "Method",
                     value.name = "SMD")

dataPlot_c <- data.frame(variable  = rownames(ExtractSmd(tabmatched2.1)),
                       Unweighted = as.numeric(ExtractSmd(tabmatched2.1)),
                       match_2.1  = as.numeric(ExtractSmd(tabmatched2.1)))

dataPlotMelt_c <- melt(data = dataPlot_c,
                     id.vars = c("variable"),
                     variable.name = "Method",
                     value.name = "SMD")


dataPlotMelt <- rbind(dataPlotMelt_a, dataPlotMelt_b[11:20, ], dataPlotMelt_c[11:20, ])
dataPlotMelt %>%
  arrange(desc(SMD), variable) %>%
  mutate(variable_ordered = factor(variable, levels = unique(variable))) %>%
  ggplot(aes(x = variable_ordered, y = SMD, group = Method, color = Method)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0.1, color = "black", linewidth = 0.1) +
  coord_flip() +
  theme_bw() +
  theme(legend.key = element_blank()) +  xlab('Variables') + 
  labs(title="Standardized Mean Differences for all Models")

```


