---
title: "Time-Varying Treatments and Marginal Structural Models"
author: "David M. Vock"
format: beamer
theme: "Boadilla"
colortheme: "whale"
fonttheme: "structurebold"
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Proposenity Score: A Review for Single Time Point

- Defined for a binary treatment to be $P(A = 1|X) = \pi(X)$
- Under the assumption of no unmeasured confounding and consistency, the propensity score is a balancing score
- A balancing score $b(X)$ is any function of the covariates $X$ such that $A$ is independent of $X$ given $b(X)$
- This means the distribution of the covariates is the same in the treatment and control group at the same level of $b(X)$
- Note that $P(A = 1|X) = P(A = 1 |W)$ --> show this


## Proposenity Score: Multiple Time Points
- Note that it is difficult (impossible?) to talk about a single propensity score when there are multiple time points
- If we fix $\overline{a}$ then we will define the propensity score (for that sequence of treatments) as $P(\overline{A} = \overline{a}| W)$ where $W$ denotes the set of all potential outcomes
- Under consistency and sequential ignorability one can show that 

$P(\overline{A} = \overline{a}| W) = P(A_0 = a_0|L_0) \times P(A_1 = a_1|A_0 = a_0, \overline{L}_1) \times \ldots \times P(A_M = a_m| \overline{A}_{M-1}, \overline{L}_M)$

- That is, the propensity score (for sequence of treatments $\overline{a}$) is the product that at each timepoint a person takes treatment $a_j$ given all the information previously collected on them

## Model for Probability of Receiving Treatment
```{r, echo = FALSE, warning = FALSE, message = FALSE, cache = TRUE}
# code from g-computation
# will be used as comparison
library(dplyr)
beta0 <- 6.2
beta1 <- 1.6
beta2 <- -0.23
beta3 <- -0.175
beta4 <- -0.5
sigma0 <- sigma <- 0.35
gamma0 <- -3.21
gamma1 <- 0.80

eta0 <- 3.68
eta1 <- -0.58
eta2 <- -1.18

n <- 1000

expit <- function(x) {
  expit <- exp(x)/(1 + exp(x))
  return(expit)
}

set.seed(1101985)

R0 <- rep(0, n)
logCD40 <- rnorm(n, beta0, sigma0)
A0<- rbinom(n, 1, expit(eta0 + eta1*logCD40 + eta2*R0))
cumA0 <- A0

R1 <- rbinom(n, 1, expit(gamma0 + gamma1*cumA0))*(cumA0 > 0 & R0 == 0) + 0*(cumA0 == 0) + 1*(R0 == 1) 
logCD41 <- logCD40 + rnorm(n, (beta1*A0 + beta2*A0*logCD40 + beta3*(1-A0))*(1-R1) + beta4*R1, sigma)
A1 <- rbinom(n, 1, expit(eta0 + eta1*logCD41 + eta2*R1))
cumA1 <- A0 + A1

R2 <- rbinom(n, 1, expit(gamma0 + gamma1*cumA1))*(cumA1 > 0 & R1 == 0) + 0*(cumA1 == 0) + 1*(R1 == 1) 
logCD42 <- logCD41 + rnorm(n, (beta1*A1 + beta2*A1*logCD41 + beta3*(1-A1))*(1-R2) + beta4*R2, sigma)
A2 <- rbinom(n, 1, expit(eta0 + eta1*logCD42 + eta2*R2))
cumA2 <- A0 + A1 + A2

R3 <- rbinom(n, 1, expit(gamma0 + gamma1*cumA2))*(cumA2 > 0 & R2 == 0) + 0*(cumA2 == 0) + 1*(R2 == 1) 
logCD43 <- logCD42 + rnorm(n, (beta1*A2 + beta2*A2*logCD42 + beta3*(1-A2))*(1-R3) + beta4*R3, sigma)
A3 <- rbinom(n, 1, expit(eta0 + eta1*logCD43 + eta2*R3))
cumA3 <- A0 + A1 + A2 + A3

R4 <- rbinom(n, 1, expit(gamma0 + gamma1*cumA3))*(cumA3 > 0 & R3 == 0) + 0*(cumA3 == 0) + 1*(R3 == 1) 
logCD44 <- logCD43 + rnorm(n, (beta1*A3 + beta2*A3*logCD43 + beta3*(1-A3))*(1-R4) + beta4*R4, sigma)

dataHW3 <- data.frame(ID = 1:n, R0 = R0, CD40 = exp(logCD40), A0 = A0,
  R1 = R1, CD41 = exp(logCD41), A1 = A1,
  R2 = R2, CD42 = exp(logCD42), A2 = A2,
  R3 = R3, CD43 = exp(logCD43), A3 = A3,
  R4 = R4, CD44 = exp(logCD44))

# create log CD4 values
dataHW3 <- mutate(dataHW3,
	logCD40 = log(CD40),
	logCD41 = log(CD41),
	logCD42 = log(CD42),
	logCD43 = log(CD43),
	logCD44 = log(CD44)
	)

# get estimates of mu0 and sigma0
mu0 <- mean(dataHW3$logCD40)
sigma0 <- sd(dataHW3$logCD40)

# transform into long dataset
library(data.table)

dataHW3_long <- melt(setDT(dataHW3), id = 1L,
	measure = patterns("^logCD4", "^CD4", "^R", "^A"),
	value.name = c("logCD4", "CD4", "R", "A"))
dataHW3_long <- data.frame(dataHW3_long)
dataHW3_long <- dataHW3_long[order(dataHW3_long$ID), ]
dataHW3_long <- 
	dataHW3_long %>%
	group_by(ID) %>%
	mutate(lag_logCD4 = lag(logCD4, n = 1, default = NA),
		lag_A = lag(A, n = 1, default = NA),
		cum_A = cumsum(A),
		lag_cum_A = lag(cum_A, n = 1, default = NA),
		lag_R = lag(R, n = 1, default = NA))

dataHW3_long <- mutate(dataHW3_long,
	time = as.numeric(variable) - 1,
	logCD4_diff = logCD4 - lag_logCD4)

# When estimating the model for change in log CD4 count and developing 
# resistance, only need data from time points 1-4 (not time point 0).

dataHW3_long_t0 <- filter(dataHW3_long, time > 0)

# Estimate parameters in the model for change in CD4

m1 <- lm(logCD4 ~ lag_logCD4 + lag_A + R +
		lag_logCD4*lag_A + lag_logCD4*R + 
		lag_A*R + lag_logCD4*lag_A*R, data = dataHW3_long_t0)

# Estimate parameters in the model for resistance
# Remember subjects who have no prior exposure to treatment will not be
# resistant and subjects who previously developed resistance are guaranteed
# to be resistant. Neither of these groups should be included in the model.

dataHW3_long_resist_model <- filter(dataHW3_long_t0, lag_R == 0 & lag_cum_A > 0)

m2 <- glm(R ~ lag_cum_A, family = "binomial", 
	data = dataHW3_long_resist_model)

coef_estimates <- c(mu0, sigma0, coef(m1), summary(m1)$sigma, coef(m2))
coef_names <- c("mu0", "sigma0", paste0("beta", 0:7), "sigma", 
	paste0("gamma", 0:1))
coef_estimates <- data.frame(Estimates = coef_estimates)
rownames(coef_estimates) <- coef_names

mean_response <- function(treat_sequence, coef_estimates, MC) {
	mu0 <- coef_estimates[1]; sigma0 <- coef_estimates[2]
	beta0 <- coef_estimates[3]; beta1 <- coef_estimates[4]; beta2 <- coef_estimates[5]; 
	beta3 <- coef_estimates[6]; beta4 <- coef_estimates[7];
	beta5 <- coef_estimates[8]; beta6 <- coef_estimates[9];
	beta7 <- coef_estimates[10]; 
	
	sigma <- coef_estimates[11];gamma0 <- coef_estimates[12];gamma1 <- coef_estimates[13]
	
		R0 <- rep(0, MC)
		logCD40 <- rnorm(MC, mu0, sigma0)
		A0 <- rep(treat_sequence[1], MC)
		cumA0 <- A0
		R1 <- rbinom(MC, 1, expit(gamma0 + gamma1*cumA0))*(cumA0 > 0 & R0 == 0) + 
			0*(cumA0 == 0) + 1*(R0 == 1) 
		logCD41 <- beta0 + beta1*logCD40 + beta2*A0 + beta3*R1 + 
			beta4*logCD40*A0 + beta5*logCD40*R1 + beta6*A0*R1 + beta7*logCD40*A0*R1 + 
			rnorm(MC, 0, sigma)
		A1 <- rep(treat_sequence[2], MC)
		cumA1 <- A0 + A1
		R2 <- rbinom(MC, 1, expit(gamma0 + gamma1*cumA1))*(cumA1 > 0 & R1 == 0) + 
			0*(cumA1 == 0) + 1*(R1 == 1) 
		logCD42 <-  beta0 + beta1*logCD41 + beta2*A1 + beta3*R2 + 
			beta4*logCD41*A1 + beta5*logCD41*R2 + beta6*A1*R2 + beta7*logCD41*A1*R2 + 
			rnorm(MC, 0, sigma)
		A2 <- rep(treat_sequence[3], MC)
		cumA2 <- A0 + A1 + A2
		R3 <- rbinom(MC, 1, expit(gamma0 + gamma1*cumA2))*(cumA2 > 0 & R2 == 0) + 
			0*(cumA2 == 0) + 1*(R2 == 1) 
		logCD43 <- beta0 + beta1*logCD42 + beta2*A2 + beta3*R3 + 
			beta4*logCD42*A2 + beta5*logCD42*R3 + beta6*A2*R3 + beta7*logCD42*A2*R3 + 
			rnorm(MC, 0, sigma)
		A3 <- rep(treat_sequence[4], MC)
		cumA3 <- A0 + A1 + A2 + A3
		R4 <- rbinom(MC, 1, expit(gamma0 + gamma1*cumA3))*(cumA3 > 0 & R3 == 0) + 
			0*(cumA3 == 0) + 1*(R3 == 1) 
		logCD44 <-  beta0 + beta1*logCD43 + beta2*A3 + beta3*R4 + 
			beta4*logCD43*A3 + beta5*logCD43*R4 + beta6*A3*R4 + beta7*logCD43*A3*R4 + 
			rnorm(MC, 0, sigma)
		
	mean_response <- c(mean(logCD44), mean(exp(logCD44)))
	names(mean_response) <- c("mean_logCD4", "mean_CD4")
	return(mean_response)
}

trt_sequences <- list(
	c(0, 0, 0, 0),
	c(0, 0, 0, 1),
	c(0, 0, 1, 1),
	c(0, 1, 1, 1),
	c(1, 1, 1, 1),
	c(1, 1, 1, 0),
	c(1, 1, 0, 0),
	c(1, 0, 0, 0)
)
set.seed(8172013)
mean_trt_response <- sapply(trt_sequences, mean_response, 
	coef_estimates = as.vector(coef_estimates[, 1]), MC = 100000)	
mean_trt_response <- t(mean_trt_response)
rownames(mean_trt_response) <- c("(0, 0, 0, 0)",
	"(0, 0, 0, 1)",
	"(0, 0, 1, 1)",
	"(0, 1, 1, 1)",
	"(1, 1, 1, 1)",
	"(1, 1, 1, 0)",
	"(1, 1, 0, 0)",
	"(1, 0, 0, 0)")

dataHW3_orig <- dataHW3
n <- nrow(dataHW3_orig)
boot <- 50

mean_trt_response_boot <- array(0, dim = c(8, 2, boot))
for(j in 1:boot) {
	dataHW3 <- dataHW3_orig[sample(1:n, n, replace = TRUE), ]
	
	# get estimates of mu0 and sigma0
	mu0 <- mean(dataHW3$logCD40)
	sigma0 <- sd(dataHW3$logCD40)

	dataHW3$ID <- 1:n
	dataHW3_long <- melt(setDT(dataHW3), id = 1L,
		measure = patterns("^logCD4", "^CD4", "^R", "^A"),
		value.name = c("logCD4", "CD4", "R", "A"))
	dataHW3_long <- data.frame(dataHW3_long)
	dataHW3_long <- dataHW3_long[order(dataHW3_long$ID), ]
	dataHW3_long <- 
		dataHW3_long %>%
		group_by(ID) %>%
		mutate(lag_logCD4 = lag(logCD4, n = 1, default = NA),
			lag_A = lag(A, n = 1, default = NA),
			cum_A = cumsum(A),
			lag_cum_A = lag(cum_A, n = 1, default = NA),
			lag_R = lag(R, n = 1, default = NA))
	
	dataHW3_long <- mutate(dataHW3_long,
		time = as.numeric(variable) - 1,
		logCD4_diff = logCD4 - lag_logCD4)
	dataHW3_long_t0 <- filter(dataHW3_long, time > 0)
	
	m1 <- lm(logCD4 ~ lag_logCD4 + lag_A + R +
		lag_logCD4*lag_A + lag_logCD4*R + 
		lag_A*R + lag_logCD4*lag_A*R, data = dataHW3_long_t0)
	
	dataHW3_long_resist_model <- filter(dataHW3_long_t0, lag_R == 0 & lag_cum_A > 0)
	
	m2 <- glm(R ~ lag_cum_A, family = "binomial", 
		data = dataHW3_long_resist_model)
	
	coef_estimates <- c(mu0, sigma0, coef(m1), summary(m1)$sigma, coef(m2))
	
	mean_trt_response <- sapply(trt_sequences, mean_response, 
		coef_estimates = coef_estimates, MC = 100000)	
	mean_trt_response <- t(mean_trt_response)
	
	mean_trt_response_boot[, , j] <- mean_trt_response 
}

se <- apply(mean_trt_response_boot, c(1, 2), sd)
rownames(se) <- c("(0, 0, 0, 0)",
	"(0, 0, 0, 1)",
	"(0, 0, 1, 1)",
	"(0, 1, 1, 1)",
	"(1, 1, 1, 1)",
	"(1, 1, 1, 0)",
	"(1, 1, 0, 0)",
	"(1, 0, 0, 0)")
colnames(se) <- c("SE_logCD4", "SE_CD4")
```


```{r, echo = FALSE, cache = TRUE}
dataHW3 <- dataHW3_orig
dataHW3 <- mutate(dataHW3,
	logCD40 = log(CD40),
	logCD41 = log(CD41),
	logCD42 = log(CD42),
	logCD43 = log(CD43),
	logCD44 = log(CD44)
	)
dataHW3_long <- melt(setDT(dataHW3), id = 1L,
	measure = patterns("^logCD4", "^CD4", "^R", "^A"),
	value.name = c("logCD4", "CD4", "R", "A"))
dataHW3_long <- data.frame(dataHW3_long)
dataHW3_long <- dataHW3_long[order(dataHW3_long$ID), ]
dataHW3_long <- 
	dataHW3_long %>%
	group_by(ID) %>%
	mutate(lag_logCD4 = lag(logCD4, n = 1, default = NA),
		lag_A = lag(A, n = 1, default = NA),
		cum_A = cumsum(A),
		lag_cum_A = lag(cum_A, n = 1, default = NA),
		lag_R = lag(R, n = 1, default = NA))

# model for probability of receiving treatment
dataHW3_long <- mutate(dataHW3_long,
	time = as.numeric(variable) - 1,
	logCD4_diff = logCD4 - lag_logCD4)
dataHW3_long <- as.data.frame(dataHW3_long)

m3 <- glm(A ~ logCD4 + R, family = "binomial", 
		data = dataHW3_long)
round(summary(m3)$coefficients, digits = 2)
```

## Inverse Probability Weighting

- Assume that we have a sample of observed data $(\overline{L}_i, \overline{A}_i, Y_i)$ for $i = 1, \ldots, n$
- We can derive a consistent estimator for $E(Y^{\overline{a}})$ by using
\begin{equation}
n^{-1} \sum_{i=1}^n \frac{I(\overline{A}_i = \overline{a})}{\prod_{j=0}^M P_{A_j|\overline{A}_{j-1}, \overline{L}_j}(a_j|\overline{a}_{j-1}, \overline{L}_{ij})}Y_i
\end{equation}
- SHOW this is consistent


## Inverse Probability Weighting for Sequence (0, 1, 1, 1)

```{r, echo = TRUE, cache = TRUE}
dataHW3_long <-  mutate(dataHW3_long, pred_trt = predict(m3, newdata = dataHW3_long, type = "response"),
	pred_trt_rule = ifelse(time > 0, pred_trt, 1-pred_trt),
	trt_rule = ifelse((time == 0 & A == 0)|(time > 0 & A ==1), 1, 0) )
den_wt <- tapply(dataHW3_long$pred_trt_rule, 
	dataHW3_long$ID, prod, na.rm = TRUE)
num_wt <- tapply(dataHW3_long$trt_rule, 
	dataHW3_long$ID, prod, na.rm = TRUE)
sum(num_wt)
round(weighted.mean(dataHW3$CD44, w = num_wt/den_wt))
```

## Limitations of Inverse Probability Weighting

- In order for this method to be useful, we cannot have too many decision points $M$ or many decision options at each point
- Otherwise, the number of individuals such that $I(\overline{A}_i =  \overline{a})$ becomes so small as to give very unstable estimators
- For example, if a binary treatment were randomly assigned with 50% probability at 4 decision points, the number of subjects such that $I(\overline{A}_i =  \overline{a})$ would be 1/16 the total sample size
- Typically, we avoid this problem of small sample sizes through modeling. For example, we use additive regression models rather than fit separate models for each covariate combination

## Modeling potential outcomes
- Consider the example in the homework - binary treatment at 4 decision points (16 different treatment combinations)
- Could (in theory) estimate a different expected outcome for each treatment sequence
- BUT we might have more precision if we could pool some of these treatment sequences together

## Marginal Structural Model

- Let's break down the name
- Structural = having to do with potential outcomes
- Marginal = not covariate adjusted
- Overall, this is a proposed model for (usually) the mean of potential outcomes based on combining different treatment sequences together
- Example: $E(Y^{\overline{a}}) = \beta_0 + \beta_1 cum(\overline{a})$ where $cum(\overline{a}) = \sum_{l=0}^M a_l$. Instead of estimating 
- Other possibilities?
- Note that at this point I have not discussed estimation 

## Estimating a Marginal Structural Model

- MSM are models for potential outcomes (which we may not observe)
- If we want to use observed data to estimate (functionals of) potential outcomes need to worry about confounding 
- In other words, we cannot consistently estimate $\beta_0$ and $\beta_1$ by fitting a regression model with $Y_i$ as the outcome and $cum(\overline{A}_i)$ as the predictor
- Instead we need to use weighted regression with weights $\frac{1}{\prod_{j=0}^M P_{A_j|\overline{A}_{j-1}, \overline{L}_j}(A_{ij}|\overline{A}_{i,j-1}, \overline{L}_{ij})}$
- This is one over the probability that a particular subject would have followed their observed treatment history 


## Marginal Structural Model for Treatment Initiation

- In the HIV example, want to compare different different times for initiating (and then sustaining) treatment
- Compare the regimes (0, 0, 0, 0); (0, 0, 0, 1); (0, 0, 1, 1); (0, 1, 1, 1); (1, 1, 1, 1);
- Note that there are 5 different regimes here so a saturated model would have 5 parameters
- Consider time period when treatment was initiated as predictor with additional quadratic term (3 parameter model including intercept); never initiators have time set to 4


## Marginal Structural Model - Denominator Weights
```{r, echo = TRUE, cache = TRUE}
dataHW3_long <-   mutate(dataHW3_long,
	pred_trt = predict(m3, newdata = dataHW3_long, type = "response"),
	pred_trt_received = (pred_trt^A)*(1-pred_trt)^(1-A))
# calculate denominator weights
dataHW3$den_wt <- tapply(dataHW3_long$pred_trt_received, 
	dataHW3_long$ID, prod, na.rm = TRUE)
```

## Marginal Structural Model - Initiation Time
```{r, echo = TRUE, cache = TRUE, size = "small"}
dataHW3 <- mutate(dataHW3,
	init_time =    NA,
	init_time = ifelse(A0 == 1 & A1 == 1 & A2 == 1 & A3 == 1, 0, init_time),
	init_time = ifelse(A0 == 0 & A1 == 1 & A2 == 1 & A3 == 1, 1, init_time),
	init_time = ifelse(A0 == 0 & A1 == 0 & A2 == 1 & A3 == 1, 2, init_time),
	init_time = ifelse(A0 == 0 & A1 == 0 & A2 == 0 & A3 == 1, 3, init_time),
	init_time = ifelse(A0 == 0 & A1 == 0 & A2 == 0 & A3 == 0, 4, init_time))
```

## Estimate Marginal Structural Model
```{r, echo = TRUE, cache = TRUE}
MSM_CD4 <- lm(CD44 ~   init_time + I(init_time^2), data = dataHW3)
round(summary(MSM_CD4)$coefficients, digits = 2)
```

## Estimate Marginal Structural Model
```{r, echo = TRUE, cache = TRUE}
newdatatime <-   data.frame(init_time = 4:0)
predCD4 <- data.frame(CD4 = predict(MSM_CD4, newdatatime))
rownames(predCD4) <- c("(0, 0, 0, 0)",
	"(0, 0, 0, 1)",
	"(0, 0, 1, 1)",
	"(0, 1, 1, 1)",
	"(1, 1, 1, 1)")
print(round(predCD4))
```
## Obtain Standard Errors Using Bootstrap

```{r, echo = FALSE, cache = TRUE}
dataHW3_orig <- dataHW3
n <- nrow(dataHW3_orig)
boot <- 50
set.seed(8172013)
predCD4.boot <- NULL
for(j in 1:boot) {
	dataHW3 <- dataHW3_orig[sample(1:n, n, replace = TRUE), ]
	dataHW3$ID <- 1:n
	dataHW3 <- mutate(dataHW3,
	logCD40 = log(CD40),
	logCD41 = log(CD41),
	logCD42 = log(CD42),
	logCD43 = log(CD43),
	logCD44 = log(CD44)
	)

	dataHW3_long <- melt(setDT(dataHW3), id = 1L,
	measure = patterns("^logCD4", "^CD4", "^R", "^A"),
	value.name = c("logCD4", "CD4", "R", "A"))
dataHW3_long <- data.frame(dataHW3_long)
dataHW3_long <- dataHW3_long[order(dataHW3_long$ID), ]
dataHW3_long <- 
	dataHW3_long %>%
	group_by(ID) %>%
	mutate(lag_logCD4 = lag(logCD4, n = 1, default = NA),
		lag_A = lag(A, n = 1, default = NA),
		cum_A = cumsum(A),
		lag_cum_A = lag(cum_A, n = 1, default = NA),
		lag_R = lag(R, n = 1, default = NA))

# model for probability of receiving treatment
dataHW3_long <- mutate(dataHW3_long,
	time = as.numeric(variable) - 1,
	logCD4_diff = logCD4 - lag_logCD4)
dataHW3_long <- as.data.frame(dataHW3_long)

m3 <- glm(A ~ logCD4 + R, family = "binomial", 
		data = dataHW3_long)

dataHW3_long <- mutate(dataHW3_long,
	pred_trt = predict(m3, newdata = dataHW3_long, type = "response"),
	pred_trt_received = (pred_trt^A)*(1-pred_trt)^(1-A))
# calculate denominator weights
dataHW3$den_wt <- tapply(dataHW3_long$pred_trt_received, 
	dataHW3_long$ID, prod, na.rm = TRUE)
dataHW3 <- mutate(dataHW3,
	init_time = NA,
	init_time = ifelse(A0 == 1 & A1 == 1 & A2 == 1 & A3 == 1, 0, init_time),
	init_time = ifelse(A0 == 0 & A1 == 1 & A2 == 1 & A3 == 1, 1, init_time),
	init_time = ifelse(A0 == 0 & A1 == 0 & A2 == 1 & A3 == 1, 2, init_time),
	init_time = ifelse(A0 == 0 & A1 == 0 & A2 == 0 & A3 == 1, 3, init_time),
	init_time = ifelse(A0 == 0 & A1 == 0 & A2 == 0 & A3 == 0, 4, init_time))

MSM_CD4 <- lm(CD44 ~ init_time + I(init_time^2), data = dataHW3)
newdatatime <- data.frame(init_time = 4:0)
predCD4.boot <- cbind(predCD4.boot, predict(MSM_CD4, newdatatime))
}

predCD4$SE <- apply(predCD4.boot, 1, sd)
print(round(predCD4))
```
## Comparison Between G-computation and MSM

```{r, echo = FALSE, cache = TRUE }
ymin <- min(c(predCD4$CD4 - predCD4$SE, mean_trt_response[1:5, 2] - se[1:5, 2]))
ymax <- max(c(predCD4$CD4 + predCD4$SE, mean_trt_response[1:5, 2] + se[1:5, 2]))
plot(4:0 - 0.1, predCD4$CD4, xlim = c(-0.1, 4.1), ylim = c(ymin, ymax),
  xlab = "Time Interval Trt Started", ylab = "Estimated Mean CD4 +/- SE")
points(4:0 - 0.1, predCD4$CD4, col = "blue", pch = 19)
points(4:0 + 0.1,  mean_trt_response[1:5, 2], col = "red", pch = 19)
for(j in 1:5) {
  lines(rep(5-j - 0.1, 2), c((predCD4$CD4 - predCD4$SE)[j], (predCD4$CD4 + predCD4$SE)[j]), col = "blue")
  lines(rep(5-j + 0.1, 2), c(mean_trt_response[j, 2] - se[j, 2], 
    mean_trt_response[j, 2] + se[j, 2]), col = "red")
}
legend("bottomleft", c("MSM","G-comp"), col = c("blue", "red"), pch = 19, lty = 1)
```

